### 2015-06-12 paper.tex
first commit

\documentclass[fleqn,10pt,twoside]{gcb15submission}
\usepackage{url}\urlstyle{same}

\title{signalHsmm - a novel semi-Markov model of signal peptides}

\author[1]{Micha\l{}  Burdukiewicz}
\author[2]{Piotr Sobczyk}
\author[1]{Pawe\l{} B\l{}a\.{z}ej}
\author[1]{Pawe\l{} Mackiewicz}
\affil[1]{University of Wroc\l{}aw, Department of Genomics, Poland}
\affil[2]{Wroc\l{}aw University of Technology, Department of Mathematics, Poland}

\keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
The proper localization of proteins in a cell is essential to perform their desired function. Information about the protein destination is included within the very protein in the form of short peptides called targeting signals. Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome and play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. Moreover, the transport of proteins through the endomembrane system is important for their correct folding and posttranslational modifications. A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase.

However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides. Therefore, the main goal of the study is to design a new probabilistic model for signal peptides, which will include knowledge about their organization, amino acid composition and variation.

The proposed model will be based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is a possibility to incorporate unique regions in the architecture of signal peptides. Therefore, HSMMs can be useful in recognition of atypical and artificial signal peptides. An aggregation of amino acids into physicochemical groups using the neural gas algorithm with k-fold cross-validation will reduce dimensionality of the problem and enable to learn the algorithm even on smaller data sets. The reduction will allow to use more complex probabilistic models and models specialized in the detection of signal peptides in different taxonomic groups of organisms, which improve the efficiency of the whole predictor. The n-gram and position-specific weight matrix methods will add more details to the probabilistic model.

Our preliminary model has showed the largest AUC=0.98 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets. Thanks to that, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasised that our model describes signal peptides from medically significant malaria parasites Plasmodium and their relatives (AUC = 0.92) more accurately than popular programs (0.84).

\end{abstract}

\begin{document}
\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}

Your introduction goes here.
Some examples of commonly used {\LaTeX} commands and features are listed below, to help you get started.

All \textbf{accepted} submissions will be published as a \textbf{preprint collection} with \textbf{PeerJ} (\url{http://www.peerj.com}).
It will be clearly visible that the work
\begin{enumerate}
\item is a preprint (although reviewed and accepted by the GCB program committee)
\item has been presented at GCB'15
\end{enumerate}
by including a special GCB logo in the final version

If you decide to submit a full version after GCB to PeerJ or to PeerJ Computer Science, the procedure is very easy.
Of course, you are free to submit your paper to any other journal that allows existing preprints.

\subsection*{About PeerJ}

PeerJ is an award-winning open access publisher covering both computer science and the biological and medical sciences. 
PeerJ provides authors with three publication venues: PeerJ and PeerJ Computer Science (peer-reviewed academic journals) and PeerJ PrePrints (a 'pre-print server'). See \url{https://peerj.com/about/publications/} for more information.

The PeerJ model allows an author to publish articles in their peer-reviewed journal via the purchase of a lifetime Publication Plan. Prices start from just \$99 (a one-off payment) which entitles an author to the lifetime ability to publish 1 article per year for free.
The fee applies to all authors of an article, which encourages a reasonable number of co-authors and is generally less than for other open-access publishers.

\section*{Some \LaTeX{} Examples}
\label{sec:examples}

Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.

\subsection*{Figures and Tables}

Use the table environment and the tabular command for basic tables --- see Table~\ref{tab:widgets}, for example.

To include a figure in your document, use the figure environment and the includegraphics command as in the code for Figure~\ref{fig:uaruhr}.

\begin{figure}[ht]\centering
\includegraphics[width=0.6\textwidth]{uaruhr}
\caption{An example image.}
\label{fig:uaruhr}
\end{figure}

\begin{table}[ht]
\centering
\begin{tabular}{l|r}
Item & Quantity \\\hline
Widgets & 42 \\
Gadgets & 13
\end{tabular}
\caption{\label{tab:widgets}An example table.}
\end{table}

\subsection*{Citations}

LaTeX formats citations and references automatically using the bibliography records in your .bib file.
Use the \verb|\cite| command for an inline citation, like \cite{MuellerRahmann+2003RobustEstimation}, and the \verb|\citep| command for a citation in parentheses \citep{MuellerRahmann+2003RobustEstimation}.

\subsection*{Mathematics}

\LaTeX{} is great at typesetting mathematics. Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i$$
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

\subsection*{Lists}

You can make lists with automatic numbering \dots

\begin{enumerate}[noitemsep] 
\item Like this,
\item and like this.
\end{enumerate}
\dots or bullet points \dots
\begin{itemize}[noitemsep] 
\item Like this,
\item and like this.
\end{itemize}
\dots or with words and descriptions \dots
\begin{description}
\item[Word] Definition
\item[Concept] Explanation
\item[Idea] Text
\end{description}

\section*{Methods}

In a bioinformatics paper, the methods section should be the most important one.
Therefore, feel free to have more than one method section or to choose a more meaningful title for it.

\subsection*{Data selection}

Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06. The positive set contained 2589 sequences with an experimentally confirmed signal peptide and its cleavage site. Sequences with more than one cleavage site were excluded from the final data set. The negative set comprised 152272 sequences without any signal peptide. Protein sequences with ambiguous symbols: X, J, Z and B were removed from the final sets. Proteins with selenocysteine (U) were also excluded from data set, because there are no records of signal peptides containing this amino acid.

\subsection*{Clustering of amino acids}

AAIndex

\subsection*{Subsection}

Here is an interesting equation that may be helpful in some situations:
\begin{equation}
\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
\label{eq:refname2}
\end{equation}

\paragraph{Paragraph}
Nothing to see here. Move on.

\paragraph{Paragraph}
Really.
See Figure~\ref{fig:results} for more interesting results.

\begin{figure}[ht]\centering
\includegraphics[width=0.6\textwidth]{results}
\caption{Can you guess which function this is?}
\label{fig:results}
\end{figure}

\section*{Results and Discussion}

You may want to separate results, discussion and conclusion, according to your needs.

Please submit the final pdf file via EasyChair to the GCB'15 program committee by June 30, 2015.

\section*{Acknowledgments}
Thank you for your support!

\bibliography{sample}

\end{document}

### 2015-06-12 aminoacid_encoding.R

@@ -1,0 +1,89 @@
+library(seqinr)
+data(aaindex)
+library(signalHsmm)
+library(seqinr)
+library(cvTools)
+library(hmeasure)
+library(pbapply)
+
+#normalized values of amino acids ----------------------------------------
+aa_nvals <- t(sapply(aaindex, function(i) {
+  vals <- i[["I"]]
+  vals <- vals - min(vals, na.rm = TRUE)
+  vals/max(vals, na.rm = TRUE)
+}))
+
+colnames(aa_nvals) <- tolower(a(colnames(aa_nvals)))
+
+
+aa_props <- sapply(aaindex, function(i) i[["D"]])
+
+write.csv2(data.frame(property = unname(aa_props)), file = "aa_props.csv")
+
+
+traits <- list(size = c(63, 72, 109, 399),
+               hydroph = c(54, 68, 151, 244),
+               polarity = c(111, 321),
+               alpha = c(3, 41, 253))
+
+# clustering of amino acids -------------------------------------------------
+
+all_traits_combn <- expand.grid(traits)
+
+
+
+all_groups <- apply(all_traits_combn, 1, function(single_trait_combn) {
+  #use ward, because Piotr does it
+  cl <- hclust(dist(t(aa_nvals[single_trait_combn, ])), method = "ward.D2")
+  gr <- cutree(cl, k = 4)
+  names(gr) <- tolower(names(gr))
+  agg_gr <- lapply(unique(gr), function(single_group) names(gr[gr == single_group]))
+  names(agg_gr) <- 1L:length(agg_gr)
+  agg_gr
+})
+
+all_groups <- all_groups[1:2]
+
+
+# read data ---------------------------------
+
+if(Sys.info()["nodename"] == "MICHALKOMP" )
+  pathway <- "C:/Users/Michal/Dropbox/signal-peptide2_data/"
+
+if(Sys.info()["nodename"] == "phobos" )
+  pathway <- "/home/michal/Dropbox/signal-peptide2_data/"
+
+
+pos_seqs <- read_uniprot(paste0(pathway, "signal_peptides.txt"), euk = TRUE, what = "signal")
+neg_seqs <- read.fasta(paste0(pathway, "nonsignal_peptides.fasta"), seqtype = "AA")
+#remove sequences with atypical aminoacids
[…]

### 2015-06-12 aminoacid_encoding.R
fisnished cross-validation code

@@ -42,9 +42,6 @@ all_groups <- apply(all_traits_combn, 1, function(single_trait_combn) {
   agg_gr
 })
 
-all_groups <- all_groups[1:2]
-
-
 # read data ---------------------------------
 
 if(Sys.info()["nodename"] == "MICHALKOMP" )
@@ -65,7 +62,7 @@ too_short <- which(sapply(pos_seqs, length) < 80)
 pos_seqs <- pos_seqs[-c(too_short)]
 
 
-# cross-validation
+# cross-validation ---------------------------------
 
 fold_res <- pblapply(1L:10, function(dummy) {
   #assure the same proteins in each fold for each repetition
@@ -80,10 +77,9 @@ fold_res <- pblapply(1L:10, function(dummy) {
         c(prob = single_pred[["sp_probability"]], cs_pred = single_pred[["sp_end"]]))),
         cs_real = sapply(test_dat, function(i) 
           ifelse(is.null(attr(i, "sig")[2]), NA, attr(i, "sig")[2])))
-      print("dzialam")
       preds
     })
   })
 })
 
-save(fold_res, all_groups, file = paste0(pathway, "fold_res_df.RData"))
+save(fold_res, all_groups, file = paste0(pathway, "fold_res_MB1.RData"))

### 2015-06-13 paper.tex
commit before the storm

@@
 
-AAIndex
+The functionality of a signal peptide results not from specific amino acids, but more generally from their physicochemical properties.
+Henceforth, the usage of raw amino acid sequences is superflous and introduces too much noise.
+Firstly, we chose several attributes based on the previous studies of signal peptides.
+We selected hydrophobicity (important for the transfer of the tagged protein), probability of being in alpha-helix (all signal peptides are alpha helices)
+
+imposed by several  requirements, not by presence Using
+
+To cluster amino acids we used several properties important for the structure and functionality of signal peptides. We chose:
+\begin{enumerate}[noitemsep] 
+\item hydrophobicity (important for )
+\end{enumerate}
+
+from AAIndex database (aaindexcitation).
+We divide properties into several groups of the interest based on the knowledge about signal peptides regional structure.
+The amino acids were clustered using euclidean distance and Ward's method.
 

### 2015-06-14 paper.tex
references added

@@
 
-In a bioinformatics paper, the methods section should be the most important one.
-Therefore, feel free to have more than one method section or to choose a more meaningful title for it.
+\subsection*{The framework structure}
+
+The functionality of a signal peptide depends not on exact sequence of specific amino acids, but on the physicochemical properties of residues in a given region.
+Henceforth, the usage of raw amino acid sequences is superflous and introduces too much noise.
+Our algorithm is trained on the set of amino acids classified into several groups.
+We use hidden semi-Markov model as the stochastic framework, because it does not imply unrealistic assumptions.
 
@@
 
-The functionality of a signal peptide results not from specific amino acids, but more generally from their physicochemical properties.
-Henceforth, the usage of raw amino acid sequences is superflous and introduces too much noise.
 Firstly, we chose several attributes based on the previous studies of signal peptides.

### 2015-06-14 aminoacid_encoding.R

@@ -1,10 +1,12 @@
 library(seqinr)
-data(aaindex)
 library(signalHsmm)
 library(seqinr)
 library(cvTools)
 library(hmeasure)
 library(pbapply)
+
+data(aaindex)
+
 
 #normalized values of amino acids ----------------------------------------
 aa_nvals <- t(sapply(aaindex, function(i) {
@@ -25,6 +27,11 @@ traits <- list(size = c(63, 72, 109, 399),
                hydroph = c(54, 68, 151, 244),
                polarity = c(111, 321),
                alpha = c(3, 41, 253))
+
+# piece of code below is useful for getting the references of measures that we used
+# sapply(aaindex[unlist(traits)], function(i)
+#   c(i[c("A", "T")]))
+
 
 # clustering of amino acids -------------------------------------------------
 

### 2015-06-15 paper.tex
fixed regexp in the endnote conversion tool

@@
 LaTeX formats citations and references automatically using the bibliography records in your .bib file.
-Use the \verb|\cite| command for an inline citation, like \cite{MuellerRahmann+2003RobustEstimation}, and the \verb|\citep| command for a citation in parentheses \citep{MuellerRahmann+2003RobustEstimation}.
+Use the \verb|\cite| command for an inline citation, like
 
@@
 The functionality of a signal peptide depends not on exact sequence of specific amino acids, but on the physicochemical properties of residues in a given region.
-Henceforth, the usage of raw amino acid sequences is superflous and introduces too much noise.
-Our algorithm is trained on the set of amino acids classified into several groups.
-We use hidden semi-Markov model as the stochastic framework, because it does not imply unrealistic assumptions.
+Henceforth, the usage of raw amino acid sequences is superflous and introduces unnecessary information.
+To utilize this property of signal peptide recognision, we cluster amino acids into several groups based on the physicochemical properties of residues.
+
+The pre-processed sequences are further analyzed by the heuristic algorithm~\citep{1998nielsenprediction}, which determines borders between three characteristic signal peptide regions.
+Next, two models are trained to recognize proteins with and without a signal peptide.
+The first one is a hidden semi-Markov model, in which each of three signal peptide regions is represented by a different hidden state.
+The additional fourth hidden state represents mature protein.
+Each state is described by the frequencies of amino acid groups  within that state.
+The distribution of hidden states durations, the number of amino acids related to each hidden state in signal peptide, is based on the empirical density of region lengths from the training set.
+The second model is a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.
 
@@
 Sequences with more than one cleavage site were excluded from the final data set.
-The negative set comprised 152272 sequences without any signal peptide.
+The negative set comprised 152272 sequences without any signal peptide annotation.
 Protein sequences with ambiguous symbols: X, J, Z and B were removed from the final sets.
@@
 
-Firstly, we chose several attributes based on the previous studies of signal peptides.
+We cluster amino acids using several criterions selected from AAindex database.
+
+e chose several attributes based on the previous studies of signal peptides.
 We selected hydrophobicity (important for the transfer of the tagged protein), probability of being in alpha-helix (all signal peptides are alpha helices)
-
-imposed by several  requirements, not by presence Using
 
@@
 
-\bibliography{sample}
+\bibliography{lokalizom}
 

### 2015-06-15 endnote_export.R

@@ -2,7 +2,7 @@ library(dplyr)
 
 bib_lines <- readLines("lokalizom.txt")
 
-starts <- c(grep("@", bib_lines), length(bib_lines) + 1)
+starts <- c(grep("^@", bib_lines), length(bib_lines) + 1)
 
 bib_names <- sapply(1L:(length(starts) - 1), function(i) {
   pub_lines <- bib_lines[starts[i]:((starts[i + 1]) - 1)]
@@ -37,8 +37,8 @@ bib_names <- sapply(1L:(length(starts) - 1), function(i) {
 })
 
 bib_names <- paste0(bib_names, ",")
-bib_starts <- grep("@", bib_lines)
+bib_starts <- grep("^@", bib_lines)
 for(i in 1L:length(bib_starts)) {
-  bib_lines[bib_starts[i]] <- paste0(bib_lines[bib_starts[i]], bib_names[i])
+  bib_lines[bib_starts[i]] <- paste0("@article{", bib_names[i])
 }
 writeLines(bib_lines, "lokalizom.bib")

### 2015-06-15 paper.tex
aded property table

@@
 \usepackage{url}\urlstyle{same}
+\usepackage{booktabs}
+\usepackage{colortbl, xcolor}
 
@@
 
-\subsection*{The framework structure}
+\subsection*{Overview}
 
 The functionality of a signal peptide depends not on exact sequence of specific amino acids, but on the physicochemical properties of residues in a given region.
-Henceforth, the usage of raw amino acid sequences is superflous and introduces unnecessary information.
-To utilize this property of signal peptide recognision, we cluster amino acids into several groups based on the physicochemical properties of residues.
+Henceforth, the usage of raw amino acid sequences is superfluous and introduces unnecessary information.
+To utilize this property of signal peptide recognition, we cluster amino acids into several groups based on the physicochemical properties of residues.
 
-The pre-processed sequences are further analyzed by the heuristic algorithm~\citep{1998nielsenprediction}, which determines borders between three characteristic signal peptide regions.
+The pre-processed sequences are further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions, the enhanced version of algorithm presented in~\cite{1998nielsenprediction}.
+Using the current information from experimentalists, we refined the region recognition criteria.
+
 Next, two models are trained to recognize proteins with and without a signal peptide.
@@
 The distribution of hidden states durations, the number of amino acids related to each hidden state in signal peptide, is based on the empirical density of region lengths from the training set.
+
 The second model is a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.
@@
 
-We cluster amino acids using several criterions selected from AAindex database.
+We cluster amino acids using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
+High values of hydrophobicity are required in the h-region, the core of signal peptides.
+Alpha-helix, the secondary structure of this region, is probably induced by the positively charged n-region.
+Polarity as well as size are important especially in the cleavage site~\citep{1994palzkillselection}.
 
-e chose several attributes based on the previous studies of signal peptides.
-We selected hydrophobicity (important for the transfer of the tagged protein), probability of being in alpha-helix (all signal peptides are alpha helices)
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+  \toprule
+Criterion name & Property name \\ 
+  \midrule
+Size & Size \\ 
+   \rowcolor[gray]{0.85}Size & Molecular weight \\ 
+  Size & Residue volume \\ 
+   \rowcolor[gray]{0.85}Size & Bulkiness \\ 
+  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins \\ 
+   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale \\ 
+  Hydrophobicity & Hydropathy index \\ 
+   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix \\ 
+  Polarity & Polarity \\ 
+   \rowcolor[gray]{0.85}Polarity & Mean polarity \\ 
+  Frequency in alpha-helices & Signal sequence helical potential \\ 
+   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix \\ 
+  Frequency in alpha-helices & Relative frequency in alpha-helix \\ 
+   \bottomrule
+\end{tabular}
+\caption{Properties used in clusterization.} 
+\label{tab:aaprop}
+\end{table}
 
-To cluster amino acids we used several properties important for the structure and functionality of signal peptides. We chose:
-\begin{enumerate}[noitemsep] 
-\item hydrophobicity (important for )
-\end{enumerate}
-
-from AAIndex database (aaindexcitation).
-We divide properties into several groups of the interest based on the knowledge about signal peptides regional structure.
-The amino acids were clustered using euclidean distance and Ward's method.
+We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}).
+Each property was attributed to a single criterion.
+Considering all combinations of single properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
 

### 2015-06-15 aminoacid_encoding.R

@@ -31,7 +31,16 @@ traits <- list(size = c(63, 72, 109, 399),
 # piece of code below is useful for getting the references of measures that we used
 # sapply(aaindex[unlist(traits)], function(i)
 #   c(i[c("A", "T")]))
-
+# tab <- data.frame(criterion = names(unlist(traits)), name = unlist(lapply(aaindex[unlist(traits)], function(i)
+#   c(i[c("D")]))), row.names = NULL)
+# tab[["criterion"]] <- factor(sapply(as.character(tab[["criterion"]]), function(i) substr(i, 0, nchar(i) - 1)))
+# levels(tab[["criterion"]]) <- c("Frequency in alpha-helices", "Hydrophobicity", "Polarity", "Size")
+# tab[["name"]] <- sapply(strsplit(as.character(tab[["name"]]), " (", fixed = TRUE), function(i) i[[1]])
+# rws <- seq(1, nrow(tab) - 1, by = 2)
+# col <- rep("\\rowcolor[gray]{0.85}", length(rws))
+# colnames(tab) <- c("Criterion name", "Property name")
+# print(xtable(tab, "Properties used in clusterization.", label = "tab:aaprop"), include.rownames = FALSE, booktabs = TRUE,
+#       add.to.row = list(pos = as.list(rws), command = col))
 
 # clustering of amino acids -------------------------------------------------
 

### 2015-06-15 paper.tex
added signal peptide figure

@@
 
-\title{signalHsmm - a novel semi-Markov model of signal peptides}
+\title{signalHsmm - a novel semi-Markov model of eukaryotic signal peptides}
 
@@
 \begin{abstract}
-The proper localization of proteins in a cell is essential to perform their desired function.
+The proper localization of proteins in a cell is essential to maintain their desired function.
 Information about the protein destination is included within the very protein in the form of short peptides called targeting signals.
 Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell.
-Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome and play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
+Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
 Moreover, the transport of proteins through the endomembrane system is important for their correct folding and posttranslational modifications.
+
 A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase.
+However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
+Therefore, we designed a new, more univeral probabilistic model for eukaryotic signal peptides, which includes knowledge about their organisation, amino acid composition and variation.
 
-However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
-Therefore, the main goal of the study is to design a new probabilistic model for signal peptides, which will include knowledge about their organization, amino acid composition and variation.
-
-The proposed model will be based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
-The big advantage of the algorithm is a possibility to incorporate unique regions in the architecture of signal peptides.
-Therefore, HSMMs can be useful in recognition of atypical and artificial signal peptides.
-An aggregation of amino acids into physicochemical groups using the neural gas algorithm with k-fold cross-validation will reduce dimensionality of the problem and enable to learn the algorithm even on smaller data sets.
-The reduction will allow to use more complex probabilistic models and models specialized in the detection of signal peptides in different taxonomic groups of organisms, which improve the efficiency of the whole predictor.
-The n-gram and position-specific weight matrix methods will add more details to the probabilistic model.
+The proposed model is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
+The big advantage of the algorithm is its extensibility.
+Using the k-mers we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.
 
@@
 
-Your introduction goes here.
-Some examples of commonly used {\LaTeX} commands and features are listed below, to help you get started.
+Proteins of eukaryotes are encoded in nuclear genomes and are synthetized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum.
+After the translation process, proteins have to be targeted to specific subcellular compartments or exported outside the cell to the extracellular environment.
+The proper localization of proteins is essential to perform their desired function.
+Information about protein destination is included within the very protein in the form of short peptides or stretches of amino acid residues called targeting or sorting signals.
+Ones of them are signal peptides, which are located at the N-terminus of proteins.
 
-All \textbf{accepted} submissions will be published as a \textbf{preprint collection} with \textbf{PeerJ} (\url{http://www.peerj.com}).
-It will be clearly visible that the work
-\begin{enumerate}
-\item is a preprint (although reviewed and accepted by the GCB program committee)
-\item has been presented at GCB'15
-\end{enumerate}
-by including a special GCB logo in the final version
+Signal peptides are responsible for targeting of proteins to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes.
+Such proteins can stay inside of these compartments, or can be inserted into cellular membranes or exported outside the cell.
+Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome.
+They play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}.
+Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.
 
-If you decide to submit a full version after GCB to PeerJ or to PeerJ Computer Science, the procedure is very easy.
-Of course, you are free to submit your paper to any other journal that allows existing preprints.
+Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious.
+Therefore, the development of new approaches in the field of computational biology and bioinformatics is desirable.
+They are not only a good support, complement or alternative for the experimental methods but also enable to understand rules encoding information about protein targeting, which is contained in the predicted targeting signals.
+However, many of the present predictors disregard full biological information carrying by signal peptides.
+Moreover, they often use orthogonal amino acids encodings, which greatly increase the model complexity without preserving physiochemical properties of residues.
+The current signal predicting software is mainly based on sequences of frequently studied species, such as yeast, angiosperms and mammals.
+However, our analysis of the large number of sequences representing diverse taxonomical groups indicates a great variability of signal peptides.
+The mentioned limitations result in inflexible models, which inefficiently extract data and henceforth cannot be used to model signal peptides restricted to smaller taxonomic groups and atypical.
 
-\subsection*{About PeerJ}
-
-PeerJ is an award-winning open access publisher covering both computer science and the biological and medical sciences. 
-PeerJ provides authors with three publication venues: PeerJ and PeerJ Computer Science (peer-reviewed academic journals) and PeerJ PrePrints (a 'pre-print server').
-See \url{https://peerj.com/about/publications/} for more information.
-
-The PeerJ model allows an author to publish articles in their peer-reviewed journal via the purchase of a lifetime Publication Plan.
-Prices start from just \$99 (a one-off payment) which entitles an author to the lifetime ability to publish 1 article per year for free.
-The fee applies to all authors of an article, which encourages a reasonable number of co-authors and is generally less than for other open-access publishers.
-
-\section*{Some \LaTeX{} Examples}
-\label{sec:examples}
-
-Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.
-
-\subsection*{Figures and Tables}
-
-Use the table environment and the tabular command for basic tables --- see Table~\ref{tab:widgets}, for example.
-
-To include a figure in your document, use the figure environment and the includegraphics command as in the code for Figure~\ref{fig:uaruhr}.
-
-\begin{figure}[ht]\centering
-\includegraphics[width=0.6\textwidth]{uaruhr}
-\caption{An example image.}
-\label{fig:uaruhr}
-\end{figure}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{l|r}
-Item & Quantity \\\hline
-Widgets & 42 \\
-Gadgets & 13
-\end{tabular}
-\caption{\label{tab:widgets}An example table.}
-\end{table}
-
-\subsection*{Citations}
-
-LaTeX formats citations and references automatically using the bibliography records in your .bib file.
-Use the \verb|\cite| command for an inline citation, like
-
-\subsection*{Mathematics}
-
-\LaTeX{} is great at typesetting mathematics.
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i$$
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
-
-\subsection*{Lists}
-
-You can make lists with automatic numbering \dots
-
-\begin{enumerate}[noitemsep] 
-\item Like this,
-\item and like this.
-\end{enumerate}
-\dots or bullet points \dots
-\begin{itemize}[noitemsep] 
-\item Like this,
-\item and like this.
-\end{itemize}
-\dots or with words and descriptions \dots
-\begin{description}
-\item[Word] Definition
-\item[Concept] Explanation
-\item[Idea] Text
-\end{description}
+Taking into account the importance of proteins with signal peptides, the aim of the project became a design for a new probabilistic model for signal peptide, which will include knowledge about their organization, amino acid composition and variation.
+The model will be based on hidden semi-Markov models.
+It will be created for different taxonomic groups of organisms, validated through a prediction algorithm and incorporated into a user-friendly software.
+The program will be also compared with other signal peptide predictors and competence area of such software will be determined.
+In the project, we are also planning to analyse differences in amino acid composition of signal peptides in different groups of organisms and recognise various types of signal peptides according to their distinctive features.
 
@@
 
+\begin{figure}[ht]\centering
+\includegraphics[width=0.5\textwidth]{SP.png}
+\caption{The structure of signal peptide.}
+\label{fig:sparch}
+\end{figure}
+
 \subsection*{Data selection}
@@
 
-We cluster amino acids using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
+Amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
 High values of hydrophobicity are required in the h-region, the core of signal peptides.
@@
 
-We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}).
-Each property was attributed to a single criterion.
-Considering all combinations of single properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
+We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
+Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
 
-\subsection*{Subsection}
+To compare encodings we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding.
+We created balanced data sets by subsampling a number of proteins without a signal peptide equal to the number of proteins with signal peptide.
+The cross-validation was repeated XXX times, to ensure that every negative protein was tested at least once with XXX probability.
 
-Here is an interesting equation that may be helpful in some situations:
-\begin{equation}
-\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
-\label{eq:refname2}
-\end{equation}
+\subsection*{Hidden semi-Markov model}
 
-\paragraph{Paragraph}
-Nothing to see here. Move on.
+Tu troche matematyki o ukrytych modelach Markova.
 
-\paragraph{Paragraph}
-Really.
-See Figure~\ref{fig:results} for more interesting results.
-
-\begin{figure}[ht]\centering
-\includegraphics[width=0.6\textwidth]{results}
-\caption{Can you guess which function this is?}
-\label{fig:results}
-\end{figure}
-
-\section*{Results and Discussion}
+\section*{Results}
 
@@
 Please submit the final pdf file via EasyChair to the GCB'15 program committee by June 30, 2015.
+
+\section*{Discussion}
 

### 2015-06-15 paper.tex
added plots and text

@@
 
-\keywords{Keyword1, Keyword2, Keyword3}
+\keywords{signal petide prediction; n-gram; hidden semi-Markov models}
 
@@
 The big advantage of the algorithm is its extensibility.
-Using the k-mers we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.
+Using the n-grams (k-mers) we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.
 
-Our preliminary model has showed the largest AUC=0.98 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets.
+Our model was validated a signal peptide pedictor.
+It has showed the largest AUC=0.98 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets.
 Thanks to that, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
@@
 
+\subsection*{Signal peptides}
+
 Proteins of eukaryotes are encoded in nuclear genomes and are synthetized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum.
@@
 
-Signal peptides are responsible for targeting of proteins to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes.
+Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes.
 Such proteins can stay inside of these compartments, or can be inserted into cellular membranes or exported outside the cell.
@@
 
+Although signal peptides are quite variable, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig.
+\ref{fig:sparch}.
+It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues.
+They probably enforce a proper topology on the polypeptide during translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}.
+The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues.
+It constitutes a core region of signal peptide and has a tendency to form $\alpha$-helix.
+The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site recognized by the signal peptidase.
+The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.
+
+\begin{figure}[ht]\centering
+\includegraphics[width=0.55\textwidth]{SP.png}
+\caption{The organization of signal peptide.}
+\label{fig:sparch}
+\end{figure}
+
+During or after translocation of the protein into the lumen of endoplasmic reticulum, the typical signal peptide is cleaved by a signal peptidase~\citep{2002paetzelsignal} and next degraded by specific proteases, whereas the rest (mature) part of protein stays in the lumen or is passed to other compartments.
+The cleavage site is characterized by a very variable amino acid composition.
+It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}.
+The site is, however, absent from some membrane proteins in which the first transmembrane-domain acts both as signal peptides and signal anchor~\citep{1988szczesnaskorupapositive}.
+
+Some data indicate that signal peptides may be universal.
+It was found, for example, that even bacterial signal peptides targeted correctly transgenic proteins to the plant~\citep{2009moellera} or mammalian secretory system~\citep{2014naganoestablishment}.
+On the other hand, signal peptides show great variation and the description presented above (Fig.
+\ref{fig:sparch}) refers to the most ‘typical’ signal peptides.
+There are also exceptionally long signal peptides, which fulfil more sophisticated roles~\citep{2009hissarchitecture}.
+For example, the fragment of signal peptide from preprolactin probably takes part in the regulation of prolactin secretion.
+Other examples are signal peptides of MHC class I, which inhibit activity of NK cells.
+Interesting functions have signal peptides of viral origin, which are involved in the immune evasion or viral life cycle~\citep{2000kappposttargeting}.
+Such diverse functions are restricted not only to the long signal peptides.
+The peptides from midkine, a protein contributing to the tumor progression, contains epitopes recognized by CD4+ T cells~\citep{2013kerzerhothe}.
+It indicates that the signal peptide may participate in a tumor immunity.
+The functional significance of these targeting signals makes that the prediction of signal peptide-containing proteins is also an important step in the drug development~\citep{2005zhangalteration, 2012netoadeimproving, 2010moellerwetmilling}.
+
+\subsection*{Signal peptide predicting software}
+
 Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious.
@@
 However, many of the present predictors disregard full biological information carrying by signal peptides.
-Moreover, they often use orthogonal amino acids encodings, which greatly increase the model complexity without preserving physiochemical properties of residues.
-The current signal predicting software is mainly based on sequences of frequently studied species, such as yeast, angiosperms and mammals.
-However, our analysis of the large number of sequences representing diverse taxonomical groups indicates a great variability of signal peptides.
-The mentioned limitations result in inflexible models, which inefficiently extract data and henceforth cannot be used to model signal peptides restricted to smaller taxonomic groups and atypical.
 
-Taking into account the importance of proteins with signal peptides, the aim of the project became a design for a new probabilistic model for signal peptide, which will include knowledge about their organization, amino acid composition and variation.
-The model will be based on hidden semi-Markov models.
-It will be created for different taxonomic groups of organisms, validated through a prediction algorithm and incorporated into a user-friendly software.
-The program will be also compared with other signal peptide predictors and competence area of such software will be determined.
-In the project, we are also planning to analyse differences in amino acid composition of signal peptides in different groups of organisms and recognise various types of signal peptides according to their distinctive features.
+Having functional importance and specific features, signal peptides became the subject of many programs to their prediction based on different methods.
+The state of the art software for predicting the presence of signal peptides often incorporates ‘black-box’ models, as neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}.
+Such models neither utilize biological information about signal peptides nor work properly on atypical signal peptides.
+
+The other type of software incorporates into the prediction process theoretical knowledge about the structure of signal peptides.
+Although these programs do not share innate flaws of ‘black-box’ models, they also demand an improvement.
+Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}.
+Others (not supported SignalP 3.0, Philius and Phobius) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework.
+However, the HMMs imply geometric distribution for duration of region length.
+We replicated the rules for extracting regions' boundaries from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the mentioned assumption is not in concordance with the reality because every region revealed the length distribution other than geometric (Fig.
+\ref{fig:reglen}).
+Moreover, the commonly used rigid region scheme (Fig.
+\ref{fig:sparch}) does not describe extremely long or short signal peptides.
+Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also unusual structures, but such probabilistic frameworks have not still been implemented.
+
+\begin{figure}[ht]\centering
+\includegraphics[width=0.9\textwidth]{reglen.png}
+\caption{Distribution of the lengths of signal peptide regions.
+The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
+\label{fig:reglen}
+\end{figure}
+
+Another feature, shared by majority of the signal peptide predicting software, is orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
+This method of encoding, however, does not preserve relationships between amino acids.
+The distance between residues is the same regardless of the similarity of their physicochemical properties.
+Such behaviour is especially undesired in signal peptides models because their regions are defined by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
+In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.
+
+All programs used in signal peptide recognition are trained on real protein sequences.
+Therefore, they succeed in the recognition of peptides similar to those in the learning set, but fail in the case of artificial signal peptides.
+Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}.
+They are especially important in industrial applications to increase yield of proteins.
+Therefore, only explicit knowledge about the organization of signal peptides allows creating sequences that will be the most efficient in the export of proteins~\citep{2013ngengineering}.
+Signal peptides have also an important application in gene therapy.
+Mimicking the natural mechanism of protein export, artificial signal peptides with tumor epitopes increase the antitumor immune response~\citep{2003heenhanced}.
+Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure.
+Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfil the designed role.
 
@@
 
-\begin{figure}[ht]\centering
-\includegraphics[width=0.5\textwidth]{SP.png}
-\caption{The structure of signal peptide.}
-\label{fig:sparch}
-\end{figure}
-
 \subsection*{Data selection}
@@
 Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06.
-The positive set contained 2589 sequences with an experimentally confirmed signal peptide and its cleavage site.
+The positive set contained 2~589 sequences with an experimentally confirmed signal peptide and its cleavage site.
 Sequences with more than one cleavage site were excluded from the final data set.
-The negative set comprised 152272 sequences without any signal peptide annotation.
+The negative set comprised 152~272 sequences without any signal peptide annotation.
 Protein sequences with ambiguous symbols: X, J, Z and B were removed from the final sets.
@@
 
-Tu troche matematyki o ukrytych modelach Markova.
+Tu troche matematyki o ukrytych modelach Markowa~\citep{1989rabinera}.
 

### 2015-06-15 sp_length.R

@@ -1,0 +1,39 @@
+library(reshape2)
+
+#distributions of lengths of dignal peptides regions
+
+if(Sys.info()["nodename"] == "MICHALKOMP" )
+  pathway <- "C:/Users/Michal/Dropbox/signal-peptide2_data/"
+
+if(Sys.info()["nodename"] == "phobos" )
+  pathway <- "/home/michal/Dropbox/signal-peptide2_data/"
+
+
+pos_seqs <- read_uniprot(paste0(pathway, "signal_peptides.txt"), euk = TRUE, what = "signal")
+
+nhc_borders <- t(sapply(pos_seqs, find_nhc))
+
+nhc_len <- sapply(2L:ncol(nhc_borders), function(i)
+  nhc_borders[, i] - nhc_borders[, i - 1])
+
+colnames(nhc_len) <- c("n", "h", "c")
+mlen <- melt(nhc_len)
+
+levels(mlen[["Var2"]]) <- paste0(levels(mlen[["Var2"]]), "-region")
+
+
+source("plot_tools.R")
+
+p1 <- ggplot(mlen, aes(x = value)) +
+  geom_density(col = "blue", fill = adjustcolor("blue", 0.25)) +
+  facet_wrap(~ Var2) +
+  scale_x_continuous("Length \n ") + 
+  scale_y_continuous("Value") +
+  my_theme
+
+
+png("reglen.png", width = 2257*0.5, height = 801*0.5)
+print(p1)
+dev.off()
+  
+

### 2015-06-15 sp_length.R
added new plot, synchronised references

@@ -1,4 +1,5 @@
 library(reshape2)
+library(signalHsmm)
 
 #distributions of lengths of dignal peptides regions
 
@@ -21,19 +22,35 @@ mlen <- melt(nhc_len)
 
 levels(mlen[["Var2"]]) <- paste0(levels(mlen[["Var2"]]), "-region")
 
+whole_sp <- data.frame(value = nhc_borders[, "cs"] - 1)
+whole_sp <- cbind(var = rep("signal peptide", nrow(whole_sp)), whole_sp)
+
+mlen <- cbind(mlen, ann = ifelse(mlen[["Var2"]] == "n-region", "B", ""))
 
 source("plot_tools.R")
+
+
+p2 <- ggplot(whole_sp, aes(x = value)) + 
+  geom_density(col = "blue", fill = adjustcolor("blue", 0.25)) +
+  scale_x_continuous("Length \n ") + 
+  facet_wrap(~ var) +
+  scale_y_continuous("Value") + 
+  #annotate("text", x = 5, y = 0.09, label = "A", fontface = "bold") +
+  my_theme
 
 p1 <- ggplot(mlen, aes(x = value)) +
   geom_density(col = "blue", fill = adjustcolor("blue", 0.25)) +
   facet_wrap(~ Var2) +
   scale_x_continuous("Length \n ") + 
   scale_y_continuous("Value") +
+  #geom_text(aes(x = 5, y = 0.3, label = ann), fontface = "bold") +
   my_theme
 
 
-png("reglen.png", width = 2257*0.5, height = 801*0.5)
-print(p1)
+png("reglen.png", width = 2257*0.5, height = 1201*0.5)
+print(arrangeGrob(textGrob("A", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p2, 
+                  textGrob("B", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p1, 
+                  nrow = 2, ncol = 2, widths = c(0.05, 0.95)))
 dev.off()
   
 

### 2015-06-16 endnote_export.R
endnote_export.R now removes note section

@@ -4,7 +4,7 @@ bib_lines <- readLines("lokalizom.txt")
 
 starts <- c(grep("^@", bib_lines), length(bib_lines) + 1)
 
-bib_names <- sapply(1L:(length(starts) - 1), function(i) {
+bib_list <- lapply(1L:(length(starts) - 1), function(i) {
   pub_lines <- bib_lines[starts[i]:((starts[i + 1]) - 1)]
   author_line <- pub_lines[grep("author = ", pub_lines)]
   if(length(author_line) > 0) {
@@ -33,12 +33,19 @@ bib_names <- sapply(1L:(length(starts) - 1), function(i) {
     final_title <- c()
   }
   
-  paste0(final_year, final_author, final_title)
+  #first line
+  pub_lines[1] <- paste0("@article{", final_year, final_author, final_title, ",")
+  
+  #remove note
+  note_start <- grep("note = {", pub_lines, fixed = TRUE)
+  if(length(note_start) > 0) {
+    pub_lines[-(note_start:(note_start + grep("\\},$", pub_lines[note_start:length(pub_lines)])[1] - 1))]
+  } else {
+    pub_lines
+  }
 })
 
-bib_names <- paste0(bib_names, ",")
-bib_starts <- grep("^@", bib_lines)
-for(i in 1L:length(bib_starts)) {
-  bib_lines[bib_starts[i]] <- paste0("@article{", bib_names[i])
-}
-writeLines(bib_lines, "lokalizom.bib")
+# bib_names <- paste0(bib_names, ",")
+# bib_starts <- grep("^@", bib_lines)
+
+writeLines(unlist(bib_list), "lokalizom.bib")

### 2015-06-16 aminoacid_encoding.R
adding pathways handling and output files

@@ -50,6 +50,12 @@ if(Sys.info()["nodename"] == "MICHALKOMP" )
 if(Sys.info()["nodename"] == "phobos" )
   pathway <- "/home/michal/Dropbox/signal-peptide2_data/"
 
+if(Sys.info()["nodename"] == "tobit" )
+  pathway <- "/home/piotr//Dropbox/signal-peptide2_data (1)//"
+
+if(Sys.info()["nodename"] == "sobczyk-pc"  )
+  pathway <- "/home/sobczyk/Dropbox//signal-peptide2_data (1)//"
+
 
 pos_seqs <- read_uniprot(paste0(pathway, "signal_peptides.txt"), euk = TRUE, what = "signal")
 neg_seqs <- read.fasta(paste0(pathway, "nonsignal_peptides.fasta"), seqtype = "AA")
@@ -64,7 +70,7 @@ pos_seqs <- pos_seqs[-c(too_short)]
 
 # cross-validation ---------------------------------
 
-fold_res <- pblapply(1L:10, function(dummy) {
+fold_res <- pblapply(1L:20, function(dummy) {
   #assure the same proteins in each fold for each repetition
   pos_ids <- cvFolds(length(pos_seqs), K = 5)
   cv_neg <- neg_seqs[sample(1L:length(neg_seqs), length(pos_seqs))]
@@ -82,4 +88,17 @@ fold_res <- pblapply(1L:10, function(dummy) {
   })
 })
 
-save(fold_res, all_groups, file = paste0(pathway, "fold_res_MB1.RData"))
+
+if(Sys.info()["nodename"] == "MICHALKOMP" )
+  output <- "fold_res_MB1.RData"
+
+if(Sys.info()["nodename"] == "phobos" )
+  output <- "fold_res_MB1.RData"
+
+if(Sys.info()["nodename"] == "tobit" )
+  output <- "fold_res_PS1.RData"
+
+if(Sys.info()["nodename"] == "sobczyk-pc"  )
+  output <- "fold_res_PS1.RData"
+
+save(fold_res, all_groups, file = paste0(pathway, output))

### 2015-06-16 aminoacid_encoding.R
passing arguments to scripts

@@ -1,3 +1,7 @@
+
+#' Run from command line
+#' Rscript aminoacid_encoding.R runID numberOfCrossValidationRuns
+
 library(seqinr)
 library(signalHsmm)
 library(seqinr)
@@ -7,6 +11,11 @@ library(pbapply)
 
 data(aaindex)
 
+args <- commandArgs(trailingOnly = TRUE)
+runID <- as.numeric(args[1])
+numberRuns <- as.numeric(args[2])
+message(paste0("aminoacid_encoding.R is running with run ID ", runID, ". Number of
+               cross-validation runs is ", numberRuns))
 
 #normalized values of amino acids ----------------------------------------
 aa_nvals <- t(sapply(aaindex, function(i) {
@@ -86,7 +95,7 @@ pos_seqs <- pos_seqs[-c(too_short)]
 
 # cross-validation ---------------------------------
 
-fold_res <- pblapply(1L:20, function(dummy) {
+fold_res <- pblapply(1L:numberRuns, function(dummy) {
   #assure the same proteins in each fold for each repetition
   pos_ids <- cvFolds(length(pos_seqs), K = 5)
   cv_neg <- neg_seqs[sample(1L:length(neg_seqs), length(pos_seqs))]
@@ -106,15 +115,16 @@ fold_res <- pblapply(1L:20, function(dummy) {
 
 
 if(Sys.info()["nodename"] == "MICHALKOMP" )
-  output <- "fold_res_MB1.RData"
+  output <- paste0("fold_res_MB", runID, ".RData")
 
 if(Sys.info()["nodename"] == "phobos" )
-  output <- "fold_res_MB1.RData"
+  output <- paste0("fold_res_MB", runID, ".RData")
 
 if(Sys.info()["nodename"] == "tobit" )
-  output <- "fold_res_PS1.RData"
+  output <- paste0("fold_res_PS", runID, ".RData")
 
 if(Sys.info()["nodename"] == "sobczyk-pc"  )
-  output <- "fold_res_PS1.RData"
+  output <- paste0("fold_res_PS", runID, ".RData")
 
 save(fold_res, all_groups, file = paste0(pathway, output))
+message(paste0("Results were saved to file ", pathway, output))

### 2015-06-16 paper.tex
first draft of HMM section

@@
 \usepackage{colortbl, xcolor}
+
+% rysunki
+\usepackage{tikz}
+\usepackage{ifthen}
+\usepackage{xxcolor}
+\usetikzlibrary{arrows}
+\usetikzlibrary[topaths]
+\usetikzlibrary{decorations.pathreplacing}
 
@@
 \includegraphics[width=0.9\textwidth]{reglen.png}
-\caption{Distribution of the lengths of signal peptide regions.
+\caption{A) Distribution of lengths of signal peptides.
+B) Distribution of the lengths of signal peptide regions.
 The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
@@
 
-Tu troche matematyki o ukrytych modelach Markowa~\citep{1989rabinera}.
+Hidden Markov Model(HMM) is a widely used machine learning technique which, in contrast to black-box methods like neural network used by SignalP, 
+is easy to interpret and allows researcher to analyze and understand structure of modeled phenomenon. 
+In case of signal-peptide, we might wish to verify whether regions really describe it. Also we might capture
+structure for some special subgroups and with HMM we can verify biological hypothesis.
+
+Suppose you have a sequence of observations e.g. amino acids. What is of interest is
+understanding underlying cause of their occurrence. 
+HMM aims to answer that question by assuming a specific yet flexible structure of the problem.
+HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
+These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
+$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
+Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
+Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
+Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
+In our case, observations are (degenerated) amino acids.
+Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
+Viterbi algorithm.
+For a good reference on HMM see~\citep{1989rabinera}.
+
+In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
+\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
+complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
+However it still not flexible enough. Empirical regions length distributions (see Figure~\ref{fig:reglen}),
+is difficult to capture in that way.
+
+%rysunek pokazujący o co chodzi w HMM
+\begin{figure}[h]
+\centering
+\tikzstyle{block} = [draw,shape=circle, top color=white, bottom color=green!50!red!70 ,minimum size=4em]
+\tikzstyle{blockComp} = [draw,shape=circle, top color=white, bottom color=red!50!black!70 ,minimum size=2.5em]
+% diameter of semicircle used to indicate that two lines are not connected
+\def\radius{.7mm} 
+\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=green!50!red!70 ,minimum size=3pt,inner sep=0pt]
+\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=red!50!black!70 ,minimum size=3pt,inner sep=0pt]
+\def\n{11}
+ \tikzstyle{line} = [draw=black, color=black!30!white!50, line width=1.5mm, -latex'] 
+ \tikzstyle{line2} = [draw=black, color=red!30!black!50, line width=1.5mm, -latex']
+\tikzstyle{frame}=[text=black,above, bottom color=white, top color=blue!50!black!70  ]
+\tikzstyle{frameComp}=[text=black,below, bottom color=white, top color=red!50!black!70  ]
+ \tikzstyle{line3} = [color=green!50!red!70, line width=1.5mm]
+\def\names{{"$O_{1,1}$","...","$O_{1,d_1}$", "...","$O_{k, 1}$", "...", "$O_{k, d_k}$"}}%
+\begin{tikzpicture}[>=latex']
+%wąsate nawiasy
+\draw [decorate, decoration={brace,amplitude=10pt},line3,xshift=0pt,yshift=0pt] (2.5,1) -> (10.5,1) node [branch,midway,yshift=0.8cm,color=black] 
+	{\textbf{Hidden states}};
+% \draw [->,decoration={brace,mirror,amplitude=10pt},line2,xshift=0pt,yshift=0pt] (11,-3) -- (3,-3) node [branch2,midway,yshift=-0.6cm,color=black] 
+% 	{\textbf{Kierunek odczytu na nici Cricka ($3' \to 5'$)}}; 
+   % Draw blocks, inputs and outputs
+
+\node[block, text width=3em] at (1+2,0) (block1) {\footnotesize 1st hidden state};
+\node[blockComp] at (-.5+2,-2)  (komp1) {\footnotesize \pgfmathparse{\names[0]}\pgfmathresult};
+\node[blockComp] at (1+2,-2)  (komp2) {\footnotesize \pgfmathparse{\names[1]}\pgfmathresult};
+\node[blockComp] at (2.5+2,-2)  (komp3) {\footnotesize \pgfmathparse{\names[2]}\pgfmathresult};
+\draw[line,->]  (block1) --  (komp1);
+\draw[line,->]  (block1) --  (komp2);
+\draw[line,->]  (block1) --  (komp3);
+\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
+
+\node[block, bottom color=green!50!red!80,] at (4.5+2,0) (block2) {...};
+\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
+\node[blockComp] at (4.5+2,-2)  (komp4) {\footnotesize \pgfmathparse{\names[3]}\pgfmathresult};
+\draw[line,->]  (block2) --  (komp4) ;
+
+\node[block,text width=3.5em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
+\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
+\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
+\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
+\draw[line,->]  (block3) --  (komp5);
+\draw[line,->]  (block3) --  (komp6);
+\draw[line,->]  (block3) --  (komp7);
+
+% \draw[<->, color=red!30!black!50, line width=1.5mm] (komp1) |- +(5,-1.5) node[line2, yshift=-0.5cm] {\textbf{Observations}} -| (komp7);
+
+\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
+(1,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
+
+\end{tikzpicture}
+\caption{Hidden semi-Markov Model scheme}
+\label{fig:hsmm}
+\end{figure}
+
+The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
+any given hidden state length distribution. Scheme is shown in figure~\ref{fig:hsmm}.
+Model is given by matrices $A$ and $B$ as before, as well as probabilities of duration length in hidden states.
+$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
+$D$ is maximal allowed duration.
+As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
+higher than in the regular HMM.
+
+The gain is not only a better fit to the data but also, as number of hidden states is smaller, easier interpretation 
+for researcher without mathematical background.
+
+HSMM can be further extend to use some specific knowledge of motives - n-grams.
 

### 2015-06-16 paper.tex
preliminary results of the analysis

--- 2015-06-15 paper.tex
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.55\textwidth]{SP.png}
+\includegraphics[width=0.55\textwidth]{figures/SP.png}
 \caption{The organization of signal peptide.}
@@
 Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}.
-Others (not supported SignalP 3.0, Philius and Phobius) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework.
+Others (Phobius, Philius and not supported SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework.
 However, the HMMs imply geometric distribution for duration of region length.
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{reglen.png}
-\caption{Distribution of the lengths of signal peptide regions.
-The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
+\includegraphics[width=0.9\textwidth]{figures/reglen.png}
+\caption{A) Distribution of lengths of signal peptides.
+B) Distribution of the lengths of signal peptide regions.
+The measure of the length is a number of amino acids in the subsequence.
+The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
 \label{fig:reglen}
@@
 
-Another feature, shared by majority of the signal peptide predicting software, is orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
+Another feature, shared by majority of the signal peptide predicting software, is the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
 This method of encoding, however, does not preserve relationships between amino acids.
@@
    \rowcolor[gray]{0.85}Size & Molecular weight \\ 
-  Size & Residue volume \\ 
-   \rowcolor[gray]{0.85}Size & Bulkiness \\ 
-  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins \\ 
-   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale \\ 
-  Hydrophobicity & Hydropathy index \\ 
-   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix \\ 
-  Polarity & Polarity \\ 
-   \rowcolor[gray]{0.85}Polarity & Mean polarity \\ 
-  Frequency in alpha-helices & Signal sequence helical potential \\ 
+  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
+   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
+  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
+   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
+  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
+   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
+  Polarity & Polarity~\citep{1974granthamamino} \\ 
+   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
+  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
    \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix \\ 
-  Frequency in alpha-helices & Relative frequency in alpha-helix \\ 
+  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
    \bottomrule
@@
 
-We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
+We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
 Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
@@
 
+\subsection*{n-gram extension}
+
+n-grams (k-mers) are vectors of n characters derived from input sequences.
+They can be added as the new hidden state to the hidden semi-Markov model.
+The rigid structure of n-grams may mirror cleavage sites, whose are more conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
+
+Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. XXX n-grams were further used in the analysis
+
 \section*{Results}
 
-You may want to separate results, discussion and conclusion, according to your needs.
-
-Please submit the final pdf file via EasyChair to the GCB'15 program committee by June 30, 2015.
+\begin{figure}[ht]\centering
+\includegraphics[width=0.9\textwidth]{figures/cvres.png}
+\caption{Results of cross-validation for 15 best performing amino acid encodings.}
+\label{fig:cvres}
+\end{figure}
 
@@
 
+Despite the presence of signal peptide predicting software, there are no exhaustive studies devoted to the analysis of the signal peptide structure and taxonomical variability.
+Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution.
+
+The proposed algorithm recognizing various signal peptides will reduce costs and speed up searching appropriate artificial signal peptides designed for protein secretion in expensive and time-consuming laboratory experiments.
+Moreover, the algorithm can be applied in recognition any sequence types on the nucleotide and amino acid levels, represented by small sets.
+Accessible as a stand-alone and browser-based software, its useful both for researches interested in the analysis of signal peptides.
+
 \section*{Acknowledgments}
-Thank you for your support!
+PS and PB devised the HSMM model of signal peptide. MB and PS co-wrote the software. MB gathered data and carried the analysis.
+MB, PS and PM drafted the manuscript. All authors revised the manuscript and approved the final version.
 

### 2015-06-16 aminoacid_encoding.R

@@ -29,18 +29,18 @@ traits <- list(size = c(63, 72, 109, 399),
                alpha = c(3, 41, 253))
 
 # piece of code below is useful for getting the references of measures that we used
-# sapply(aaindex[unlist(traits)], function(i)
-#   c(i[c("A", "T")]))
-# tab <- data.frame(criterion = names(unlist(traits)), name = unlist(lapply(aaindex[unlist(traits)], function(i)
-#   c(i[c("D")]))), row.names = NULL)
-# tab[["criterion"]] <- factor(sapply(as.character(tab[["criterion"]]), function(i) substr(i, 0, nchar(i) - 1)))
-# levels(tab[["criterion"]]) <- c("Frequency in alpha-helices", "Hydrophobicity", "Polarity", "Size")
-# tab[["name"]] <- sapply(strsplit(as.character(tab[["name"]]), " (", fixed = TRUE), function(i) i[[1]])
-# rws <- seq(1, nrow(tab) - 1, by = 2)
-# col <- rep("\\rowcolor[gray]{0.85}", length(rws))
-# colnames(tab) <- c("Criterion name", "Property name")
-# print(xtable(tab, "Properties used in clusterization.", label = "tab:aaprop"), include.rownames = FALSE, booktabs = TRUE,
-#       add.to.row = list(pos = as.list(rws), command = col))
+sapply(aaindex[unlist(traits)], function(i)
+  c(i[c("A", "T")]))
+tab <- data.frame(criterion = names(unlist(traits)), name = unlist(lapply(aaindex[unlist(traits)], function(i)
+  c(i[c("D")]))), row.names = NULL)
+tab[["criterion"]] <- factor(sapply(as.character(tab[["criterion"]]), function(i) substr(i, 0, nchar(i) - 1)))
+levels(tab[["criterion"]]) <- c("Frequency in alpha-helices", "Hydrophobicity", "Polarity", "Size")
+tab[["name"]] <- sapply(strsplit(as.character(tab[["name"]]), " (", fixed = TRUE), function(i) i[[1]])
+rws <- seq(1, nrow(tab) - 1, by = 2)
+col <- rep("\\rowcolor[gray]{0.85}", length(rws))
+colnames(tab) <- c("Criterion name", "Property name")
+print(xtable(tab, "Properties used in clusterization.", label = "tab:aaprop"), include.rownames = FALSE, booktabs = TRUE,
+      add.to.row = list(pos = as.list(rws), command = col))
 
 # clustering of amino acids -------------------------------------------------
 

### 2015-06-16 results_analysis.R

@@ -1,0 +1,114 @@
+library(hmeasure)
+library(reshape2)
+library(seqinr)
+library(dplyr)
+library(magrittr)
+library(pbapply)
+
+load(paste0(pathway, "fold_res_MB1.RData"))
+
+source("plot_tools.R")
+
+
+perf_rep <- function(folds, threshold = 0.5)
+  do.call(rbind, lapply(1L:length(folds), function(repetition_id) {
+    res <- t(sapply(folds[[repetition_id]], function(single_fold)
+      rowMeans(sapply(single_fold, function(single_group) {
+        unlist(HMeasure(as.numeric(!is.na(single_group[, "cs_real"]) - 1),
+                        single_group[, "prob"], threshold = threshold)[["metrics"]])
+        }))))
+    
+    res <- melt(res)
+    
+    colnames(res) <- c("encoding", "measure", "value")
+    
+    res[["encoding"]] <- factor(res[["encoding"]])
+    cbind(repetition = factor(rep(repetition_id, nrow(res))), res)
+  }))
+
+
+
+
+cutoff_opt <- pblapply(c(0.5, 0.8, 0.9, 0.95, 0.99), function(i)
+  perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec")) %>%
+    group_by(encoding, measure) %>% summarise(mean_value = mean(value)) %>%
+    group_by(measure) %>% summarise(max(mean_value))
+)
+
+#0.95 the best specificity/sensitivity
+
+
+rep_res <- perf_rep(fold_res, 0.95)
+
+# rep_res %>% filter(measure %in% c("Sens")) %>% 
+#   group_by(encoding) %>% 
+#   summarise(mAUC = mean(value)) %>% 
+#   ggplot(aes(x = mAUC)) + geom_density()
+# 
+# best_encodings <- rep_res %>% filter(measure %in% c("AUC")) %>% 
+#   group_by(encoding) %>% 
+#   summarise(mAUC = mean(value)) %>% 
+#   filter(mAUC > quantile(mAUC, 0.9)) %>%
+#   droplevels %>%
+#   arrange(desc(mAUC)) %>%
+#   select(encoding) %>%
+#   unlist %>%
+#   as.character %>%
+#   as.numeric
+#   
+# 
+# 
+# p1 <- rep_res %>% filter(encoding %in% best_encodings, measure %in% c("AUC")) %>%
+#   droplevels %>% 
[…]

### 2015-06-16 paper.tex

--- 2015-06-16 paper.tex
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{reglen.png}
+\includegraphics[width=0.9\textwidth]{figures/reglen.png}
 \caption{A) Distribution of lengths of signal peptides.
 B) Distribution of the lengths of signal peptide regions.
-The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
+The measure of the length is a number of amino acids in the subsequence.
+The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
 \label{fig:reglen}

### 2015-06-16 paper.tex
adding model schemes

--- 2015-06-16 paper.tex
@@
 
-Hidden Markov Model(HMM) is a widely used machine learning technique which, in contrast to black-box methods like neural network used by SignalP, 
-is easy to interpret and allows researcher to analyze and understand structure of modeled phenomenon. 
-In case of signal-peptide, we might wish to verify whether regions really describe it. Also we might capture
-structure for some special subgroups and with HMM we can verify biological hypothesis.
-
-Suppose you have a sequence of observations e.g. amino acids. What is of interest is
+Let us first define Hidden Markov Model (HMM). Suppose you have a sequence of observations e.g. amino acids. What is of interest is
 understanding underlying cause of their occurrence. 
-HMM aims to answer that question by assuming a specific yet flexible structure of the problem.
+HMM aims to answer that by assuming a specific yet flexible structure of the problem.
 HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
@@
 complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
-However it still not flexible enough. Empirical regions length distributions (see Figure~\ref{fig:reglen}),
-is difficult to capture in that way.
+It is However still not flexible enough. Empirical regions length distributions (see figure~\ref{fig:reglen})
+are difficult to capture in that way.
 
@@
 
-The gain is not only a better fit to the data but also, as number of hidden states is smaller, easier interpretation 
+The gain is not only a better fit to the data but also, as 
+model structure (see figure \ref{fig:model_hsmm}) is straightforward, easier interpretation 
 for researcher without mathematical background.
 
-HSMM can be further extend to use some specific knowledge of motives - n-grams.
+%model structure
+\begin{figure}
+ \begin{minipage}{.5\linewidth} 
+ \caption{Scheme of signalHsmm}
+     \resizebox{6.5cm}{!}{%
+    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
+    \label{fig:model_hsmm}
+      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
+      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
+         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
+      %nodes
+      \node[main node]            (start') 	[]						{Start};	     
+      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
+      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
+      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
+      \node[main node, bottom color=green!70!black!70] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};
+
+      %lines
+      \path [line] (start')   edge node [left, color=black] {} (nregion');
+      \path [line] (nregion') edge node [below, color=black] { } (hregion');
+      \path [line] (hregion') edge node [below, color=black] { } (cregion');
+      \path [line] (cregion') edge node [left, color=black] { } (mature');
+      \draw [line] (start') to[out=340,in=200] (mature');
+    \end{tikzpicture} }
+\end{minipage}
+ \begin{minipage}{.5\linewidth}
+   \caption{Scheme of signalHsmm with cleavage site motives}
+     \resizebox{6.5cm}{!}{%
+    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
+    \label{fig:model_hsmm_kMer}
+      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
+      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
+         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
+      %nodes
+      \node[main node]            (start')   []  					{Start};	     
+      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-25mm, yshift=35mm] 	{n-region};
+      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-25mm,yshift=35mm] 	{h-region};
+      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-10mm,yshift=-25mm] 	{c-region};
+      \node[main node, bottom color=green!70!black!70] 	(mature') 	[below of=hregion',xshift=15mm, yshift=-90mm] 	{Mature protein};
+
+      \node[main node, bottom color=yellow!70!black!70]   (kmer1) 	[below of=cregion',xshift=35mm,yshift=30mm] 	{Aminoacid 1};
+      \node[main node, bottom color=yellow!70!black!70]   (kmer2) 	[below of=kmer1,xshift=-5mm,yshift=5mm] 	{Aminoacid 2};
+      \node[main node, bottom color=yellow!70!black!70]   (kmer3) 	[below of=kmer2,xshift=-5mm,yshift=5mm] 	{Aminoacid 3};
+      %lines
+      \path [line] (start')   edge node [left, color=black] {} (nregion');
+      \path [line] (nregion') edge node [below, color=black] { } (hregion');
+      \path [line] (hregion') edge node [below, color=black] { } (cregion');
+      \path [line] (cregion') to[out=230,in=90] (mature');
+      \draw [line] (start') to[out=300,in=180] (mature');
+
+      \draw [line] (cregion') edge node [below, color=black] { } (kmer1);
+      \draw [line] (kmer1) edge node [left, color=black] {} (kmer2);
+      \draw [line] (kmer2) edge node [left, color=black] {} (kmer3);
+      \draw [line] (kmer3) edge node [left, color=black] {} (mature');
+
+      %\draw [decorate,line width=3pt,decoration={brace,amplitude=10pt}] ([xshift=5mm, yshift=5mm]kmer1) -- ([yshift=-25mm]kmer3) node [black,midway,xshift=0.6cm] {\footnotesize $P_1$};
+      \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
+      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{yellow!50!black!90} k-mer};
+    \end{tikzpicture} }
+    \end{minipage}
+\end{figure}
+
+HSMM can be further extend to use some specific motives that might occur in proximity of cleavage site.
+Such an extension requires defining additional hidden states. 
+It proves to be extremely useful in recognizing cleavage site.
+Scheme can be seen in figure \ref{fig:model_hsmm_kMer}.
 

### 2015-06-16 paper.tex

@@
 The gain is not only a better fit to the data but also, as 
-model structure (see figure \ref{fig:model_hsmm}) is straightforward, easier interpretation 
+model structure (see figure~\ref{fig:model_hsmm}) is straightforward, easier interpretation 
 for researcher without mathematical background.
@@
 
-HSMM can be further extend to use some specific motives that might occur in proximity of cleavage site.
-Such an extension requires defining additional hidden states. 
-It proves to be extremely useful in recognizing cleavage site.
-Scheme can be seen in figure \ref{fig:model_hsmm_kMer}.
-
 \subsection*{n-gram extension}
 
-n-grams (k-mers) are vectors of n characters derived from input sequences.
-They can be added as the new hidden state to the hidden semi-Markov model.
-The rigid structure of n-grams may mirror cleavage sites, whose are more conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
+HSMM can be further extended to use some specific motives that might occur in proximity of cleavage site.
+Such an extension requires defining additional hidden states. 
+To find such motives we analyze n-grams (k-mers) -- vectors of n characters derived from input sequences. 
+The rigid structure of n-grams may mirror cleavage sites, which are more 
+conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
+Scheme of extended HSMM can be seen in figure~\ref{fig:model_hsmm_kMer}.
 
-Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. XXX n-grams were further used in the analysis
+Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. 
+XXX n-grams were further used in the analysis
 

### 2015-06-16 results_analysis.R
cvres figure done

@@ -92,13 +92,21 @@ best_enc <- lapply(c("AUC", "Sens", "Spec"), function(i)
 #   scale_y_continuous("Value") + 
 #   my_theme
 
-p1_dat <- rep_res %>% filter(encoding %in% unique(unlist(best_enc)), measure %in% c("AUC", "Spec", "Sens")) %>% droplevels %>%
+p1_dat <- rep_res %>% filter(#encoding %in% unique(unlist(best_enc)), 
+  measure %in% c("AUC", "Spec", "Sens")) %>% droplevels %>%
   group_by(encoding, measure) %>% summarize(mean = mean(value)) %>%
   dcast(encoding ~ measure, value.var = "mean")
 
+p1_dat <- p1_dat[!duplicated(p1_dat[, -1]), ]
+p1_dat[, "encoding"] <- rep("", nrow(p1_dat))
+p1_dat[p1_dat[, "Sens"] > 0.95, "encoding"] <- "1"
+p1_dat[p1_dat[, "Sens"] > 0.935 & p1_dat[, "Spec"] > 0.85, "encoding"] <- "2"
+p1_dat[p1_dat[, "Spec"] > 0.93, "encoding"] <- "3"
+p1_dat[, "encoding"] <- as.factor(p1_dat[, "encoding"])
+
 p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding)) +
-  geom_point(size = 5) +
-  #geom_text(size = 9, position = "jitter") +
+  geom_point(size = 5, colour = "blue", fill = adjustcolor("blue", 0.25), shape = 21) +
+  geom_text(size = 9, hjust = -0.4, vjust = -0.1) +
   scale_x_continuous("Sensitivity") +
   scale_y_continuous("Specificity\n") + 
   my_theme +
@@ -109,5 +117,6 @@ png("./figures/cvres.png", width = 2257*0.5, height = 1201*0.5)
 print(p1)
 dev.off()
 
-
-
+mean(p1_dat[p1_dat[, "encoding"] == "1", "AUC"])
+mean(p1_dat[p1_dat[, "encoding"] == "2", "AUC"])
+mean(p1_dat[p1_dat[, "encoding"] == "3", "AUC"])

### 2015-06-16 paper.tex
cvres figure done
minor changes to hsmm section

@@
 
-Let us first define Hidden Markov Model (HMM). Suppose you have a sequence of observations e.g. amino acids. What is of interest is
-understanding underlying cause of their occurrence. 
+Let us first focus on Hidden Markov Model (HMM). 
+Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
 HMM aims to answer that by assuming a specific yet flexible structure of the problem.
@@
 complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
-It is However still not flexible enough. Empirical regions length distributions (see figure~\ref{fig:reglen})
+It is, however, still not flexible enough. Empirical regional length distributions (see figure~\ref{fig:reglen})
 are difficult to capture in that way.
@@
 The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
-any given hidden state length distribution. Scheme is shown in figure~\ref{fig:hsmm}.
-Model is given by matrices $A$ and $B$ as before, as well as probabilities of duration length in hidden states.
+any given hidden state duration distribution. Scheme is shown in figure~\ref{fig:hsmm}.
+In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
 $$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
-$D$ is maximal allowed duration.
+where $D$ is maximal allowed duration.
 As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
-higher than in the regular HMM.
+higher than with the regular HMM.
 
-The gain is not only a better fit to the data but also, as 
+The advantage of HSMM model is not only a better performance but also, as 
 model structure (see figure~\ref{fig:model_hsmm}) is straightforward, easier interpretation 
@@
 
-Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. 
+Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides. 
 XXX n-grams were further used in the analysis
@@
 \includegraphics[width=0.9\textwidth]{figures/cvres.png}
-\caption{Results of cross-validation for 15 best performing amino acid encodings.}
+\caption{Results of cross-validation.
+1.
+An encoding providing the best sensitivity (AUC = 0.9337).
+2.
+A group of encodings with be best sensitivity/specificity ratio (mean AUC = 0.9650).
+3.
+An encoding providing the best specificity (AUC = 0.9686).}
 \label{fig:cvres}

### 2015-06-16 paper.tex
adding references to paper that use hsmm in bioinformatics

@@
 \subsection*{Hidden semi-Markov model}
-
-Let us first focus on Hidden Markov Model (HMM). 
+Hidden semi-Markov model (HSMM) is an extension of popular method  hidden Markov model (HMM).
+In many fields it proved to yield better accuracy of bioinformatics algorithms.
+Gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction  \citep{16571137}
+or automatic detection of heartbeats~\citep{7043102}.
+Let us first focus on hidden Markov model (HMM). 
 Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
@@
 
-\node[block,text width=3.5em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
+\node[block,text width=3em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
 \node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};

### 2015-06-17 paper.tex
fixed dangerous bug in results_analysis

@@
  \begin{minipage}{.5\linewidth} 
- \caption{Scheme of signalHsmm}
      \resizebox{6.5cm}{!}{%
@@
     \end{tikzpicture} }
+    \caption{The diagram of signalHsmm.}
+    \label{fig:signalHsmm}
 \end{minipage}
  \begin{minipage}{.5\linewidth}
-   \caption{Scheme of signalHsmm with cleavage site motives}
      \resizebox{6.5cm}{!}{%
@@
     \end{tikzpicture} }
+    \caption{Scheme of signalHsmm extended wtich the n-gram cleavage site model.}
+    \label{fig:ngramext}
     \end{minipage}
@@
 
+\subsection*{Cross-validation}
+
 \begin{figure}[ht]\centering
 \includegraphics[width=0.9\textwidth]{figures/cvres.png}
-\caption{Results of cross-validation.
-1.
-An encoding providing the best sensitivity (AUC = 0.9337).
-2.
-A group of encodings with be best sensitivity/specificity ratio (mean AUC = 0.9650).
-3.
-An encoding providing the best specificity (AUC = 0.9686).}
+\caption{}
 \label{fig:cvres}

### 2015-06-17 results_analysis.R

@@ -6,6 +6,17 @@ library(magrittr)
 library(pbapply)
 
 load(paste0(pathway, "fold_res_MB1.RData"))
+fold_res1 <- fold_res
+
+load(paste0(pathway, "fold_res_PS1.RData"))
+fold_res2 <- fold_res
+
+load(paste0(pathway, "fold_res_PS4.RData"))
+fold_res3 <- fold_res
+
+fold_res <- c(fold_res1,
+              fold_res2,
+              fold_res3)
 
 source("plot_tools.R")
 
@@ -14,9 +25,16 @@ perf_rep <- function(folds, threshold = 0.5)
   do.call(rbind, lapply(1L:length(folds), function(repetition_id) {
     res <- t(sapply(folds[[repetition_id]], function(single_fold)
       rowMeans(sapply(single_fold, function(single_group) {
-        unlist(HMeasure(as.numeric(!is.na(single_group[, "cs_real"]) - 1),
-                        single_group[, "prob"], threshold = threshold)[["metrics"]])
-        }))))
+        metrics <- unlist(HMeasure(as.numeric(!is.na(single_group[, "cs_real"])),
+                                   single_group[, "prob"], threshold = threshold)[["metrics"]])
+        TP <- metrics["TP"]
+        TN <- metrics["TN"]
+        FP <- metrics["FP"]
+        FN <- metrics["FN"]
+        
+        c(metrics, 
+          mcc = unname((TP*TN - FP*FN)/sqrt((TP + FP)*(TP + FN)*(TN + FP)*(TN + FN))))
+      }))))
     
     res <- melt(res)
     
@@ -29,8 +47,8 @@ perf_rep <- function(folds, threshold = 0.5)
 
 
 
-cutoff_opt <- pblapply(c(0.5, 0.8, 0.9, 0.95, 0.99), function(i)
-  perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec")) %>%
+cutoff_opt <- pblapply(c(0.5, 0.2, 0.1, 0.05, 0.01), function(i)
+  perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec", "mcc")) %>%
[…]

### 2015-06-17 paper.tex
added MCC

@@
 Amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
-High values of hydrophobicity are required in the h-region, the core of signal peptides.
-Alpha-helix, the secondary structure of this region, is probably induced by the positively charged n-region.
-Polarity as well as size are important especially in the cleavage site~\citep{1994palzkillselection}.
+High hydrophobicity is a determinant of the h-region, the core of signal peptides.
+Alpha-helix, the secondary structure of the h-region, is probably induced by the positively charged n-region.
+High polarity as well as smaller size are important features of the residues in the cleavage site~\citep{1994palzkillselection}.
 
@@
 We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
-Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
+We created 96 permutations of the properties, where each permutation contains only one property associated with a given criterion.
+Considering all permutations of properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
+Further, we cut the clusterings to create four-group encodings.
+
+\begin{figure}[ht]\centering
+\includegraphics[width=0.9\textwidth]{figures/cvres.png}
+\caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9346, MCC = 0.6479). 3. An encoding providing the best specificity (AUC = 0.9687, MCC = 0.8689).}
+\label{fig:cvres}
+\end{figure}
 
@@
 The advantage of HSMM model is not only a better performance but also, as 
-model structure (see figure~\ref{fig:model_hsmm}) is straightforward, easier interpretation 
+model structure (see figure~\ref{fig:hsmm}) is straightforward, easier interpretation 
 for researcher without mathematical background.
@@
 
-\begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{figures/cvres.png}
-\caption{}
-\label{fig:cvres}
-\end{figure}
-
 \section*{Discussion}

### 2015-06-17 results_analysis.R

@@ -53,10 +53,10 @@ cutoff_opt <- pblapply(c(0.5, 0.2, 0.1, 0.05, 0.01), function(i)
     group_by(measure) %>% summarise(max(mean_value))
 )
 
-#0.95 the best specificity/sensitivity
+#0.05 the best specificity/sensitivity
 
 
-rep_res <- perf_rep(fold_res[1L:2], 0.95)
+rep_res <- perf_rep(fold_res, 0.05)
 
 # rep_res %>% filter(measure %in% c("Sens")) %>% 
 #   group_by(encoding) %>% 
@@ -111,15 +111,15 @@ best_enc <- lapply(c("AUC", "Sens", "Spec"), function(i)
 #   my_theme
 
 p1_dat <- rep_res %>% filter(#encoding %in% unique(unlist(best_enc)), 
-  measure %in% c("AUC", "Spec", "Sens")) %>% droplevels %>%
+  measure %in% c("AUC", "Spec", "Sens", "mcc")) %>% droplevels %>%
   group_by(encoding, measure) %>% summarize(mean = mean(value)) %>%
   dcast(encoding ~ measure, value.var = "mean")
 
 p1_dat <- p1_dat[!duplicated(p1_dat[, -1]), ]
 p1_dat[, "encoding"] <- rep("", nrow(p1_dat))
-p1_dat[p1_dat[, "Sens"] > 0.95, "encoding"] <- "1"
-p1_dat[p1_dat[, "Sens"] > 0.94 & p1_dat[, "Spec"] > 0.85, "encoding"] <- "2"
-p1_dat[p1_dat[, "Spec"] > 0.93, "encoding"] <- "3"
+p1_dat[p1_dat[, "Spec"] > 0.955, "encoding"] <- "1"
+#p1_dat[p1_dat[, "Sens"] > 0.85 & p1_dat[, "Spec"] > 0.94, "encoding"] <- "3"
+p1_dat[p1_dat[, "Sens"] > 0.93, "encoding"] <- "2"
 p1_dat[, "encoding"] <- as.factor(p1_dat[, "encoding"])
 
 p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding)) +
@@ -127,8 +127,7 @@ p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding)) +
   geom_text(size = 9, hjust = -0.4, vjust = -0.1) +
   scale_x_continuous("Sensitivity") +
[…]

### 2015-06-17 paper.tex
improved cv section

@@
 We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
-We created 96 permutations of the properties, where each permutation contains only one property associated with a given criterion.
+We established 96 permutations of the properties, where each permutation contains only one property associated with a given criterion.
 Considering all permutations of properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
 Further, we cut the clusterings to create four-group encodings.
+As expected, some encodings (31\%) were identical.
 
@@
 We created balanced data sets by subsampling a number of proteins without a signal peptide equal to the number of proteins with signal peptide.
+The analysis was carried using also redundant encodings, but results are not shown (HSMM is a deterministic algorithm and predictions for identical groupings of amino acids were the same).
 The cross-validation was repeated XXX times, to ensure that every negative protein was tested at least once with XXX probability.
+
+As the performance measure interest we choose
 

### 2015-06-17 results_analysis.R

@@ -45,7 +45,7 @@ perf_rep <- function(folds, threshold = 0.5)
   }))
 
 
-
+# optimize cut-off
 
 cutoff_opt <- pblapply(c(0.5, 0.2, 0.1, 0.05, 0.01), function(i)
   perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec", "mcc")) %>%
@@ -122,12 +122,16 @@ p1_dat[p1_dat[, "Spec"] > 0.955, "encoding"] <- "1"
 p1_dat[p1_dat[, "Sens"] > 0.93, "encoding"] <- "2"
 p1_dat[, "encoding"] <- as.factor(p1_dat[, "encoding"])
 
-p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding)) +
-  geom_point(size = 5, colour = "blue", fill = adjustcolor("blue", 0.25), shape = 21) +
-  geom_text(size = 9, hjust = -0.4, vjust = -0.1) +
+p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding, colour = encoding == "", fill = encoding == "")) +
+  geom_point(size = 5, shape = 21) +
+  geom_text(size = 9, hjust = -0.5, vjust = 0) +
+  scale_colour_manual(values = c("red","blue")) + 
+  scale_fill_manual(values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 
   scale_x_continuous("Sensitivity") +
   scale_y_continuous("Specificity\n") + 
-  my_theme
+  my_theme +
+  guides(colour = FALSE, fill = FALSE)
+  
 
 
 png("./figures/cvres.png", width = 2257*0.5, height = 1201*0.5)

### 2015-06-17 paper.tex
added two best groups

@@
 
-\begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{figures/cvres.png}
-\caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9346, MCC = 0.6479). 3. An encoding providing the best specificity (AUC = 0.9687, MCC = 0.8689).}
-\label{fig:cvres}
-\end{figure}
-
 To compare encodings we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding.
@@
 The cross-validation was repeated XXX times, to ensure that every negative protein was tested at least once with XXX probability.
-
-As the performance measure interest we choose
 
@@
 
+\begin{figure}[ht]\centering
+\includegraphics[width=0.9\textwidth]{figures/cvres.png}
+\caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9687, MCC = 0.8689). 2. An encoding providing the best specificity (AUC = 0.9346, MCC = 0.6479).}
+\label{fig:cvres}
+\end{figure}
+
+We used four performance measures to evaluate results of cross-validation: specificity, sensitivity, Matthew's Correlation Coefficient ($\phi$ coefficient) and Area Under the Curve (AUC).
+All encodings provide very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
+The encodings have the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
+The final signalHsmm algorithm uses the encoding, that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
+
+\begin{table}[ht]
+\begin{minipage}{.5\linewidth} 
+\centering
+\begin{tabular}{l}
+  \toprule
+Groups \\ 
+  \midrule
+A, E, K, Q, R \\ 
+   \rowcolor[gray]{0.85}D, G, N, P, S, T \\ 
+  C, H, I, L, M, V \\ 
+   \rowcolor[gray]{0.85}F, W, Y \\ 
+   \bottomrule
+\end{tabular}
+\caption{Best-sensitivity (final) encoding} 
+\label{tab:best}
+\end{minipage}
+\begin{minipage}{.5\linewidth} 
+\centering
+\begin{tabular}{l}
+  \toprule
+Groups \\ 
+  \midrule
+A, C \\ 
+   \rowcolor[gray]{0.85}D, E, H, K, N, Q, R \\ 
+  G, P, S, T, Y \\ 
+   \rowcolor[gray]{0.85}F, I, L, M, V, W \\ 
+   \bottomrule
+\end{tabular}
+\caption{Best-specificity encoding} 
+\label{tab:worst}
+\end{minipage}
+\end{table}
+
 \section*{Discussion}

### 2015-06-17 results_analysis.R

@@ -117,9 +117,9 @@ p1_dat <- rep_res %>% filter(#encoding %in% unique(unlist(best_enc)),
 
 p1_dat <- p1_dat[!duplicated(p1_dat[, -1]), ]
 p1_dat[, "encoding"] <- rep("", nrow(p1_dat))
-p1_dat[p1_dat[, "Spec"] > 0.955, "encoding"] <- "1"
+p1_dat[p1_dat[, "Spec"] > 0.955, "encoding"] <- "2"
 #p1_dat[p1_dat[, "Sens"] > 0.85 & p1_dat[, "Spec"] > 0.94, "encoding"] <- "3"
-p1_dat[p1_dat[, "Sens"] > 0.93, "encoding"] <- "2"
+p1_dat[p1_dat[, "Sens"] > 0.93, "encoding"] <- "1"
 p1_dat[, "encoding"] <- as.factor(p1_dat[, "encoding"])
 
 p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding, colour = encoding == "", fill = encoding == "")) +
@@ -127,8 +127,8 @@ p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding, colour = encoding
   geom_text(size = 9, hjust = -0.5, vjust = 0) +
   scale_colour_manual(values = c("red","blue")) + 
   scale_fill_manual(values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 
-  scale_x_continuous("Sensitivity") +
-  scale_y_continuous("Specificity\n") + 
+  scale_x_continuous("Sensitivity\n") +
+  scale_y_continuous("Specificity") + 
   my_theme +
   guides(colour = FALSE, fill = FALSE)
   
@@ -139,13 +139,37 @@ print(p1)
 dev.off()
 
 #caption for cvres
-paste0("Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = ", 
+paste0("Results of cross-validation. 1. The encoding providing the best sensitivity (AUC = ", 
        round(mean(p1_dat[p1_dat[, "encoding"] == "1", "AUC"]), 4),
        ", MCC = ", 
        round(mean(p1_dat[p1_dat[, "encoding"] == "1", "mcc"]), 4),
-       "). 3. An encoding providing the best specificity (AUC = ", 
+       "). 2. The encoding providing the best specificity (AUC = ", 
[…]

### 2015-06-17 paper.tex
adding info in signalHSMM diagram

--- 2015-06-17 paper.tex
@@
 \subsection*{Hidden semi-Markov model}
-Hidden semi-Markov model (HSMM) is an extension of popular method  hidden Markov model (HMM).
-In many fields it proved to yield better accuracy of bioinformatics algorithms.
-Gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction  \citep{16571137}
+Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM).
+As such, it proved to yield better performance in many fields of bioinformatics including
+gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction~\citep{16571137}
 or automatic detection of heartbeats~\citep{7043102}.
-Let us first focus on hidden Markov model (HMM). 
-Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
-HMM aims to answer that by assuming a specific yet flexible structure of the problem.
+Let us first briefly describe the idea that lies behind HMM. 
+Suppose you have a sequence of observations e.g. amino acids, and you are interested in understanding underlying cause of their occurrence. 
+HMM aims to answer that question by assuming a specific, yet flexible, structure of the problem.
 HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
-These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
+They are "the cause" of the observations. At every step $t$, hidden state might change according to transition matrix
 $A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
@@
 
+\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
+(1,-1.8) -- (5,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_1$}};
+
+\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
+(8,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_k$}};
+
 \draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
-(1,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
+(1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
 
 \end{tikzpicture}
-\caption{Hidden semi-Markov Model scheme}
+\caption{General scheme of hidden semi-Markov model}
 \label{fig:hsmm}
@@
 The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
-any given hidden state duration distribution. Scheme is shown in figure~\ref{fig:hsmm}.
+any given hidden state duration distribution. Diagram of HSMM is presented in figure~\ref{fig:hsmm}.
 In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
 $$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
-where $D$ is maximal allowed duration.
+where $D$ is maximum allowed duration.
 As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
@@
 
-The advantage of HSMM model is not only a better performance but also, as 
-model structure (see figure~\ref{fig:hsmm}) is straightforward, easier interpretation 
-for researcher without mathematical background.
+The model we use has a very specific structure. Hidden states are signal peptide regions.
+Almost all entries in transition matrix $A$ are zeros, because regions are sequential.
+Possible transitions are depicted as arrows in figure~\ref{fig:signalHsmm}.
+Probabilities of observations for hidden states and hidden states durations were estimated from training data.
+The advantage of HSMM model is not only a better performance but also its straightforwardness.
+Figure~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.
 
@@
     \end{tikzpicture} }
-    \caption{Scheme of signalHsmm extended wtich the n-gram cleavage site model.}
+    \caption{The diagram of signalHsmm extended with the n-gram cleavage site model.}
     \label{fig:ngramext}
@@
 conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
-Scheme of extended HSMM can be seen in figure~\ref{fig:model_hsmm_kMer}.
+Scheme of extended HSMM can be seen in figure~\ref{fig:ngramext}.
 

### 2015-06-17 results_analysis.R
addded enccomp figure

@@ -149,11 +149,12 @@ paste0("Results of cross-validation. 1. The encoding providing the best sensitiv
        round(mean(p1_dat[p1_dat[, "encoding"] == "2", "mcc"]), 4),
        ").")
 
-#interesting encodings
+#interesting encodings -------------------------------------
 int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
 all_groups[int_enc]
-all_traits_combn[int_enc, ]
 
+
+# tables of groups for interesting encodings --------------------------------
 group2df <- function(group_list, caption = NULL, label = NULL) {
   tab <- data.frame(Groups = sapply(group_list, function(i)
     paste0(toupper(sort(i)), collapse = ", ")))
@@ -168,8 +169,47 @@ cat(group2df(all_groups[int_enc][[2]],
              "tab:best"))
 
 cat(group2df(all_groups[int_enc][[1]],
-             "Best-specificity encoding",
+             "Worst-sensitivity encoding",
              "tab:worst"))
 
+#figure comparing encodings
+data(aaindex)
+group_properties <- function(group) {
+  res <- do.call(rbind, lapply(1L:length(group), function(subgroup_id)
+    melt(data.frame(group = paste0("Group ", as.character(rep(subgroup_id, 4))), critertion = c("size", "hydroph", "polarity", "alpha"),
+                    aa_nvals[unlist(all_traits_combn[int_enc[2], ]), group[[subgroup_id]]]))))
+  levels(res[["variable"]]) <- toupper(levels(res[["variable"]]))
+  res
+}
+
+dat_best <- group_properties(all_groups[int_enc][[2]])
+group_worst <- all_groups[int_enc][[1]]
+#re-arrange group for better comparision
+group_worst <- group_worst[c(2, 3, 4, 1)]
+
+dat_worst <- group_properties(group_worst)
+
+p1 <- ggplot(dat_best, aes(x = critertion, y = value, label = variable)) +
+  geom_point(size = 5, shape = 21, colour = "blue", fill = adjustcolor("blue", 0.25)) +
+  #geom_text(hjust = -1) +
+  facet_wrap(~group, ncol = 4) +
+  scale_x_discrete("", labels = c("size" = "Size","hydroph" = "Hydroph.",
[…]

### 2015-06-17 results_analysis.R
commit before dinner

@@ -153,6 +153,9 @@ paste0("Results of cross-validation. 1. The encoding providing the best sensitiv
 int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
 all_groups[int_enc]
 
+group_worst <- all_groups[int_enc][[1]]
+#re-arrange group for better comparision
+group_worst <- group_worst[c(2, 3, 4, 1)]
 
 # tables of groups for interesting encodings --------------------------------
 group2df <- function(group_list, caption = NULL, label = NULL) {
@@ -168,7 +171,7 @@ cat(group2df(all_groups[int_enc][[2]],
              "Best-sensitivity (final) encoding",
              "tab:best"))
 
-cat(group2df(all_groups[int_enc][[1]],
+cat(group2df(group_worst,
              "Worst-sensitivity encoding",
              "tab:worst"))
 
@@ -183,10 +186,6 @@ group_properties <- function(group) {
 }
 
 dat_best <- group_properties(all_groups[int_enc][[2]])
-group_worst <- all_groups[int_enc][[1]]
-#re-arrange group for better comparision
-group_worst <- group_worst[c(2, 3, 4, 1)]
-
 dat_worst <- group_properties(group_worst)
 
 p1 <- ggplot(dat_best, aes(x = critertion, y = value, label = variable)) +

### 2015-06-17 results_analysis.R
re-arranged the worst group

@@ -151,7 +151,10 @@ paste0("Results of cross-validation. 1. The encoding providing the best sensitiv
 
 #interesting encodings -------------------------------------
 int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
-all_groups[int_enc]
+group_worst <- all_groups[int_enc][[1]]
+#re-arrange group for better comparision
+group_worst <- group_worst[c(2, 3, 4, 1)]
+
 
 
 # tables of groups for interesting encodings --------------------------------
@@ -168,7 +171,7 @@ cat(group2df(all_groups[int_enc][[2]],
              "Best-sensitivity (final) encoding",
              "tab:best"))
 
-cat(group2df(all_groups[int_enc][[1]],
+cat(group2df(group_worst,
              "Worst-sensitivity encoding",
              "tab:worst"))
 
@@ -183,9 +186,6 @@ group_properties <- function(group) {
 }
 
 dat_best <- group_properties(all_groups[int_enc][[2]])
-group_worst <- all_groups[int_enc][[1]]
-#re-arrange group for better comparision
-group_worst <- group_worst[c(2, 3, 4, 1)]
 
 dat_worst <- group_properties(group_worst)
 

### 2015-06-17 paper.tex
addded enccomp figure

@@
 
-\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
 (1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
@@
 \end{tikzpicture}
-\caption{General scheme of hidden semi-Markov model}
+\caption{General scheme of hidden semi-Markov model.}
 \label{fig:hsmm}
@@
 \end{tabular}
-\caption{Best-sensitivity (final) encoding} 
+\caption{Best-sensitivity (final) encoding.} 
 \label{tab:best}
@@
   \midrule
-A, C \\ 
-   \rowcolor[gray]{0.85}D, E, H, K, N, Q, R \\ 
-  G, P, S, T, Y \\ 
-   \rowcolor[gray]{0.85}F, I, L, M, V, W \\ 
+D, E, H, K, N, Q, R \\ 
+   \rowcolor[gray]{0.85}G, P, S, T, Y \\ 
+  F, I, L, M, V, W \\ 
+   \rowcolor[gray]{0.85}A, C \\ 
    \bottomrule
 \end{tabular}
-\caption{Best-specificity encoding} 
+\caption{Worst-sensitivity encoding.} 
 \label{tab:worst}
@@
 \end{table}
+
+\begin{figure}[ht]\centering
+\includegraphics[width=0.9\textwidth]{figures/enccomp.png}
+\caption{A comparision of encodings.}
+\label{fig:enccomp}
+\end{figure}
 

### 2015-06-17 results_analysis.R
slightly improved descriptions

@@ -215,3 +215,5 @@ print(arrangeGrob(textGrob("A", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p1,
                   textGrob("B", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p2, 
                   nrow = 2, ncol = 2, widths = c(0.05, 0.95)))
 dev.off()
+
+

### 2015-06-17 encodings_vs_regions.R
improved comparision of encodings

@@ -1,0 +1,141 @@
+library(reshape2)
+library(signalHsmm)
+library(biogram)
+
+#distributions of lengths of dignal peptides regions
+
+if(Sys.info()["nodename"] == "MICHALKOMP" )
+  pathway <- "C:/Users/Michal/Dropbox/signal-peptide2_data/"
+
+if(Sys.info()["nodename"] == "phobos" )
+  pathway <- "/home/michal/Dropbox/signal-peptide2_data/"
+
+
+pos_seqs <- read_uniprot(paste0(pathway, "signal_peptides.txt"), euk = TRUE, what = "signal")
+
+#interesting encodings -------------------------------------
+#p1_dat comes from results_analysis.R
+int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
+group_worst <- all_groups[int_enc][[1]]
+#re-arrange group for better comparision
+group_worst <- group_worst[c(2, 3, 4, 1)]
+names(group_worst) <- 1L:4
+
+group_best <- all_groups[int_enc][[2]]
+
+nhc_borders <- t(sapply(pos_seqs, find_nhc))
+
+deg_pos <- lapply(list(group_best, group_worst), function(single_encoding)
+  lapply(1L:length(pos_seqs), function(seq_id) {
+    single_seq <- pos_seqs[[seq_id]]
+    aa_n <- single_seq[1L:(nhc_borders[seq_id, "start_h"] - 1)]
+    aa_h <- single_seq[nhc_borders[seq_id, "start_h"]:(nhc_borders[seq_id, "start_c"] - 1)]
+    aa_c <- single_seq[nhc_borders[seq_id, "start_c"]:(nhc_borders[seq_id, "cs"])]
+    aa_mature <- single_seq[(nhc_borders[seq_id, "cs"] + 1):length(single_seq)]
+    res <- list(aa_n = aa_n,
+                aa_h = aa_h,
+                aa_c = aa_c,
+                aa_m = aa_mature)
+    lapply(res, function(i) 
+      degenerate(i, single_encoding))
+  }))
+
+deg_region <- do.call(rbind, lapply(deg_pos, function(single_encoding)
+  cbind(region = unlist(lapply(c("n", "h", "c", "m"), function(i) rep(i, 4))), 
+        do.call(rbind, lapply(1L:4, function(single_region) {
+          res <- factor(unlist(lapply(single_encoding, function(single_seq)
+            single_seq[[single_region]])), levels = as.character(1L:4))
[…]

### 2015-06-17 results_analysis.R

@@ -156,64 +156,5 @@ group_worst <- all_groups[int_enc][[1]]
 group_worst <- group_worst[c(2, 3, 4, 1)]
 
 
-group_worst <- all_groups[int_enc][[1]]
-#re-arrange group for better comparision
-group_worst <- group_worst[c(2, 3, 4, 1)]
-
-# tables of groups for interesting encodings --------------------------------
-group2df <- function(group_list, caption = NULL, label = NULL) {
-  tab <- data.frame(Groups = sapply(group_list, function(i)
-    paste0(toupper(sort(i)), collapse = ", ")))
-  rws <- seq(1, nrow(tab) - 1, by = 2)
-  col <- rep("\\rowcolor[gray]{0.85}", length(rws))
-  print(xtable(tab, caption = caption, label = label), include.rownames = FALSE, booktabs = TRUE,
-        add.to.row = list(pos = as.list(rws), command = col), print.results = FALSE)
-}
-
-cat(group2df(all_groups[int_enc][[2]],
-             "Best-sensitivity (final) encoding",
-             "tab:best"))
-
-cat(group2df(group_worst,
-             "Worst-sensitivity encoding",
-             "tab:worst"))
-
-#figure comparing encodings
-data(aaindex)
-group_properties <- function(group) {
-  res <- do.call(rbind, lapply(1L:length(group), function(subgroup_id)
-    melt(data.frame(group = paste0("Group ", as.character(rep(subgroup_id, 4))), critertion = c("size", "hydroph", "polarity", "alpha"),
-                    aa_nvals[unlist(all_traits_combn[int_enc[2], ]), group[[subgroup_id]]]))))
-  levels(res[["variable"]]) <- toupper(levels(res[["variable"]]))
-  res
-}
-
-dat_best <- group_properties(all_groups[int_enc][[2]])
-dat_worst <- group_properties(group_worst)
-
-p1 <- ggplot(dat_best, aes(x = critertion, y = value, label = variable)) +
-  geom_point(size = 5, shape = 21, colour = "blue", fill = adjustcolor("blue", 0.25)) +
-  #geom_text(hjust = -1) +
-  facet_wrap(~group, ncol = 4) +
-  scale_x_discrete("", labels = c("size" = "Size","hydroph" = "Hydroph.",
[…]

### 2015-06-17 encodings_vs_regions.R
improved plot

@@ -74,7 +74,7 @@ p3 <- ggplot(deg_region, aes(x = group, y = freq, fill = enc, colour = enc)) +
 data(aaindex)
 group_properties <- function(group) {
   res <- do.call(rbind, lapply(1L:length(group), function(subgroup_id)
-    melt(data.frame(group = paste0("Group ", as.character(rep(subgroup_id, 4))), critertion = c("size", "hydroph", "polarity", "alpha"),
+    melt(data.frame(group = as.character(rep(subgroup_id, 4)), critertion = c("size", "hydroph", "polarity", "alpha"),
                     aa_nvals[unlist(all_traits_combn[int_enc[2], ]), group[[subgroup_id]]]))))
   levels(res[["variable"]]) <- toupper(levels(res[["variable"]]))
   res
@@ -104,12 +104,20 @@ dat_bestworst <- cbind(enc = unlist(lapply(c("best", "worst"), function(i) rep(i
 #   scale_y_continuous("Value") + 
 #   my_theme
 
-p1 <- ggplot(dat_bestworst, aes(x = critertion, y = value, col = enc, fill = enc)) +
+facet_names <- list("size" = "Size",
+                    "hydroph" = "Hydrophobicity",
+                    "polarity" = "Polarity",
+                    "alpha" = expression(paste(alpha, "-helix")))
+
+facet_labeller <- function(variable, value) 
+  facet_names[value]
+
+
+p1 <- ggplot(dat_bestworst, aes(x = group, y = value, col = enc, fill = enc)) +
   geom_point(size = 5, shape = 21, position = position_dodge(width=0.5)) +
   #geom_text(hjust = -1) +
-  facet_grid(~group) +
-  scale_x_discrete("Criterion\n", labels = c("size" = "Size","hydroph" = "Hydroph.",
-                                  "polarity" = "Polarity","alpha" = expression(paste(alpha, "-helix")))) +
+  facet_grid(~critertion, labeller = facet_labeller) +
+  scale_x_discrete("Group\n") +
   scale_y_continuous("Value") + 
   scale_colour_manual("Encoding: ", values = c("red", "blue")) +
   scale_fill_manual("Encoding: ", values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 

### 2015-06-17 encodings_vs_regions.R
wasted too much time on enccomp figure

@@ -58,15 +58,17 @@ deg_region[["region"]] <- factor(deg_region[["region"]], levels = c("n", "h", "c
 
 source("plot_tools.R")
 
-levels(deg_region[["region"]]) <- c("n-region", "h-region", "c-region", "Mature protein")
+levels(deg_region[["region"]]) <- c("n-region", "h-region", "c-region", "Mature\nprotein")
+levels(deg_region[["group"]]) <- paste0("Group ", levels(deg_region[["group"]]))
 
-p3 <- ggplot(deg_region, aes(x = group, y = freq, fill = enc, colour = enc)) +
+p3 <- ggplot(deg_region, aes(x = region, y = freq, fill = enc, colour = enc)) +
   geom_bar(stat = "identity", position = "dodge") + 
-  facet_wrap(~ region, nrow = 1) + 
+  facet_wrap(~ group, nrow = 1) + 
   scale_y_continuous("Frequency") +
-  scale_x_discrete("Group\n") +
+  scale_x_discrete("Region\n") +
   scale_colour_manual("Encoding: ", values = c("red", "blue")) +
   scale_fill_manual("Encoding: ", values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 
+  guides(colour = FALSE) +
   my_theme
 
 
@@ -74,7 +76,7 @@ p3 <- ggplot(deg_region, aes(x = group, y = freq, fill = enc, colour = enc)) +
 data(aaindex)
 group_properties <- function(group) {
   res <- do.call(rbind, lapply(1L:length(group), function(subgroup_id)
-    melt(data.frame(group = as.character(rep(subgroup_id, 4)), critertion = c("size", "hydroph", "polarity", "alpha"),
+    melt(data.frame(group = paste0("Group ", as.character(rep(subgroup_id, 4))), critertion = c("size", "hydroph", "polarity", "alpha"),
                     aa_nvals[unlist(all_traits_combn[int_enc[2], ]), group[[subgroup_id]]]))))
   levels(res[["variable"]]) <- toupper(levels(res[["variable"]]))
   res
@@ -104,20 +106,12 @@ dat_bestworst <- cbind(enc = unlist(lapply(c("best", "worst"), function(i) rep(i
 #   scale_y_continuous("Value") + 
 #   my_theme
 
-facet_names <- list("size" = "Size",
-                    "hydroph" = "Hydrophobicity",
-                    "polarity" = "Polarity",
[…]

### 2015-06-17 results_analysis.R

@@ -47,11 +47,11 @@ perf_rep <- function(folds, threshold = 0.5)
 
 # optimize cut-off
 
-cutoff_opt <- pblapply(c(0.5, 0.2, 0.1, 0.05, 0.01), function(i)
-  perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec", "mcc")) %>%
-    group_by(encoding, measure) %>% summarise(mean_value = mean(value)) %>%
-    group_by(measure) %>% summarise(max(mean_value))
-)
+# cutoff_opt <- pblapply(c(0.5, 0.2, 0.1, 0.05, 0.01), function(i)
+#   perf_rep(fold_res, threshold = i) %>% filter(measure %in% c("AUC", "Sens", "Spec", "mcc")) %>%
+#     group_by(encoding, measure) %>% summarise(mean_value = mean(value)) %>%
+#     group_by(measure) %>% summarise(max(mean_value))
+# )
 
 #0.05 the best specificity/sensitivity
 
@@ -149,11 +149,7 @@ paste0("Results of cross-validation. 1. The encoding providing the best sensitiv
        round(mean(p1_dat[p1_dat[, "encoding"] == "2", "mcc"]), 4),
        ").")
 
-#interesting encodings -------------------------------------
-int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
-group_worst <- all_groups[int_enc][[1]]
-#re-arrange group for better comparision
-group_worst <- group_worst[c(2, 3, 4, 1)]
+
 
 
 

### 2015-06-17 encodings_vs_regions.R
fixed awful msitake - confused the best and the worst encodings

@@ -16,12 +16,12 @@ pos_seqs <- read_uniprot(paste0(pathway, "signal_peptides.txt"), euk = TRUE, wha
 #interesting encodings -------------------------------------
 #p1_dat comes from results_analysis.R
 int_enc <- as.numeric(rownames(p1_dat[p1_dat[, "encoding"] != "", ]))
-group_worst <- all_groups[int_enc][[1]]
+group_best <- all_groups[int_enc][[1]]
 #re-arrange group for better comparision
-group_worst <- group_worst[c(2, 3, 4, 1)]
-names(group_worst) <- 1L:4
+group_best <- group_best[c(2, 3, 4, 1)]
+names(group_best) <- 1L:4
 
-group_best <- all_groups[int_enc][[2]]
+group_worst <- all_groups[int_enc][[2]]
 
 nhc_borders <- t(sapply(pos_seqs, find_nhc))
 

### 2015-06-17 results_analysis.R

@@ -151,6 +151,6 @@ paste0("Results of cross-validation. 1. The encoding providing the best sensitiv
 
 
 
-
+rep_res
 
 

### 2015-06-17 benchmark_signalHsmm.R
added benchmark

@@ -1,0 +1,26 @@
+library(dplyr)
+library(biogram)
+
+other_soft <- read.csv2("benchmark_other_new.csv")
+
+signalHsmm2010 <- train_hsmm(read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal"), 
+                             aa_group = group_best)
+signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = "signal"), 
+                             aa_group = group_best)
+benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
+benchmark_data2 <- lapply(benchmark_data, function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
+
+real_labels <- c(rep(1, 218), rep(0, 218))
+
+all_preds <- data.frame(other_soft,
+                       signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],
+                       signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_data2))[["sp.probability"]])
+
+bench_metrics <- HMeasure(real_labels, all_preds,
+                          threshold = c(rep(0.5, 5), 0.05, 0.05))[["metrics"]]
+
+all_preds[["signalHsmm2010"]] <- all_preds[["signalHsmm2010"]] > 0.05
+all_preds[["signalHsmm1989"]] <- all_preds[["signalHsmm1989"]] > 0.05
+
+FN_preds <- all_preds[1L:218, ]
+rownames(FN_preds[!FN_preds[["signalHsmm2010"]] | !FN_preds[["signalHsmm1989"]], c("signalHsmm2010", "signalHsmm1989")])

### 2015-06-17 benchmark_signalHsmm.R
added best encoding

@@ -3,6 +3,10 @@ library(biogram)
 
 other_soft <- read.csv2("benchmark_other_new.csv")
 
+group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
+     `2` = c("g", "p", "s", "t", "y"), `3` = c("i", "l", "m", "f", "w", "v"), 
+     `4` = c("a", "c"))
+
 signalHsmm2010 <- train_hsmm(read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal"), 
                              aa_group = group_best)
 signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = "signal"), 

### 2015-06-17 benchmark_signalHsmm.R
benchmark with k-mer states

@@ -1,5 +1,8 @@
 library(dplyr)
 library(biogram)
+library(signalHsmm)
+library(seqinr)
+library(hmeasure)
 
 other_soft <- read.csv2("benchmark_other_new.csv")
 
@@ -21,7 +24,7 @@ all_preds <- data.frame(other_soft,
                        signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_data2))[["sp.probability"]])
 
 bench_metrics <- HMeasure(real_labels, all_preds,
-                          threshold = c(rep(0.5, 5), 0.05, 0.05))[["metrics"]]
+                          threshold = c(rep(0.5, 5), 0.1, 0.1))[["metrics"]]
 
 all_preds[["signalHsmm2010"]] <- all_preds[["signalHsmm2010"]] > 0.05
 all_preds[["signalHsmm1989"]] <- all_preds[["signalHsmm1989"]] > 0.05

### 2015-06-17 benchmark_with_k_mer_states.R

@@ -1,0 +1,176 @@
+
+signalHSMM_proteins <- read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal")
+b <- lapply(signalHSMM_proteins, function(x){
+  cut <- attr(x, "sig")[2]
+  deg_sample <- as.numeric(degenerate(toupper(x)[1L:50], group_best))
+  #remove atypical amino acids
+  deg_sample <- na.omit(deg_sample)
+  deg_sample[(cut-2):(cut+2)]
+})
+
+b <- matrix(unlist(b), ncol=5, byrow = TRUE)
+# print.xtable(xtable(b, digits = 0), type = "html", include.rownames = FALSE)
+# kable(b)
+(table(apply(b, 1, function(x) paste0(x, collapse=""))))
+(table(apply(b, 1, function(x) paste0(x[1:4], collapse=""))))
+(table(apply(b, 1, function(x) paste0(x[4:1], collapse=""))))
+(table(apply(b, 1, function(x) paste0(x[1:3], collapse=""))))
+(table(apply(b, 1, function(x) paste0(x[3:1], collapse=""))))
+
+sum(b[,1]==2 & b[,3] %in% c(2,4))/length(signalHSMM_proteins)
+sum(b[,1]==4 & b[,3] %in% c(2,4))/length(signalHSMM_proteins)
+sum(b[,1]%in%c(4) & b[,2] %in% c(1,2,3,4) & b[,3] %in% c(2,4))/length(signalHSMM_proteins)
+sum(b[,1]%in%c(2,4) & b[,3] %in% c(4))/length(signalHSMM_proteins)
+sum(b[,1]%in%c(3) & b[,2] %in% c(1,2,3) & b[,3]%in%c(2))/length(signalHSMM_proteins)
+
+sum(b[,2]%in%c(1,2) & b[,3] %in% c(2,4) & b[,4]==3)/length(signalHSMM_proteins)
+sum(b[,1]%in%c(2,3,4) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4) & b[,4]%in%c(1))/length(signalHSMM_proteins)
+
+sum(b[,1]%in%c(2) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2))/length(signalHSMM_proteins)
+
+
[…]

### 2015-06-18 paper.tex
slightly improved descriptions
improved comparision of encodings
fixed awful msitake - confused the best and the worst encodings

@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{figures/reglen.png}
+\includegraphics[width=0.95\textwidth]{figures/reglen.png}
 \caption{A) Distribution of lengths of signal peptides.
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{figures/cvres.png}
+\includegraphics[width=0.95\textwidth]{figures/cvres.png}
 \caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9687, MCC = 0.8689). 2. An encoding providing the best specificity (AUC = 0.9346, MCC = 0.6479).}
@@
 
+The best and the worst encoding share some similarities.
+The both groupings have strongly pronounced highly hydrophobic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}.
+
 \begin{table}[ht]
+\begin{minipage}{.5\linewidth} 
+\centering
+\begin{tabular}{l}
+  \toprule
+Groups \\ 
+  \midrule
+D, E, H, K, N, Q, R \\ 
+   \rowcolor[gray]{0.85}G, P, S, T, Y \\ 
+  F, I, L, M, V, W \\ 
+   \rowcolor[gray]{0.85}A, C \\ 
+   \bottomrule
+\end{tabular}
+\caption{The best sensitivity (final) encoding.} 
+\label{tab:best}
+\end{minipage}
 \begin{minipage}{.5\linewidth} 
@@
 \end{tabular}
-\caption{Best-sensitivity (final) encoding.} 
-\label{tab:best}
-\end{minipage}
-\begin{minipage}{.5\linewidth} 
-\centering
-\begin{tabular}{l}
-  \toprule
-Groups \\ 
-  \midrule
-D, E, H, K, N, Q, R \\ 
-   \rowcolor[gray]{0.85}G, P, S, T, Y \\ 
-  F, I, L, M, V, W \\ 
-   \rowcolor[gray]{0.85}A, C \\ 
-   \bottomrule
-\end{tabular}
-\caption{Worst-sensitivity encoding.} 
+\caption{The worst sensitivity encoding.} 
 \label{tab:worst}
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.9\textwidth]{figures/enccomp.png}
-\caption{A comparision of encodings.}
+\includegraphics[width=0.95\textwidth]{figures/enccomp.png}
+\caption{A comparision of encodings.
+A) Properties of amino acids in the best and the worst sensitivity encoding.
+B) Frequencies of amino acids beloning to the encodings in regions of signal peptide and mature protein.}
 \label{fig:enccomp}

### 2015-06-18 paper.tex
updated code and dataset in fasta file

@@
 
-The best and the worst encoding share some similarities.
-The both groupings have strongly pronounced highly hydrophobic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}.
+The best and the worst encoding share some similarities. In both cases Group 1 tends to contain polar
+
+The both groupings have strongly pronounced non-polar and highly hydrophobic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}) containing isoleucine, leucine, methionine and valine.
 

### 2015-06-18 benchmark_signalHsmm.R

@@ -17,7 +17,7 @@ signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = 
 benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
 benchmark_data2 <- lapply(benchmark_data, function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
 
-real_labels <- c(rep(1, 218), rep(0, 218))
+real_labels <- c(rep(1, 214), rep(0, 218))
 
 all_preds <- data.frame(other_soft,
                        signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],

### 2015-06-18 paper.tex
added rambling about encodings

--- 2015-06-18 paper.tex
@@
 
-We used four performance measures to evaluate results of cross-validation: specificity, sensitivity, Matthew's Correlation Coefficient ($\phi$ coefficient) and Area Under the Curve (AUC).
-All encodings provide very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
-The encodings have the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
-The final signalHsmm algorithm uses the encoding, that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
-
-The best and the worst encoding share some similarities.
-The both groupings have strongly pronounced highly hydrophobic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}.
-
 \begin{table}[ht]
@@
 A) Properties of amino acids in the best and the worst sensitivity encoding.
-B) Frequencies of amino acids beloning to the encodings in regions of signal peptide and mature protein.}
+B) Frequencies of amino acids beloning to the encodings in regions of signal peptide and mature proteins.}
 \label{fig:enccomp}
 \end{figure}
+
+We used four performance measures to evaluate results of cross-validation: specificity, sensitivity, Matthew's Correlation Coefficient ($\phi$ coefficient) and Area Under the Curve (AUC).
+All encodings provide very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
+The encodings have the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
+The final signalHsmm algorithm uses the encoding, that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
+
+The best and the worst encoding share some similarities.
+In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals.
+In the best encoding, all charged amino acids, both acids and even weak bases as histidine, also belong to this group.
+This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding.
+In the worst encoding, where polar and charged character of the Group 1 is not that explicit, the difference in frequency between regions is also less visible.
+
+The amino acids belonging to the Group 2 have quite low probability of occuring in an alpha-helix, but are still very diverse, but all .
+Polar but uncharged serine and threonine neighbour with alyphatic glycine and proline.
+In the worst encoding the polar character of the group is emphasized by the addition of asparagine.
+The best grouping contains also highly hydrophobic tyrosine.
+Despite these differences, the frequency of Group 2 seems to be comparable between two encodings.
+
+The both groupings have strongly non-polar and aliphatic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}) containing isoleucine, leucine, methionine and valine.
+Hyrophobic properties of this group in the best encoding are even more pronounced by the presence of tryptophan and phenylalanine.
+On the contrary, to the Group 3 in the worst encoding belong also more polar histidine.
+Hence the hydrophobic character of this group, it is dominant over the h-region in case of both amino acid classifications.
+
+The fourth group is the most diverse among both encodings.
+The key is obvious in a case of the worst encoding.
+Phenylalanine, tryptophan and tyrosine are aromatic, highly hydrophobic amino acids.
+In contrast, the Group 4 in the best encoding contains only alanine and cysteine, which both are rather small amino acids and tend to appear in alpha-helices.
+This very unique composition seems to be typical for the c-region of signal peptide.
 

### 2015-06-18 paper.tex
commit before dinner

@@
 
-The best and the worst encoding share some similarities.
+\subsection*{Comparision of encodings}
+
+We examined more deeply encodings with the best and the worst sensitivity.
 In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals.
@@
 
+The first set of predictive rules derived by our method lies in the encoding of amino acids.
+Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also found new relationships.
+We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins.
+The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.
+
 The proposed algorithm recognizing various signal peptides will reduce costs and speed up searching appropriate artificial signal peptides designed for protein secretion in expensive and time-consuming laboratory experiments.

### 2015-06-18 paper.tex
added results from all repetitions

@@
 The analysis was carried using also redundant encodings, but results are not shown (HSMM is a deterministic algorithm and predictions for identical groupings of amino acids were the same).
-The cross-validation was repeated XXX times, to ensure that every negative protein was tested at least once with XXX probability.
+The cross-validation was repeated 60 times.
+At first, we planned 160 repetitions to ensure with probability 0.95, that every protein without signal peptide was in the learning set, but very small variance of performance measure allowed us to relax this condition.
 
@@
 \includegraphics[width=0.95\textwidth]{figures/cvres.png}
-\caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9687, MCC = 0.8689). 2. An encoding providing the best specificity (AUC = 0.9346, MCC = 0.6479).}
+\caption{Results of cross-validation. 1. The encoding providing the best sensitivity (AUC = 0.9683, MCC = 0.8677). 2. The encoding providing the best specificity (AUC = 0.9338, MCC = 0.6474).}
 \label{fig:cvres}
@@
 
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrr}
+  \toprule
+Measure & Mean & SD \\ 
+  \midrule
+AUC & 0.9683 & 0.0024 \\ 
+   \rowcolor[gray]{0.85}MCC & 0.8677 & 0.0049 \\ 
+  Sensitivity & 0.9406 & 0.0008 \\ 
+   \rowcolor[gray]{0.85}Specificity & 0.9269 & 0.0050 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Performance measures for the best encoding. 60 repetitions of cross-validation.} 
+\label{tab:perfmeas}
+\end{table}
+
 \subsection*{Comparision of encodings}

### 2015-06-18 paper.tex
added some discussion

@@
 Henceforth, the usage of raw amino acid sequences is superfluous and introduces unnecessary information.
-To utilize this property of signal peptide recognition, we cluster amino acids into several groups based on the physicochemical properties of residues.
+To utilize this property of signal peptide recognition, we cluster amino acids into several groups based on the physicochemical features of residues.
 
@@
 
-Amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
+It is worth to note, that signalHsmm is not a first software to use amino acid encoding in signal peptide prediction.
+BLOMAP~\citep{maetschke2005blomap} also employed similar strategy, but considered only encodings based on the substitution matrices.
+Substitution matrices are computed using protein sequences and knowning how atypical signal peptides are, we had chosen different approach.
+
+Instead amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
 High hydrophobicity is a determinant of the h-region, the core of signal peptides.
@@
   \midrule
-Size & Size \\ 
-   \rowcolor[gray]{0.85}Size & Molecular weight \\ 
+Size & Size~\citep{dawson1972size} \\ 
+   \rowcolor[gray]{0.85}Size & Molecular weight~\citep{fasman1976handbook}\\ 
   Size & Residue volume~\citep{1973goldsackcontribution} \\ 
@@
   Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
-   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix \\ 
+   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
   Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
@@
 
-We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Table~\ref{tab:aaprop}), each attributed to a single criterion.
+We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Tab.~\ref{tab:aaprop}), each attributed to a single criterion.
 We established 96 permutations of the properties, where each permutation contains only one property associated with a given criterion.
@@
 complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
-It is, however, still not flexible enough. Empirical regional length distributions (see figure~\ref{fig:reglen})
+It is, however, still not flexible enough. Empirical regional length distributions (see Fig.~\ref{fig:reglen})
 are difficult to capture in that way.
@@
 The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
-any given hidden state duration distribution. Diagram of HSMM is presented in figure~\ref{fig:hsmm}.
+any given hidden state duration distribution. Diagram of HSMM is presented in Fig.~\ref{fig:hsmm}.
 In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
@@
 Almost all entries in transition matrix $A$ are zeros, because regions are sequential.
-Possible transitions are depicted as arrows in figure~\ref{fig:signalHsmm}.
+Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
 Probabilities of observations for hidden states and hidden states durations were estimated from training data.
 The advantage of HSMM model is not only a better performance but also its straightforwardness.
-Figure~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.
+Fig.~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.
 
@@
 conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
-Scheme of extended HSMM can be seen in figure~\ref{fig:ngramext}.
+Scheme of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
 
@@
 
-The both groupings have strongly non-polar and aliphatic group 3 (see Table~\ref{tab:best}, Table~\ref{tab:best} and Figure~\ref{fig:enccomp}) containing isoleucine, leucine, methionine and valine.
+The both groupings have strongly non-polar and aliphatic group 3 (see Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}) containing isoleucine, leucine, methionine and valine.
 Hyrophobic properties of this group in the best encoding are even more pronounced by the presence of tryptophan and phenylalanine.
@@
 
+\subsection*{Benchmark test}
+
+To provide the honest comparison, we trained our model on 2311 signal peptide-containing sequences deposited in Uniprot until 2010 year (an iteration of signalHsmm called signalHsmm-2010).
+The data set should correspond to the set used to train SignalP 4.1.
+Interestingly, our algorithm performed very well even if it was trained on a very small sets including only 336 sequences collected till 1987 year, just after the first method predicting signal peptide was published~\citep{1986vonheijnea}.
+The signalHsmm-1987 was able to very accurately predict a signal peptide, even better than predictors trained on richer data sets.
+It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets.
+
+To check universality of our probabilistic model, we also validated it on signal peptides belonging to specific taxonomic groups.
+As the example, we chose the Plasmodiidae family including malaria parasites.
+The testing data set extracted from UniProt data base contained 111 proteins with a signal peptide and 361 without it.
+We repeated the benchmark scheme for different classifiers – Tab.
+XXX.
+
 \section*{Discussion}
@@
 
-The first set of predictive rules derived by our method lies in the encoding of amino acids.
+The good performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence, but on the physicochemical features of the amino acids.
+Although this finding may seem to offer a minimal improvement when we consider the most popular signal peptides, this generalization pays off when we analyze more specific cases, as signal peptides of \textit{Plasmodiidae}.
+
 Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also found new relationships.
@@
 The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.
+
+The encoding of amino acid plays crucial role in the recognition of signal peptide, but does not affect identification of proteins without a signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
+The mature protein state in the HSMM model tends to have more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
 

### 2015-06-18 paper.tex
added code for benchmark table

@@
 The signalHsmm-1987 was able to very accurately predict a signal peptide, even better than predictors trained on richer data sets.
-It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets.
+It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets~\ref{tab:bench2010}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrr}
+  \toprule
+Software name & AUC & Sensitivity & Specificity & MCC \\ 
+  \midrule
+signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9424 & 0.9720 & 0.9128 & 0.8859 \\ 
+   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9675 & 0.9579 & 0.9771 & 0.9353 \\ 
+  PrediSi \citep{2004hillerpredisi} & 0.8959 & 0.9065 & 0.8853 & 0.7919 \\ 
+   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.9515 & 0.9673 & 0.9358 & 0.9033 \\ 
+  Philius \citep{2008reynoldstransmembrane} & 0.9376 & 0.9533 & 0.9220 & 0.8755 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 & 0.9429 & 0.9393 & 0.8761 & 0.8166 \\ 
+  signalHsmm-1989 & 0.9455 & 0.9486 & 0.8945 & 0.8440 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.} 
+\label{tab:bench2010}
+\end{table}
 

### 2015-06-18 benchmark_signalHsmm.R

@@ -1,9 +1,8 @@
-library(dplyr)
 library(biogram)
 library(signalHsmm)
 library(seqinr)
 library(hmeasure)
-
+library(dplyr)
 other_soft <- read.csv2("benchmark_other_new.csv")
 
 group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
@@ -23,11 +22,31 @@ all_preds <- data.frame(other_soft,
                        signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],
                        signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_data2))[["sp.probability"]])
 
-bench_metrics <- HMeasure(real_labels, all_preds,
-                          threshold = c(rep(0.5, 5), 0.1, 0.1))[["metrics"]]
+calc_mcc <- function(metrics) {
+  TP <- as.numeric(metrics[["TP"]])
+  TN <- metrics[["TN"]]
+  FP <- metrics[["FP"]]
+  FN <- metrics[["FN"]]
+  data.frame(metrics, mcc = unname((TP*TN - FP*FN)/sqrt((TP + FP)*(TP + FN)*(TN + FP)*(TN + FN))))
+}
 
-all_preds[["signalHsmm2010"]] <- all_preds[["signalHsmm2010"]] > 0.05
-all_preds[["signalHsmm1989"]] <- all_preds[["signalHsmm1989"]] > 0.05
+bench_metrics <- calc_mcc(HMeasure(real_labels, all_preds,
+                                   threshold = c(rep(0.5, 5), 0.05, 0.05))[["metrics"]])
 
-FN_preds <- all_preds[1L:218, ]
-rownames(FN_preds[!FN_preds[["signalHsmm2010"]] | !FN_preds[["signalHsmm1989"]], c("signalHsmm2010", "signalHsmm1989")])
+bench_metrics <- data.frame(rownames(bench_metrics), bench_metrics[, c("AUC", "Sens", "Spec", "mcc")])
+colnames(bench_metrics) <- c("Software name", "AUC", "Sensitivity", "Specificity", "MCC")
+
+levels(bench_metrics[["Software name"]]) <- c("Philius \\citep{2008reynoldstransmembrane}", "Phobius \\citep{2004klla}", 
+                                              "PrediSi \\citep{2004hillerpredisi}", 
+                                              "signalHsmm-1989", 
+                                              "signalHsmm-2010", 
[…]

### 2015-06-18 benchmark_signalHsmm.R
added plasmodium data

@@ -3,7 +3,7 @@ library(signalHsmm)
 library(seqinr)
 library(hmeasure)
 library(dplyr)
-other_soft <- read.csv2("benchmark_other_new.csv")
+other_soft <- read.csv2("benchmark_other.csv")
 
 group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
      `2` = c("g", "p", "s", "t", "y"), `3` = c("i", "l", "m", "f", "w", "v"), 
@@ -50,3 +50,19 @@ col <- rep("\\rowcolor[gray]{0.85}", length(rws))
 print(xtable(bench_metrics, caption = "Performance measures for the best encoding. 60 repetitions of cross-validation.", 
              label = "tab:bench2010", digits = 4), include.rownames = FALSE, booktabs = TRUE,
       add.to.row = list(pos = as.list(rws), command = col), sanitize.text.function = identity)
+
+
+# plasmodium ------------------------------------------
+
+other_soft_plas <- read.csv2("benchmark_plas_other.csv")[, -1]
+
+benchmark_plas_data <- read.fasta("benchmark_plas_data.fasta", seqtype = "AA")
+
+real_labels_plas <- c(rep(1, 102), rep(0, 358))
+
+all_preds_plas <- data.frame(other_soft_plas,
+                             signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_plas_data))[["sp.probability"]],
+                             signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_plas_data))[["sp.probability"]])
+
+bench_metrics <- calc_mcc(HMeasure(real_labels_plas, all_preds_plas,
+                                   threshold = c(rep(0.5, 4), 0.05, 0.05))[["metrics"]])

### 2015-06-18 benchmark_signalHsmm.R
added plasmodium data

@@ -3,7 +3,7 @@ library(signalHsmm)
 library(seqinr)
 library(hmeasure)
 library(dplyr)
-other_soft <- read.csv2("benchmark_other_new.csv")
+other_soft <- read.csv2("benchmark_other.csv")
 
 group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
      `2` = c("g", "p", "s", "t", "y"), `3` = c("i", "l", "m", "f", "w", "v"), 
@@ -50,3 +50,15 @@ col <- rep("\\rowcolor[gray]{0.85}", length(rws))
 print(xtable(bench_metrics, caption = "Performance measures for the best encoding. 60 repetitions of cross-validation.", 
              label = "tab:bench2010", digits = 4), include.rownames = FALSE, booktabs = TRUE,
       add.to.row = list(pos = as.list(rws), command = col), sanitize.text.function = identity)
+
+
+# plasmodium ------------------------------------------
+
+other_soft_plas <- read.csv2("benchmark_plasmodium_other.csv")
+
+benchmark_plas_data <- read.fasta("benchmark_plas_data.fasta", seqtype = "AA")
+
+all_preds_plas <- data.frame(other_soft_plas,
+                             signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_plas_data))[["sp.probability"]],
+                             signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_plas_data))[["sp.probability"]])
+

### 2015-06-18 paper.tex
commit before miktex update

@@
 The testing data set extracted from UniProt data base contained 111 proteins with a signal peptide and 361 without it.
-We repeated the benchmark scheme for different classifiers – Tab.
-XXX.
+We repeated the benchmark scheme for different classifiers (Tab.~\ref{tab:bench2010plas}).
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrr}
+  \toprule
+Software name & AUC & Sensitivity & Specificity & MCC \\ 
+  \midrule
+signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
+   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & 0.9469 & 0.4028 \\ 
+  Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
+   \rowcolor[gray]{0.85}Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
+  signalHsmm-2010 & 0.9340 & 0.7941 & 0.8939 & 0.6526 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & 0.7531 \\ 
+   \bottomrule
+\end{tabular}
+\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
+\label{tab:bench2010plas}
+\end{table}
 

### 2015-06-18 benchmark_signalHsmm.R

@@ -47,7 +47,7 @@ library(xtable)
 
 rws <- seq(1, nrow(bench_metrics) - 1, by = 2)
 col <- rep("\\rowcolor[gray]{0.85}", length(rws))
-print(xtable(bench_metrics, caption = "Performance measures for the best encoding. 60 repetitions of cross-validation.", 
+print(xtable(bench_metrics, caption = "Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.", 
              label = "tab:bench2010", digits = 4), include.rownames = FALSE, booktabs = TRUE,
       add.to.row = list(pos = as.list(rws), command = col), sanitize.text.function = identity)
 
@@ -69,3 +69,19 @@ all_preds_plas <- data.frame(other_soft_plas,
 bench_metrics <- calc_mcc(HMeasure(real_labels_plas, all_preds_plas,
                                    threshold = c(rep(0.5, 4), 0.05, 0.05))[["metrics"]])
 
+bench_metrics <- data.frame(rownames(bench_metrics), bench_metrics[, c("AUC", "Sens", "Spec", "mcc")])
+colnames(bench_metrics) <- c("Software name", "AUC", "Sensitivity", "Specificity", "MCC")
+
+levels(bench_metrics[["Software name"]]) <- c("Philius \\citep{2008reynoldstransmembrane}", "Phobius \\citep{2004klla}", 
+                                              "PrediSi \\citep{2004hillerpredisi}", 
+                                              "signalHsmm-1989", 
+                                              "signalHsmm-2010", 
+                                              "signalP 4.1 (no tm) \\citep{2011petersensignalp}", 
+                                              "signalP 4.1 (tm) \\citep{2011petersensignalp}")
+
+
+rws <- seq(1, nrow(bench_metrics) - 1, by = 2)
+col <- rep("\\rowcolor[gray]{0.85}", length(rws))
+print(xtable(bench_metrics, caption = "Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.", 
+             label = "tab:bench2010plas", digits = 4), include.rownames = FALSE, booktabs = TRUE,
+      add.to.row = list(pos = as.list(rws), command = col), sanitize.text.function = identity)

### 2015-06-19 paper.tex
remade tables
more discussion improvments

@@
 \centering
+\caption{Properties used in clusterization.} 
 \begin{tabular}{ll}
@@
 \end{tabular}
-\caption{Properties used in clusterization.} 
 \label{tab:aaprop}
@@
 
-\section*{Results}
+\section*{Results and discussion}
 
@@
 \centering
+\caption{The best sensitivity (final) encoding.} 
 \begin{tabular}{l}
@@
 \end{tabular}
-\caption{The best sensitivity (final) encoding.} 
 \label{tab:best}
@@
 \centering
+\caption{The worst sensitivity encoding.} 
 \begin{tabular}{l}
@@
 \end{tabular}
-\caption{The worst sensitivity encoding.} 
 \label{tab:worst}
@@
 \centering
+\caption{Performance measures for the best encoding. 60 repetitions of cross-validation.} 
 \begin{tabular}{lrr}
@@
 \end{tabular}
-\caption{Performance measures for the best encoding. 60 repetitions of cross-validation.} 
 \label{tab:perfmeas}
@@
 In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals.
-In the best encoding, all charged amino acids, both acids and even weak bases as histidine, also belong to this group.
-This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding.
+In the best encoding, all charged amino acids, both acids and bases (even as weak bases as histidine), also belong to this group.
+This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding, because is nearly absent in the h-region.
 In the worst encoding, where polar and charged character of the Group 1 is not that explicit, the difference in frequency between regions is also less visible.
@@
 
+The encoding of amino acid plays crucial role in the recognition of signal peptide, but does not affect identification of proteins without a signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
+The mature protein state in the HSMM model tends to have more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
+
 \subsection*{Benchmark test}
@@
 \centering
+\caption{Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.} 
 \begin{tabular}{lrrrr}
@@
 \end{tabular}
-\caption{Comparison of Area Under the Curve, Sensitivity, Specificity and Matthews Correlation Coefficient for different classifiers.} 
 \label{tab:bench2010}
@@
 \centering
+\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
 \begin{tabular}{lrrrr}
@@
 \end{tabular}
-\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
 \label{tab:bench2010plas}
@@
 
-\section*{Discussion}
+\subsection*{Conclusions}
 
@@
 Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution.
+Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also discovered new relationships.
+We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins.
+The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.
 
@@
 
-Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also found new relationships.
-We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins.
-The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.
+The flexibility and efficient information recovery makes our model unique among similar software.
+The signalHsmm not only shows very good performance but also is adjustable to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.
 
-The encoding of amino acid plays crucial role in the recognition of signal peptide, but does not affect identification of proteins without a signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
-The mature protein state in the HSMM model tends to have more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
+\subsection*{Availability and implementation}
+The signalHsmm web-server is avaible under address: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
 
-The proposed algorithm recognizing various signal peptides will reduce costs and speed up searching appropriate artificial signal peptides designed for protein secretion in expensive and time-consuming laboratory experiments.
-Moreover, the algorithm can be applied in recognition any sequence types on the nucleotide and amino acid levels, represented by small sets.
-Accessible as a stand-alone and browser-based software, its useful both for researches interested in the analysis of signal peptides.
+signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signal.hsmm}.
 

### 2015-06-19 paper.tex
refined the text, added few words about n-grams

@@
 A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase.
-However, our studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
+However, studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
 Therefore, we designed a new, more univeral probabilistic model for eukaryotic signal peptides, which includes knowledge about their organisation, amino acid composition and variation.
@@
 
-Although signal peptides are quite variable, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig.
+Despite the low sequence homology between signal peptides~\citep{1999ladungaphysean}, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig.
 \ref{fig:sparch}.
@@
 \includegraphics[width=0.55\textwidth]{figures/SP.png}
-\caption{The organization of signal peptide.}
+\caption{The organization of signal peptide. The figure is not drawn to scale.}
 \label{fig:sparch}
@@
 Henceforth, the usage of raw amino acid sequences is superfluous and introduces unnecessary information.
-To utilize this property of signal peptide recognition, we cluster amino acids into several groups based on the physicochemical features of residues.
+To utilize this property of signal peptide recognition, we clustered amino acids into several groups based on the physicochemical features of residues.
 
-The pre-processed sequences are further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions, the enhanced version of algorithm presented in~\cite{1998nielsenprediction}.
-Using the current information from experimentalists, we refined the region recognition criteria.
+The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions, the enhanced version of algorithm presented in~\cite{1998nielsenprediction}.
+We refined some region recognition criteria to attune the algorithm to less typical signal peptides.
 
@@
 The distribution of hidden states durations, the number of amino acids related to each hidden state in signal peptide, is based on the empirical density of region lengths from the training set.
+Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites.
 
@@
 Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06.
-The positive set contained 2~589 sequences with an experimentally confirmed signal peptide and its cleavage site.
+The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, its start and cleavage site.
 Sequences with more than one cleavage site were excluded from the final data set.
@@
       \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
-      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{yellow!50!black!90} k-mer};
+      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{black!90} n-gram};
     \end{tikzpicture} }
@@
 
-HSMM can be further extended to use some specific motives that might occur in proximity of cleavage site.
-Such an extension requires defining additional hidden states. 
-To find such motives we analyze n-grams (k-mers) -- vectors of n characters derived from input sequences. 
-The rigid structure of n-grams may mirror cleavage sites, which are more 
-conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
-Scheme of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
+Hidden semi-Markov models may be flexibly enhanced by adding additional hidden states.
+To improve our model, we added few supplementary states representing specific motives that might occur in proximity of cleavage site.
+The structure of cleavage sites, more conservative  than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.
 
 Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides.
-XXX n-grams were further used in the analysis
+The analyzed sequences were already encodeded using amino acid classification providing the best sensitivity of a general model (see Tab.\ref{tab:best}.
+Selected n-grams, representing less common cleavage site motifs, were included in the HSMM model as the alternative paths at the end of c-region.
+The achitecture of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
 
@@
 To check universality of our probabilistic model, we also validated it on signal peptides belonging to specific taxonomic groups.
-As the example, we chose the Plasmodiidae family including malaria parasites.
-The testing data set extracted from UniProt data base contained 111 proteins with a signal peptide and 361 without it.
+As the example, we chose the Plasmodiidae family including malaria parasites, because of her medical significance.
+The testing data set extracted from UniProt data base contained 102 proteins with a signal peptide and 358 without it.
 We repeated the benchmark scheme for different classifiers (Tab.~\ref{tab:bench2010plas}).
@@
 
-Despite the presence of signal peptide predicting software, there are no exhaustive studies devoted to the analysis of the signal peptide structure and taxonomical variability.
-Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution.
+The architecture of existing signal peptide predicting software is usually opaque, preventing analysis of decision rules included in the model.
+Without ability to trace the computations, researchers cannot discover which parameters are responsible for predictions.
+Considering the biological context of the problem, we see a need for a transparent model of signal peptide.
+signalHsmm is a first step in this direction, providing users with both prediction results and the extandable, readable predictive model.
+
 Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also discovered new relationships.
@@
 \subsection*{Availability and implementation}
-The signalHsmm web-server is avaible under address: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
+The signalHsmm prediction web-server is avaible under address: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
 
 signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signal.hsmm}.
+Stand-alone version offers also tools to build and test own signal peptide models.
 

### 2015-06-19 paper.tex
Fixed some typos too.

@@
 Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
-Moreover, the transport of proteins through the endomembrane system is important for their correct folding and posttranslational modifications.
 
-A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognised by a signal peptidase.
+A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognized by a signal peptidase.
 However, studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
-Therefore, we designed a new, more univeral probabilistic model for eukaryotic signal peptides, which includes knowledge about their organisation, amino acid composition and variation.
+Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation.
 
@@
 
-Our model was validated a signal peptide pedictor.
-It has showed the largest AUC=0.98 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets.
+Our model was validated as a signal peptide predictor.
+It has showed the largest AUC=0.97 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets.
 Thanks to that, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
-It should be emphasised that our model describes signal peptides from medically significant malaria parasites Plasmodium and their relatives (AUC = 0.92) more accurately than popular programs (0.84).
+It should be emphasized that our model describes signal peptides from medically significant malaria parasites Plasmodium and their relatives (AUC = 0.92) more accurately than popular programs (0.84).
+
+The web-server of signalHsmm is avaible under address \url{http://smorfland.uni.wroc.pl/signalhsmm}.
 
@@
 
-Proteins of eukaryotes are encoded in nuclear genomes and are synthetized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum.
+Proteins of eukaryotes are encoded in nuclear genomes and are synthesized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum.
 After the translation process, proteins have to be targeted to specific subcellular compartments or exported outside the cell to the extracellular environment.
@@
 \ref{fig:sparch}) refers to the most ‘typical’ signal peptides.
-There are also exceptionally long signal peptides, which fulfil more sophisticated roles~\citep{2009hissarchitecture}.
+There are also exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}.
 For example, the fragment of signal peptide from preprolactin probably takes part in the regulation of prolactin secretion.
@@
 BLOMAP~\citep{maetschke2005blomap} also employed similar strategy, but considered only encodings based on the substitution matrices.
-Substitution matrices are computed using protein sequences and knowning how atypical signal peptides are, we had chosen different approach.
+Substitution matrices are computed using protein sequences and knowing how atypical signal peptides are, we had chosen different approach.
 
@@
 The cross-validation was repeated 60 times.
-At first, we planned 160 repetitions to ensure with probability 0.95, that every protein without signal peptide was in the learning set, but very small variance of performance measure allowed us to relax this condition.
+At first, we planned 160 repetitions to ensure with probability 0.95, that every protein without signal peptide was in the learning set, but very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) allowed us to relax this condition wihout sacrificing credibility of cross-validation.
 
@@
 Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides.
-The analyzed sequences were already encodeded using amino acid classification providing the best sensitivity of a general model (see Tab.\ref{tab:best}.
+The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of a general model (see Tab.\ref{tab:best}.
 Selected n-grams, representing less common cleavage site motifs, were included in the HSMM model as the alternative paths at the end of c-region.
-The achitecture of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
+The architecture of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
 
@@
 \includegraphics[width=0.95\textwidth]{figures/cvres.png}
-\caption{Results of cross-validation. 1. The encoding providing the best sensitivity (AUC = 0.9683, MCC = 0.8677). 2. The encoding providing the best specificity (AUC = 0.9338, MCC = 0.6474).}
+\caption{Sensitivity and specificity of encodings after cross-validation.
+1.
+The encoding providing the best sensitivity (AUC = 0.9683, MCC = 0.8677).
+2.
+The encoding providing the best specificity (AUC = 0.9338, MCC = 0.6474).}
 \label{fig:cvres}
@@
 \includegraphics[width=0.95\textwidth]{figures/enccomp.png}
-\caption{A comparision of encodings.
+\caption{A comparison of encodings.
 A) Properties of amino acids in the best and the worst sensitivity encoding.
-B) Frequencies of amino acids beloning to the encodings in regions of signal peptide and mature proteins.}
+B) Frequencies of amino acids belonging to the encodings in regions of signal peptide and mature proteins.}
 \label{fig:enccomp}
@@
 All encodings provide very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
-The encodings have the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
+The classification of amino acids has the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
 The final signalHsmm algorithm uses the encoding, that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
@@
 
-\subsection*{Comparision of encodings}
+\subsection*{Comparison of encodings}
 
@@
 
-The amino acids belonging to the Group 2 have quite low probability of occuring in an alpha-helix, but are still very diverse, but all .
+The amino acids belonging to the Group 2 have quite low probability of occurring in an alpha-helix, but are still very diverse, but all .
 Polar but uncharged serine and threonine neighbour with alyphatic glycine and proline.
@@
 
+signalHsmm has much better performance that other algorithms on proteins belonging to organisms from Plasmodiidae family.
+Since our learns more general decision rules, it is able to grasp the similarities between very variable sequences.
+Without losing the ability to predict the more common cases, signalHsmm is able to also deal well with very atypical signal peptides.
+
 \subsection*{Conclusions}
@@
 We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins.
-The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such overrepresentation of group containing alanine in the c-region.
+The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such over-representation of group containing alanine in the c-region.
 
@@
 The flexibility and efficient information recovery makes our model unique among similar software.
-The signalHsmm not only shows very good performance but also is adjustable to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.
+The signalHsmm not only shows very good performance but is also adjustable enough to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.
+Morrover, our method can effectively extract information from very small data sets, which in future may lead to new predictors, able to recognize less known signaling sequences.
 
@@
 
-signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signal.hsmm}.
-Stand-alone version offers also tools to build and test own signal peptide models.
+signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signalHsmm}.
+Stand-alone version, in addition to the prediction of signal peptides, offers tools to build, train and test novel signal peptide models.
 

### 2015-06-19 signalHSMM_kMer_report.Rmd
working signalHSMM with k-mers. some auxilliary files and report with final results

---
title: "signalHsmm with k-mer hidden states"
author: "Piotr Sobczyk"
date: "19 czerwca 2015"
output: html_document
---

We shall compare performance of currently used software with signalHSMM trained on data obtained by 1989, by 2010,
with and without k-mer hidden state.

```{r, message=FALSE}
library(dplyr)
library(biogram)
library(signalHsmm)
library(seqinr)
library(hmeasure)
library(ggplot2)
library(knitr)
library(foreach)
library(doMC)
```

First let us load data
```{r, cache=TRUE}
other_soft <- read.csv2("benchmark_other_new.csv")

group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
     `2` = c("g", "p", "s", "t", "y"), `3` = c("i", "l", "m", "f", "w", "v"), 
     `4` = c("a", "c"))

signalHsmm2010 <- train_hsmm(read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal"), 
                             aa_group = group_best)
signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = "signal"), 
                             aa_group = group_best)
signalHSMM_proteins <- read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal")
benchmark_full <- read_uniprot("sp2010_2015.txt", euk = TRUE, what = "signal")
benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:218)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])

# usuniecie 4 obserwacji negatywnych
negs <- c(432, 215, 280, 417)
# negs <- sample(215:432, 4)
# [1] 432 215 280 414
benchmark_data2 <- benchmark_data2[-negs]
other_soft <- other_soft[-negs,]
real_labels <- c(rep(1, 214), rep(0, 214))
```

Now let us compare predictions on that data
```{r, warning=FALSE, cache=TRUE}
all_preds <- data.frame(other_soft,
                       signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],
                       signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_data2))[["sp.probability"]])

bench_metrics <- HMeasure(real_labels, all_preds,
                          threshold = c(rep(0.5, 5), 0.1, 0.1))[["metrics"]]
kable(bench_metrics)
```

Ok, let's move to defining k-mer states.

Show me what you got
```{r}
b <- lapply(signalHSMM_proteins, function(x){
  cut <- attr(x, "sig")[2]
  deg_sample <- as.numeric(degenerate(toupper(x)[1L:min(70,length(x))], group_best))
  #remove atypical amino acids
  deg_sample <- na.omit(deg_sample)
  deg_sample[(cut-2):(cut+2)]
})

b <- matrix(unlist(b), ncol=5, byrow = TRUE)
ggplot(data.frame(table(apply(b, 1, function(x) paste0(x[2:4], collapse="")))),
       aes(x=Var1, y=Freq)) + geom_point()
ggplot(data.frame(table(apply(b, 1, function(x) paste0(x[1:3], collapse="")))),
       aes(x=Var1, y=Freq)) + geom_point()
```

From that we will extract some most promising k-mers

```{r}
p4 <- 0.1
p3 <- 0.25/(1-p4)
p2 <- 0.35/(1-p4)/(1-p3)
p1 <- 0.1/(1-p4)/(1-p3)/(1-p2)

sum(b[,1]%in%c(3) & b[,2]%in%c(1) & b[,3] %in% c(2) & b[,4]%in%c(1) & b[,4]%in%c(1,3))/nrow(b)
kMer1 <- list(3, 1, 2, 1, c(1,3))
pState1 <- 3
nState1 <- 4
pTrans1 <- p1

# -3|222
sum(b[,1]%in%c(2,3) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
kMer2 <- list(c(2,3),c(1,2), c(2,4))
pState2 <- 3
nState2 <- 4
pTrans2 <- p2

#-3|4(1,2,3,4)(2,4)
sum(b[,1]%in%c(4) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
kMer3 <- list(4, c(1,2,3), c(2,4))
pState3 <- 3
nState3 <- 4
pTrans3 <- p3

# kMer4
sum(b[,1]%in%c(2) & b[,2]%in%c(1) & b[,3] %in% c(1,4) & b[,4]%in%c(1,2,3,4))/nrow(b)
kMer4 <- list(2, 1, c(1,4))
pState4 <- 3
nState4 <- 4
pTrans4 <- p4

lastState1 = 4+1+length(kMer1)
lastState2 = lastState1+1+length(kMer2)
lastState3 = lastState2+1+length(kMer3)
lastState4 = lastState3+1+length(kMer4)
```

Now let us build a new model on **signalHsmm2010**
```{r}
  parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
                                     signalHsmm2010$params, pState1, nState1, pTrans1, d=3)
  parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
                                     od = parametersSet2$od, params = parametersSet2$params,
                                     pState2, nState2, pTrans2, d=3)
  parametersSet4 <- add_k_mer_state2(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
                                     od = parametersSet3$od, params = parametersSet3$params, 
                                     pState3, nState3, pTrans3, d=3)
  parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
                                     od = parametersSet4$od, params = parametersSet4$params, 
                                     pState4, nState4, pTrans4, d=3)
  overall_probs_log <- signalHsmm2010$overall_probs_log
  maxSignal <- 45
```

```{r}
registerDoMC(cores=7)
n <- length(benchmark_data2)
results <- foreach(i=1:n) %dopar% {
    prot <- benchmark_data2[[i]]
    deg_sample <- na.omit(as.numeric(degenerate(toupper(prot)[1L:min(maxSignal, length(prot))], group_best)))

    viterbi_res <- duration_viterbi2(deg_sample-1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, 
                                     signalHsmm2010$od, signalHsmm2010$params)
    viterbi_path <- viterbi_res[["path"]]+1
    c_site <- ifelse(any(viterbi_path == 4), 
                     max(which(viterbi_path == 3)), 
                     length(deg_sample))
    prob.signal <- viterbi_res[["viterbi"]][c_site, viterbi_path[c_site]]
    prob.non <- Reduce(function(x, y) x + overall_probs_log[y], deg_sample[1L:c_site], 0)
    prob.total <- exp(prob.signal - prob.non)

    viterbi_res5 <- duration_viterbi2(deg_sample-1, parametersSet5$pipar, parametersSet5$tpmpar, 
                                      parametersSet5$od, parametersSet5$params)
    viterbi_path5 <- viterbi_res5[["path"]]+1
    c_site5 <- ifelse(any(viterbi_path5 == 4), 
                      min(which(viterbi_path5 == 4))-1, 
                      length(deg_sample))
    c_site5 <- ifelse(any(viterbi_path5 == lastState1), 
                      min(which(viterbi_path5 == lastState1)), 
                      c_site5)
    c_site5 <- ifelse(any(viterbi_path5 == lastState2), 
                      min(which(viterbi_path5 == lastState2)), 
                      c_site5)
    c_site5 <- ifelse(any(viterbi_path5 == lastState3), 
                      min(which(viterbi_path5 == lastState3)), 
                      c_site5)
    c_site5 <- ifelse(any(viterbi_path5 == lastState4), 
                      min(which(viterbi_path5 == lastState4)), 
                      c_site5)
    #get probabilities of signal peptide model
    prob.signal5 <- viterbi_res5[["viterbi"]][c_site5, viterbi_path5[c_site5]]
    #get probabilities of no signal peptide model
    prob.non5 <- Reduce(function(x, y) x + overall_probs_log[y], deg_sample[1L:c_site5], 0)
    prob.total5 <- exp(prob.signal5 - prob.non5)

    c(1 - 1/(1 + prob.total), 1 - 1/(1 + prob.total5), c_site, c_site5)
}
probs <- t(sapply(results, function (x) head(x, 2)))
cut <- t(sapply(results, function (x) tail(x, 2)))
```

```{r, echo=FALSE}
ggplot(data.frame(noMer=probs[,1],kMer=probs[,2], sig = factor(real_labels)), aes(x=noMer,y=kMer, color=sig)) +
  geom_point() + xlab("No k-mer") + ylab("with k-mer") + 
  scale_color_discrete("Signal peptide") +geom_abline(slope=1)
```

```{r, echo=FALSE}
all_preds2 = all_preds
all_preds2$signalKmer = probs[,2]

bench_metrics2 <- HMeasure(real_labels, all_preds2,
                             threshold = c(rep(0.5, 5), 0.1, 0.1, 0.5))[["metrics"]]
kable(bench_metrics2)
```

```{r}
#zle <- which(real_labels==1 & probs[,2]<0.75)
zle <- which(real_labels==1 & probs[,2]<probs[,1] & probs[,2]<0.99)
b2 <- lapply(benchmark_full[zle], function(x){
  cut <- attr(x, "sig")[2]
  deg_sample <- as.numeric(degenerate(toupper(x)[1L:min(50, length(x))], group_best))
  deg_sample <- na.omit(deg_sample)
  deg_sample[(cut-2):(cut+2)]
})

b2 <- matrix(unlist(b2), ncol=5, byrow = TRUE)
kable(data.frame(probs[zle,], seq=apply(b2, 1, function(x) paste0(x, collapse="")), 
      csite=sapply(benchmark_full[zle], function(x) cut <- attr(x, "sig")[2])))

kable(data.frame(which(real_labels==0 & probs[,2]>0.5),
      probs[which(real_labels==0 & probs[,2]>0.5),1],
      probs[which(real_labels==0 & probs[,2]>0.5),2]))
```

### 2015-06-19 benchmark_signalHsmm.R

@@ -14,10 +14,17 @@ signalHsmm2010 <- train_hsmm(read_uniprot("sp1950_2010.txt", euk = TRUE, what = 
                              aa_group = group_best)
 signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = "signal"), 
                              aa_group = group_best)
+benchmark_full <- read_uniprot("sp2010_2015.txt", euk = TRUE, what = "signal")
 benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
-benchmark_data2 <- lapply(benchmark_data, function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
+benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:218)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
 
-real_labels <- c(rep(1, 218), rep(0, 218))
+# usuniecie 4 obserwacji negatywnych
+negs <- c(432, 215, 280, 417)
+# negs <- sample(215:432, 4)
+# [1] 432 215 280 414
+benchmark_data2 <- benchmark_data2[-negs]
+other_soft <- other_soft[-negs,]
+real_labels <- c(rep(1, 214), rep(0, 214))
 
 all_preds <- data.frame(other_soft,
                        signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],
@@ -25,9 +32,9 @@ all_preds <- data.frame(other_soft,
 
 bench_metrics <- HMeasure(real_labels, all_preds,
                           threshold = c(rep(0.5, 5), 0.1, 0.1))[["metrics"]]
-
-all_preds[["signalHsmm2010"]] <- all_preds[["signalHsmm2010"]] > 0.05
-all_preds[["signalHsmm1989"]] <- all_preds[["signalHsmm1989"]] > 0.05
-
-FN_preds <- all_preds[1L:218, ]
-rownames(FN_preds[!FN_preds[["signalHsmm2010"]] | !FN_preds[["signalHsmm1989"]], c("signalHsmm2010", "signalHsmm1989")])
+bench_metrics
+# all_preds[["signalHsmm2010"]] <- all_preds[["signalHsmm2010"]] > 0.05
+# all_preds[["signalHsmm1989"]] <- all_preds[["signalHsmm1989"]] > 0.05
+# 
[…]

### 2015-06-19 benchmark_with_k_mer_states.R

@@ -26,7 +26,7 @@ sum(b[,1]%in%c(3) & b[,2] %in% c(1,2,3) & b[,3]%in%c(2))/length(signalHSMM_prote
 sum(b[,2]%in%c(1,2) & b[,3] %in% c(2,4) & b[,4]==3)/length(signalHSMM_proteins)
 sum(b[,1]%in%c(2,3,4) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4) & b[,4]%in%c(1))/length(signalHSMM_proteins)
 
-sum(b[,1]%in%c(2) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2))/length(signalHSMM_proteins)
+sum(b[,1]%in%c(2) & b[,2]%in%c(1,2) & b[,3] %in% c(2,3))/length(signalHSMM_proteins)
 
 
 sum(b[,4]==3, na.rm=T)/length(signalHSMM_proteins)
@@ -153,7 +153,7 @@ auc = as.numeric(performance(ROCRpredTest, "auc")@y.values)
 auc 
   
 zle <- which(real_labels==1 & a[,1]>a[,3])
-b2 <- lapply(signalHSMM_proteins[zle], function(x){
+b2 <- lapply(benchmark_full[zle], function(x){
   cut <- attr(x, "sig")[2]
   deg_sample <- as.numeric(degenerate(toupper(x)[1L:50], group_best))
   deg_sample <- na.omit(deg_sample)

### 2015-06-19 experiments_k_mers.R

@@ -1,0 +1,148 @@
+
+  p4 <- 0
+  p3 <- 0.34/(1-p4)
+  p2 <- 0.26/(1-p4)/(1-p3)
+  p1 <- 0/(1-p4)/(1-p3)/(1-p2)
+  
+  # -3|(2,3,4)(1,2,3)(2,4)(1,3)
+  sum(b[,1]%in%c(1,3) & b[,2]%in%c(1,2,3) & b[,3] %in% c(1,2,3,4) & b[,4]%in%c(1,3,4))/nrow(b)
+  kMer1 <- list(c(1,3), c(1,2,3), c(1,2,3,4), c(1,3,4))
+  pState1 <- 3
+  nState1 <- 4
+  pTrans1 <- p1
+  
+  # -3|222
+  sum(b[,1]%in%c(2,3) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
+  kMer2 <- list(c(2,3),c(1,2), c(2,4))
+  pState2 <- 3
+  nState2 <- 4
+  pTrans2 <- p2
+  
+  #-3|4(1,2,3,4)(2,4)
+  sum(b[,1]%in%c(4) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
+  kMer3 <- list(4, c(1,2,3), c(2,4))
+  pState3 <- 3
+  nState3 <- 4
+  pTrans3 <- p3
+  
+#   # -2|1_44
+#   sum(b[,1]%in%c(2) & b[,2]%in%c(2,3) & b[,3] %in% c(2,4))/length(signalHSMM_proteins)
+#   kMer4 <- list(2, c(2,3), c(2,4))
+#   pState4 <- 3
+#   nState4 <- 4
+#   pTrans4 <- p4
+  
+  
+  parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
+                                     signalHsmm2010$params, pState1, nState1, pTrans1, d=4)
+  parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
+                                     od = parametersSet2$od, params = parametersSet2$params,
+                                     pState2, nState2, pTrans2, d=3)
+  # parametersSet5 <- parametersSet3
[…]

### 2015-06-19 finding_k_mers.R

@@ -1,0 +1,47 @@
+#' 
+#' tools for heuristic recognition of important k-mers
+#' 
+
+library(ggplot2)
+
+signalHSMM_proteins <- read_uniprot("sp1950_2010.txt", euk = TRUE, what = "signal")
+b <- lapply(benchmark_full, function(x){
+  cut <- attr(x, "sig")[2]
+  deg_sample <- as.numeric(degenerate(toupper(x)[1L:min(70,length(x))], group_best))
+  #remove atypical amino acids
+  deg_sample <- na.omit(deg_sample)
+  deg_sample[(cut-2):(cut+2)]
+})
+
+b <- matrix(unlist(b), ncol=5, byrow = TRUE)
+table(apply(b, 1, function(x) paste0(x, collapse="")))
+table(apply(b, 1, function(x) paste0(x[1:4], collapse="")))
+table(apply(b, 1, function(x) paste0(x[4:1], collapse="")))
+ggplot(data.frame(table(apply(b, 1, function(x) paste0(x[4:2], collapse="")))),
+       aes(x=Var1, y=Freq)) + geom_point()
+
+# 2 (1,2,3,4) (2,3,4)
+# (2,3,4) (4) (2,4)
+# (2,3,4) (1,2,3) (2,4)
+# _ 4 (2,4) (2,3,4) 
+table(apply(b, 1, function(x) paste0(x[3:1], collapse="")))
+ggplot(data.frame(table(apply(b, 1, function(x) paste0(x[1:3], collapse="")))),
+       aes(x=Var1, y=Freq)) + geom_point()
+
+sum(b[,1]%in%c(2,3,4) & b[,2] %in% c(4) & b[,3]%in%c(2,4))/length(signalHSMM_proteins)
+sum(b[,1]%in%c(2,3,4) & b[,2] %in% c(1,2,3) & b[,3]%in%c(2,4))/length(signalHSMM_proteins)
+sum(b[,2]%in%c(4) & b[,3] %in% c(2,4) & b[,4]%in%c(2,3,4))/length(signalHSMM_proteins)
+
+
+benchmark_full <- read_uniprot("sp2010_2015.txt", euk = TRUE, what = "signal")
+b2 <- lapply(benchmark_full, function(x){
+  cut <- attr(x, "sig")[2]
+  deg_sample <- as.numeric(degenerate(toupper(x)[1L:50], group_best))
[…]

### 2015-06-19 good_models.R

@@ -1,0 +1,68 @@
+p4 <- 0
+p3 <- 0.2/(1-p4)
+p2 <- 0.4/(1-p4)/(1-p3)
+p1 <- 0/(1-p4)/(1-p3)/(1-p2)
+
+# -3|(2,3,4)(1,2,3)(2,4)(1,3)
+sum(b[,1]%in%c(1) & b[,2]%in%c(1,3) & b[,3] %in% c(1,3,4) & b[,4]%in%c(1,3,4))/nrow(b)
+kMer1 <- list(1, c(1,3), c(1,3,4), c(1,3,4))
+pState1 <- 3
+nState1 <- 4
+pTrans1 <- p1
+
+# -3|222
+sum(b[,1]%in%c(2,3) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
+kMer2 <- list(c(2,3),c(1,2), c(2,4))
+pState2 <- 3
+nState2 <- 4
+pTrans2 <- p2
+
+#-3|4(1,2,3,4)(2,4)
+sum(b[,1]%in%c(4) & b[,2]%in%c(1,2,3) & b[,3] %in% c(2,4))/nrow(b)
+kMer3 <- list(4, c(1,2,3), c(2,4))
+pState3 <- 3
+nState3 <- 4
+pTrans3 <- p3
+
+
+
+p4 <- 0
+p3 <- 0.25/(1-p4)
+p2 <- 0.35/(1-p4)/(1-p3)
+p1 <- 0/(1-p4)/(1-p3)/(1-p2)
+
+
+##
+
+p4 <- 0.2
+p3 <- 0.1/(1-p4)
+p2 <- 0.05/(1-p4)/(1-p3)
+p1 <- 0.55/(1-p4)/(1-p3)/(1-p2)
+
+# kMer1
+sum(b[,1]%in%c(2,3,4) & b[,2]%in%c(1) & b[,3] %in% c(2,4) & b[,4]%in%c(1,2,3,4))/nrow(b)
+kMer1 <- list(c(2,3,4), c(1,2,3), c(2,4))
+pState1 <- 3
+nState1 <- 4
+pTrans1 <- p1
+
+# kMer2
+sum(b[,1]%in%c(3) & b[,2]%in%c(1) & b[,3] %in% c(2) & b[,4]%in%c(1) & b[,4]%in%c(1,3))/nrow(b)
+kMer2 <- list(3, 1, 2, 1, c(1,3))
[…]

### 2015-06-19 signalHSMM_kMer_report.Rmd
updating benchmark to include model with k-mers

@@
 negs <- c(432, 215, 280, 417)
-# negs <- sample(215:432, 4)
-# [1] 432 215 280 414
 benchmark_data2 <- benchmark_data2[-negs]

### 2015-06-19 benchmark_signalHsmm.R

@@ -3,6 +3,8 @@ library(signalHsmm)
 library(seqinr)
 library(hmeasure)
 library(dplyr)
+library(foreach)
+library(doMC)
 other_soft <- read.csv2("benchmark_other.csv")
 
 group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
@@ -15,12 +17,10 @@ signalHsmm1989 <- train_hsmm(read_uniprot("sp1950_1989.txt", euk = TRUE, what = 
                              aa_group = group_best)
 benchmark_full <- read_uniprot("sp2010_2015.txt", euk = TRUE, what = "signal")
 benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
-benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:218)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
+benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:214)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
 
 # usuniecie 4 obserwacji negatywnych
 negs <- c(432, 215, 280, 417)
-# negs <- sample(215:432, 4)
-# [1] 432 215 280 414
 benchmark_data2 <- benchmark_data2[-negs]
 other_soft <- other_soft[-negs,]
 real_labels <- c(rep(1, 214), rep(0, 214))
@@ -28,6 +28,91 @@ real_labels <- c(rep(1, 214), rep(0, 214))
 all_preds <- data.frame(other_soft,
                        signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_data2))[["sp.probability"]],
                        signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_data2))[["sp.probability"]])
+
+p4 <- 0.1
+p3 <- 0.25/(1-p4)
+p2 <- 0.35/(1-p4)/(1-p3)
+p1 <- 0.1/(1-p4)/(1-p3)/(1-p2)
+
+sum(b[,1]%in%c(3) & b[,2]%in%c(1) & b[,3] %in% c(2) & b[,4]%in%c(1) & b[,4]%in%c(1,3))/nrow(b)
+kMer1 <- list(3, 1, 2, 1, c(1,3))
[…]

### 2015-06-19 paper.tex
ready for dr Mackiewicz

@@
   \midrule
-signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9424 & 0.9720 & 0.9128 & 0.8859 \\ 
-   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9675 & 0.9579 & 0.9771 & 0.9353 \\ 
-  PrediSi \citep{2004hillerpredisi} & 0.8959 & 0.9065 & 0.8853 & 0.7919 \\ 
-   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.9515 & 0.9673 & 0.9358 & 0.9033 \\ 
-  Philius \citep{2008reynoldstransmembrane} & 0.9376 & 0.9533 & 0.9220 & 0.8755 \\ 
-   \rowcolor[gray]{0.85}signalHsmm-2010 & 0.9429 & 0.9393 & 0.8761 & 0.8166 \\ 
-  signalHsmm-1989 & 0.9455 & 0.9486 & 0.8945 & 0.8440 \\ 
+signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9416 & 0.9720 & 0.9112 & 0.8848 \\ 
+   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9673 & 0.9579 & 0.9766 & 0.9347 \\ 
+  PrediSi \citep{2004hillerpredisi} & 0.8949 & 0.9065 & 0.8832 & 0.7899 \\ 
+   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.9509 & 0.9673 & 0.9346 & 0.9024 \\ 
+  Philius \citep{2008reynoldstransmembrane} & 0.9369 & 0.9533 & 0.9206 & 0.8743 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 & 0.9526 & 0.9533 & 0.8832 & 0.8385 \\ 
+  signalHsmm-1989 & 0.9562 & 0.9626 & 0.8972 & 0.8617 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 with k-mers & 0.9679 & 0.9673 & 0.9112 & 0.8799 \\ 
    \bottomrule
@@
    \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & 0.7531 \\ 
+  signalHsmm-2010 with k-mers & 0.9334 & 0.9902 & 0.7989 & 0.6767 \\ 
    \bottomrule

### 2015-06-19 benchmark_signalHsmm.R

@@ -80,7 +80,7 @@ parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar =
 overall_probs_log <- signalHsmm2010$overall_probs_log
 maxSignal <- 45
 
-registerDoMC(cores=6)
+registerDoMC(cores=4)
 n <- length(benchmark_data2)
 results <- foreach(i=1:n) %dopar% {
   prot <- benchmark_data2[[i]]
@@ -158,9 +158,42 @@ all_preds_plas <- data.frame(other_soft_plas,
                              signalHsmm2010 = pred2df(predict(signalHsmm2010, benchmark_plas_data))[["sp.probability"]],
                              signalHsmm1989 = pred2df(predict(signalHsmm1989, benchmark_plas_data))[["sp.probability"]])
 
+registerDoMC(cores=4)
+n <- length(benchmark_plas_data)
+results <- foreach(i=1:n) %dopar% {
+  prot <- benchmark_plas_data[[i]]
+  deg_sample <- na.omit(as.numeric(degenerate(toupper(prot)[1L:min(maxSignal, length(prot))], group_best)))
+  
+  viterbi_res <- duration_viterbi2(deg_sample-1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, 
+                                   signalHsmm2010$od, signalHsmm2010$params)
+  viterbi_path <- viterbi_res[["path"]]+1
+  c_site <- ifelse(any(viterbi_path == 4), max(which(viterbi_path == 3)), length(deg_sample))
+  prob.signal <- viterbi_res[["viterbi"]][c_site, viterbi_path[c_site]]
+  prob.non <- Reduce(function(x, y) x + overall_probs_log[y], deg_sample[1L:c_site], 0)
+  prob.total <- exp(prob.signal - prob.non)
+  
+  viterbi_res5 <- duration_viterbi2(deg_sample-1, parametersSet5$pipar, parametersSet5$tpmpar, 
+                                    parametersSet5$od, parametersSet5$params)
+  viterbi_path5 <- viterbi_res5[["path"]]+1
+  c_site5 <- ifelse(any(viterbi_path5 == 4), min(which(viterbi_path5 == 4))-1, length(deg_sample))
+  c_site5 <- ifelse(any(viterbi_path5 == lastState1), min(which(viterbi_path5 == lastState1)), c_site5)
+  c_site5 <- ifelse(any(viterbi_path5 == lastState2), min(which(viterbi_path5 == lastState2)), c_site5)
[…]

### 2015-06-19 paper.tex
improved plots

@@
 \usepackage{colortbl, xcolor}
+\usepackage{epstopdf}
 
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/reglen.png}
+\includegraphics[width=0.95\textwidth]{figures/reglen.eps}
 \caption{A) Distribution of lengths of signal peptides.
@@
     \label{fig:model_hsmm}
-      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
-      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
-         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
+      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=3mm, -latex']
+      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\Large\itshape,
+         align=center,  ] %font=\sffamily\small\bfseries,
       %nodes
       \node[main node]            (start') 	[]						{Start};	     
-      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
-      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
-      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
-      \node[main node, bottom color=green!70!black!70] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};
+      \node[main node, fill=purple!80!black!40 ] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
+      \node[main node, fill=orange!70!black!40] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
+      \node[main node, fill=yellow!70!black!40] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
+      \node[main node, fill=green!70!black!40] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};
 
@@
     \label{fig:model_hsmm_kMer}
-      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
-      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
-         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
+      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=3mm, -latex']
+      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\Large\itshape,
+         align=center,  ]
       %nodes
       \node[main node]            (start')   []  					{Start};	     
-      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-25mm, yshift=35mm] 	{n-region};
-      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-25mm,yshift=35mm] 	{h-region};
-      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-10mm,yshift=-25mm] 	{c-region};
-      \node[main node, bottom color=green!70!black!70] 	(mature') 	[below of=hregion',xshift=15mm, yshift=-90mm] 	{Mature protein};
+      \node[main node, fill=purple!80!black!40] 	(nregion') 	[right of=start',xshift=-25mm, yshift=35mm] 	{n-region};
+      \node[main node, fill=orange!70!black!40] 	(hregion') 	[right of=nregion',xshift=-15mm,yshift=35mm] 	{h-region};
+      \node[main node, fill=yellow!70!black!40] 	(cregion') 	[right of=hregion',xshift=-10mm,yshift=-10mm] 	{c-region};
+      \node[main node, fill=green!70!black!40] 	(mature') 	[below of=hregion',xshift=15mm, yshift=-110mm] 	{Mature protein};
 
-      \node[main node, bottom color=yellow!70!black!70]   (kmer1) 	[below of=cregion',xshift=35mm,yshift=30mm] 	{Aminoacid 1};
-      \node[main node, bottom color=yellow!70!black!70]   (kmer2) 	[below of=kmer1,xshift=-5mm,yshift=5mm] 	{Aminoacid 2};
-      \node[main node, bottom color=yellow!70!black!70]   (kmer3) 	[below of=kmer2,xshift=-5mm,yshift=5mm] 	{Aminoacid 3};
+      \node[main node, fill=green!50!red!60!black!20]   (kmer1) 	[below of=cregion',xshift=35mm,yshift=30mm] 	{Aminoacid 1};
+      \node[main node, fill=green!50!red!60!black!20]   (kmer2) 	[below of=kmer1,xshift=-5mm,yshift=-5mm] 	{Aminoacid 2};
+      \node[main node, fill=green!50!red!60!black!20]   (kmer3) 	[below of=kmer2,xshift=-5mm,yshift=-5mm] 	{Aminoacid 3};
       %lines
@@
       \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
-      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{black!90} n-gram};
+      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\LARGE \color{black!90} n-gram};
     \end{tikzpicture} }
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/cvres.png}
+\includegraphics[width=0.95\textwidth]{figures/cvres.eps}
 \caption{Sensitivity and specificity of encodings after cross-validation.
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/enccomp.png}
+\includegraphics[width=0.95\textwidth]{figures/enccomp.eps}
 \caption{A comparison of encodings.

### 2015-06-19 aminoacid_encoding.R

@@ -29,7 +29,7 @@ colnames(aa_nvals) <- tolower(a(colnames(aa_nvals)))
 
 aa_props <- sapply(aaindex, function(i) i[["D"]])
 
-write.csv2(data.frame(property = unname(aa_props)), file = "aa_props.csv")
+#write.csv2(data.frame(property = unname(aa_props)), file = "aa_props.csv")
 
 
 traits <- list(size = c(63, 72, 109, 399),

### 2015-06-19 encodings_vs_regions.R

@@ -107,7 +107,7 @@ dat_bestworst <- cbind(enc = unlist(lapply(c("best", "worst"), function(i) rep(i
 #   my_theme
 
 p1 <- ggplot(dat_bestworst, aes(x = critertion, y = value, col = enc, fill = enc)) +
-  geom_point(size = 5, shape = 21, position = position_dodge(width=0.5)) +
+  geom_point(size = 2.5, shape = 21, position = position_dodge(width=0.5)) +
   #geom_text(hjust = -1) +
   facet_grid(~group) +
   scale_x_discrete("Criterion\n", labels = c("size" = "Size","hydroph" = "Hydroph.",
@@ -117,7 +117,7 @@ p1 <- ggplot(dat_bestworst, aes(x = critertion, y = value, col = enc, fill = enc
   scale_fill_manual("Encoding: ", values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 
   my_theme
 
-png("./figures/enccomp.png", width = 2257*0.5, height = 1201*0.85)
+cairo_ps("./figures/enccomp.eps", width = 9, height = 8, onefile = FALSE)
 print(arrangeGrob(textGrob("A", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p1, 
                   textGrob("B", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p3,
                   nrow = 2, ncol = 2, widths = c(0.05, 0.95)))

### 2015-06-19 plot_tools.R

@@ -2,7 +2,7 @@ library(ggplot2)
 library(grid)
 library(gridExtra)
 
-size_mod <- 2
+size_mod <- -5
 
 my_theme <- theme(plot.background=element_rect(fill = "transparent",
                                                colour = "transparent"),

### 2015-06-19 results_analysis.R

@@ -134,7 +134,7 @@ source("plot_tools.R")
 
 p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding, colour = encoding == "", fill = encoding == "")) +
   geom_point(size = 5, shape = 21) +
-  geom_text(size = 9, hjust = -0.5, vjust = 0) +
+  geom_text(size = 6, hjust = -0.75, vjust = 1) +
   scale_colour_manual(values = c("red","blue")) + 
   scale_fill_manual(values = c(adjustcolor("red", 0.25), adjustcolor("blue", 0.25))) + 
   scale_x_continuous("Sensitivity\n") +
@@ -143,8 +143,7 @@ p1 <- ggplot(p1_dat, aes(x = Sens, y = Spec, label = encoding, colour = encoding
   guides(colour = FALSE, fill = FALSE)
   
 
-
-png("./figures/cvres.png", width = 2257*0.5, height = 1201*0.5)
+cairo_ps("./figures/cvres.eps", width = 9, height = 5, onefile = FALSE)
 print(p1)
 dev.off()
 

### 2015-06-19 sp_length.R

@@ -47,7 +47,7 @@ p1 <- ggplot(mlen, aes(x = value)) +
   my_theme
 
 
-png("./figures/reglen.png", width = 2257*0.5, height = 1201*0.5)
+cairo_ps("./figures/reglen.eps", width = 9, height = 5, onefile = FALSE)
 print(arrangeGrob(textGrob("A", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p2, 
                   textGrob("B", x = 0.75, y = 0.9, gp=gpar(fontsize=22)), p1, 
                   nrow = 2, ncol = 2, widths = c(0.05, 0.95)))

### 2015-06-19 paper.tex
fixing hsmm diagram

@@
 
-(1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
+\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
+(1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
 

### 2015-06-20 paper.tex
improvements from PM + resized tables and figures to fit on 10 pages

--- 2015-06-19 paper.tex
@@
 
-\title{signalHsmm - a novel semi-Markov model of eukaryotic signal peptides}
+\title{SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides}
 
@@
 
-\keywords{signal petide prediction; n-gram; hidden semi-Markov models}
+\keywords{eukaryote; hidden semi-Markov models; n-gram; signal peptide prediction}
 
 \begin{abstract}
-The proper localization of proteins in a cell is essential to maintain their desired function.
-Information about the protein destination is included within the very protein in the form of short peptides called targeting signals.
-Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell.
+Short stretches of amino acid residues called targeting signals are involved in the proper localization of proteins in a cell, which is essential to maintain their desired function.
+Ones of them are signal peptides, which are responsible for targeting of proteins to endomembrane system and their export outside the cell.
 Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
-
-A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognized by a signal peptidase.
-However, studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides.
+Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction.
 Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation.
-
-The proposed model is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
+The proposed approach is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
 The big advantage of the algorithm is its extensibility.
-Using the n-grams (k-mers) we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.
-
-Our model was validated as a signal peptide predictor.
-It has showed the largest AUC=0.97 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets.
-Thanks to that, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
-It should be emphasized that our model describes signal peptides from medically significant malaria parasites Plasmodium and their relatives (AUC = 0.92) more accurately than popular programs (0.84).
-
-The web-server of signalHsmm is avaible under address \url{http://smorfland.uni.wroc.pl/signalhsmm}.
+Using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
+Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even after training on very small data sets.
+Moreover, it appeared very stable regardless of types of learning data sets.
+Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
+It should be emphasized that our model described signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
+The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.
 
@@
 
-\subsection*{Signal peptides}
+\subsection*{Roles and features of signal peptides}
 
 Proteins of eukaryotes are encoded in nuclear genomes and are synthesized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum.
-After the translation process, proteins have to be targeted to specific subcellular compartments or exported outside the cell to the extracellular environment.
+After translation, proteins are targeted to specific subcellular compartments or exported outside the cell.
 The proper localization of proteins is essential to perform their desired function.
-Information about protein destination is included within the very protein in the form of short peptides or stretches of amino acid residues called targeting or sorting signals.
+Information about the protein destination is included within the very protein in short stretches of amino acid residues called targeting or sorting signals.
 Ones of them are signal peptides, which are located at the N-terminus of proteins.
 
-Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes.
-Such proteins can stay inside of these compartments, or can be inserted into cellular membranes or exported outside the cell.
-Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome.
-They play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}.
+Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum and Golgi apparatus.
+Such proteins can stay inside of these compartments, can be inserted into cellular membranes or exported outside the cell.
+Proteins equipped with signal peptides play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}.
 Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.
@@
 It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues.
-They probably enforce a proper topology on the polypeptide during translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}.
+They probably enforce a proper topology on the polypeptide during its translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}.
 The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues.
-It constitutes a core region of signal peptide and has a tendency to form $\alpha$-helix.
-The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site recognized by the signal peptidase.
+It constitutes a core region of signal peptide and usually forms $\alpha$-helix.
+The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}.
+The cleavage site is characterized by a variable amino acid composition.
+It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}.
+This site is, however, absent from some membrane proteins in which the first transmembrane domain acts both as a signal peptide and signal anchor~\citep{1988szczesnaskorupapositive}.
 The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.
@@
 \includegraphics[width=0.55\textwidth]{figures/SP.png}
-\caption{The organization of signal peptide. The figure is not drawn to scale.}
+\caption{The organization of typical signal peptide. The figure is not drawn in scale.}
 \label{fig:sparch}
@@
 
-During or after translocation of the protein into the lumen of endoplasmic reticulum, the typical signal peptide is cleaved by a signal peptidase~\citep{2002paetzelsignal} and next degraded by specific proteases, whereas the rest (mature) part of protein stays in the lumen or is passed to other compartments.
-The cleavage site is characterized by a very variable amino acid composition.
-It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}.
-The site is, however, absent from some membrane proteins in which the first transmembrane-domain acts both as signal peptides and signal anchor~\citep{1988szczesnaskorupapositive}.
-
 Some data indicate that signal peptides may be universal.
-It was found, for example, that even bacterial signal peptides targeted correctly transgenic proteins to the plant~\citep{2009moellera} or mammalian secretory system~\citep{2014naganoestablishment}.
+It was found, for example, that even bacterial signal peptides targeted correctly transgenic proteins to plant~\citep{2009moellera} or mammalian secretory systems~\citep{2014naganoestablishment}.
 On the other hand, signal peptides show great variation and the description presented above (Fig.
 \ref{fig:sparch}) refers to the most ‘typical’ signal peptides.
-There are also exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}.
-For example, the fragment of signal peptide from preprolactin probably takes part in the regulation of prolactin secretion.
-Other examples are signal peptides of MHC class I, which inhibit activity of NK cells.
-Interesting functions have signal peptides of viral origin, which are involved in the immune evasion or viral life cycle~\citep{2000kappposttargeting}.
-Such diverse functions are restricted not only to the long signal peptides.
-The peptides from midkine, a protein contributing to the tumor progression, contains epitopes recognized by CD4+ T cells~\citep{2013kerzerhothe}.
-It indicates that the signal peptide may participate in a tumor immunity.
+There are  exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}.
+The fragment of signal peptide from preprolactin takes part in the regulation of prolactin secretion, whereas signal peptides of MHC class I inhibit activity of NK cells.
+Signal peptides of viral origin are involved in the immune evasion or viral life cycle~\citep{2000kappposttargeting}.
+The signal peptide from midkine contributing to the tumor progression contains epitopes recognized by CD4+ T cells~\citep{2013kerzerhothe}.
 The functional significance of these targeting signals makes that the prediction of signal peptide-containing proteins is also an important step in the drug development~\citep{2005zhangalteration, 2012netoadeimproving, 2010moellerwetmilling}.
 
-\subsection*{Signal peptide predicting software}
+\subsection*{Software predicting signal peptides}
 
 Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious.
-Therefore, the development of new approaches in the field of computational biology and bioinformatics is desirable.
-They are not only a good support, complement or alternative for the experimental methods but also enable to understand rules encoding information about protein targeting, which is contained in the predicted targeting signals.
-However, many of the present predictors disregard full biological information carrying by signal peptides.
-
-Having functional importance and specific features, signal peptides became the subject of many programs to their prediction based on different methods.
-The state of the art software for predicting the presence of signal peptides often incorporates ‘black-box’ models, as neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}.
-Such models neither utilize biological information about signal peptides nor work properly on atypical signal peptides.
-
-The other type of software incorporates into the prediction process theoretical knowledge about the structure of signal peptides.
-Although these programs do not share innate flaws of ‘black-box’ models, they also demand an improvement.
-Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}.
-Others (Phobius, Philius and not supported SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework.
-However, the HMMs imply geometric distribution for duration of region length.
-We replicated the rules for extracting regions' boundaries from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the mentioned assumption is not in concordance with the reality because every region revealed the length distribution other than geometric (Fig.
+Therefore, signal peptides became the subject of many computational programs to their prediction.
+Many software incorporates ‘black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}.
+However, these models do not provide direct biological information about organization of signal peptides and are not able to predict properly atypical signal peptides.
+Although there are programs that do not share the innate flaws of ‘black-box’ models, they also demand an improvement.
+Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}.
+Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect regions of signal peptides without realizing limitations of their probabilistic frameworks.
+The used HMMs imply a geometric distribution for duration of regions' length.
+We verified the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was other than geometric (Fig.
 \ref{fig:reglen}).
-Moreover, the commonly used rigid region scheme (Fig.
-\ref{fig:sparch}) does not describe extremely long or short signal peptides.
-Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also unusual structures, but such probabilistic frameworks have not still been implemented.
+Moreover, the commonly used rigid scheme of signal peptide's organization (Fig.
+\ref{fig:sparch}) does not describe extremely long or short peptides.
+Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures but such probabilistic frameworks have not still been implemented.
 
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/reglen.png}
-\caption{A) Distribution of lengths of signal peptides.
-B) Distribution of the lengths of signal peptide regions.
-The measure of the length is a number of amino acids in the subsequence.
+\includegraphics[width=0.95\textwidth]{figures/reglen.eps}
+\caption{Distribution of lengths of signal peptides (A) and their regions (B) expressed in the number of amino acid residues.
 The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
@@
 
-Another feature, shared by majority of the signal peptide predicting software, is the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
-This method of encoding, however, does not preserve relationships between amino acids.
-The distance between residues is the same regardless of the similarity of their physicochemical properties.
-Such behaviour is especially undesired in signal peptides models because their regions are defined by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
-In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.
-
 All programs used in signal peptide recognition are trained on real protein sequences.
-Therefore, they succeed in the recognition of peptides similar to those in the learning set, but fail in the case of artificial signal peptides.
+Therefore, they succeed in the recognition of peptides similar to those in the learning set but fail in the case of artificial signal peptides.
 Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}.
@@
 Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure.
-Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfil the designed role.
+Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfill the designed role.
+
+Majority of the signal peptide predicting software use the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
+This method of encoding, however, does not regard relationships between amino acids and differences in their physicochemical properties.
+This is disadvantage of such signal peptides' models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
+In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.
+Therefore, we elaborated a new approach based on hidden semi-Markov models using grouping of amino acids into physicochemical groups characteristic of signal peptides.
+The new methods proved better in comparison to the current software.
 
@@
 
-The functionality of a signal peptide depends not on exact sequence of specific amino acids, but on the physicochemical properties of residues in a given region.
-Henceforth, the usage of raw amino acid sequences is superfluous and introduces unnecessary information.
-To utilize this property of signal peptide recognition, we clustered amino acids into several groups based on the physicochemical features of residues.
-
-The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions, the enhanced version of algorithm presented in~\cite{1998nielsenprediction}.
+Since the functionality of signal peptides depends the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics.
+The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\cite{1998nielsenprediction}.
 We refined some region recognition criteria to attune the algorithm to less typical signal peptides.
-
-Next, two models are trained to recognize proteins with and without a signal peptide.
-The first one is a hidden semi-Markov model, in which each of three signal peptide regions is represented by a different hidden state.
-The additional fourth hidden state represents mature protein.
-Each state is described by the frequencies of amino acid groups  within that state.
-The distribution of hidden states durations, the number of amino acids related to each hidden state in signal peptide, is based on the empirical density of region lengths from the training set.
+Next, two models were trained to recognize proteins with and without a signal peptide.
+The first one was a hidden semi-Markov model, in which each of three signal peptide regions was represented by a different hidden state.
+The additional fourth hidden state represented a mature protein.
+Each state was described by its frequencies of amino acid groups.
+The distribution of hidden states durations, i.e.
+the number of amino acids, was based on the empirical distribution of region lengths from the training set.
 Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites.
-
-The second model is a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.
+The second model was a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.
 
@@
 The negative set comprised 152~272 sequences without any signal peptide annotation.
-Protein sequences with ambiguous symbols: X, J, Z and B were removed from the final sets.
-Proteins with selenocysteine (U) were also excluded from data set, because there are no records of signal peptides containing this amino acid.
+Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were removed from the final sets.
 
@@
 
-It is worth to note, that signalHsmm is not a first software to use amino acid encoding in signal peptide prediction.
-BLOMAP~\citep{maetschke2005blomap} also employed similar strategy, but considered only encodings based on the substitution matrices.
-Substitution matrices are computed using protein sequences and knowing how atypical signal peptides are, we had chosen different approach.
-
-Instead amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size.
-High hydrophobicity is a determinant of the h-region, the core of signal peptides.
-Alpha-helix, the secondary structure of the h-region, is probably induced by the positively charged n-region.
-High polarity as well as smaller size are important features of the residues in the cleavage site~\citep{1994palzkillselection}.
+SignalHsmm is not the first software which uses amino acid encoding in signal peptide prediction.
+BLOMAP~\citep{maetschke2005blomap} also employed the similar strategy but considered only substitution matrices.
+However, we applied different approach.
+We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size.
+The high hydrophobicity is determinants of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region.
+The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.
 
 \begin{table}[ht]
+\small
 \centering
-\caption{Properties used in clusterization.} 
+\caption{Properties used in amino acid clusterization.} 
 \begin{tabular}{ll}
   \toprule
-Criterion name & Property name \\ 
+Property name & Amino acid scale and its reference \\ 
   \midrule
@@
    \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
-  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
+  Hydrophobicity & Normalized hydrophobicity scales for $\alpha$-proteins~\citep{1992cidhydrophobicity} \\ 
    \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
   Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
-   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
+   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in $\alpha$-helix~\citep{1980ponnuswamyhydrophobic} \\ 
   Polarity & Polarity~\citep{1974granthamamino} \\ 
    \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
-  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
-   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
-  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
+  Frequency in $\alpha$-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
+   \rowcolor[gray]{0.85}Frequency in $\alpha$-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
+  Frequency in $\alpha$-helices & Relative frequency in $\alpha$-helix~\citep{1990prabhakaranthe} \\ 
    \bottomrule
@@
 
-We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Tab.~\ref{tab:aaprop}), each attributed to a single criterion.
-We established 96 permutations of the properties, where each permutation contains only one property associated with a given criterion.
-Considering all permutations of properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
-Further, we cut the clusterings to create four-group encodings.
-As expected, some encodings (31\%) were identical.
-
-To compare encodings we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding.
-We created balanced data sets by subsampling a number of proteins without a signal peptide equal to the number of proteins with signal peptide.
-The analysis was carried using also redundant encodings, but results are not shown (HSMM is a deterministic algorithm and predictions for identical groupings of amino acids were the same).
+We considered in total 13 amino acid scales from AAIndex database~\citep{2008kawashimaaaindex} (Tab.~\ref{tab:aaprop}).
+We selected one scale per property and carried out all possible 96 permutations of them.
+Based on that, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.
+Next, we cut the clusterings to create four group of amino acids.
+In 31\% of cases, the groupings were identical.
+To compare the usefulness of encodings, we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding.
+We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide.
 The cross-validation was repeated 60 times.
-At first, we planned 160 repetitions to ensure with probability 0.95, that every protein without signal peptide was in the learning set, but very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) allowed us to relax this condition wihout sacrificing credibility of cross-validation.
+At first, we planned 160 repetitions to ensure that every protein without signal peptide was in the learning set with the probability of 0.95, but a very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) for smaller number of repetitions allowed us to relax this condition without decreasing credibility of cross-validation.
 
@@
 Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM). 
-As such, it proved to yield better performance in many fields of bioinformatics including
-gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction~\citep{16571137}
-or automatic detection of heartbeats~\citep{7043102}.
-Let us first briefly describe the idea that lies behind HMM. 
-Suppose you have a sequence of observations e.g. amino acids, and you are interested in understanding underlying cause of their occurrence. 
-HMM aims to answer that question by assuming a specific, yet flexible, structure of the problem.
-HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
-They are "the cause" of the observations. At every step $t$, hidden state might change according to transition matrix
-$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
-Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
-Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
+Let us briefly describe the idea of the HMM. 
+Suppose we have a sequence of observations, e.g. amino acids, and we are interested in understanding an underlying cause of their occurrence. 
+HMM aims to answer that question assuming a specific and flexible structure of the problem.
+HMM consists of two stochastic processes. The first is a discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
+They are "the cause" of the observations. At every step $t$, hidden state might change according to a transition matrix
+$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
+Th second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$, which are assumed to occur independently.
+Their distribution depends only on the hidden state which emits it.
 Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
-In our case, observations are (degenerated) amino acids.
-Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
-Viterbi algorithm.
-For a good reference on HMM see~\citep{1989rabinera}.
+In our case, observations are (encoded) amino acids.
+The main goal in signalHsmm is to find for a given peptide most probable regions boundaries. This is achieved with Viterbi algorithm.
+For a good reference on HMM, see~\citep{1989rabinera}.
 
-In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
-\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
-complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
-It is, however, still not flexible enough. Empirical regional length distributions (see Fig.~\ref{fig:reglen})
-are difficult to capture in that way.
+In the regular HMM, the hidden state duration, i.e.
+the number of observations emitted by the hidden state, has a geometric distribution.
+\citep{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational
+complexity. Similar ideas were used for signal peptide recognition, for example by~\citep{2004klla}. 
+However, it is still not flexible enough because the empirical regional length distributions (see Fig.~\ref{fig:reglen})
+are difficult to capture in this way.
 
@@
 
-(1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
-
+\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
+(1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
 \end{tikzpicture}
@@
 
-The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
-any given hidden state duration distribution. Diagram of HSMM is presented in Fig.~\ref{fig:hsmm}.
-In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
+The model we used is in fact Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
+In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
 $$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
-where $D$ is maximum allowed duration.
+where $D$ is the maximum allowed duration.
 As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
-higher than with the regular HMM.
+higher than in the regular HMM.
 
-The model we use has a very specific structure. Hidden states are signal peptide regions.
-Almost all entries in transition matrix $A$ are zeros, because regions are sequential.
+Our model has a very specific structure. Ihe hidden states represent signal peptide regions.
+Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
 Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
-Probabilities of observations for hidden states and hidden states durations were estimated from training data.
+Probabilities of observations for the hidden states and hidden states durations were estimated from training data.
 The advantage of HSMM model is not only a better performance but also its straightforwardness.
@@
 
-\subsection*{n-gram extension}
+\subsection*{Extension of n-grams}
 
 Hidden semi-Markov models may be flexibly enhanced by adding additional hidden states.
-To improve our model, we added few supplementary states representing specific motives that might occur in proximity of cleavage site.
-The structure of cleavage sites, more conservative  than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.
-
-Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides.
-The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of a general model (see Tab.\ref{tab:best}.
-Selected n-grams, representing less common cleavage site motifs, were included in the HSMM model as the alternative paths at the end of c-region.
-The architecture of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.
+To improve our model, we added few supplementary states representing specific motives that might occur in the proximity of cleavage site.
+The structure of cleavage sites, more conserved than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.
+Using the biogram software~\citep{biogramPackage}, we extracted n-grams from cleavage sites of signal peptides.
+The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of the general model (see Tab.\ref{tab:best}.
+Selected n-grams representing less common cleavage site motifs were included in the HSMM model as alternative paths at the end of c-region (Fig.~\ref{fig:ngramext}).
 
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/cvres.png}
-\caption{Sensitivity and specificity of encodings after cross-validation.
-1.
-The encoding providing the best sensitivity (AUC = 0.9683, MCC = 0.8677).
-2.
-The encoding providing the best specificity (AUC = 0.9338, MCC = 0.6474).}
+\includegraphics[width=0.95\textwidth, height=7cm]{figures/cvres.eps}
+\caption{Sensitivity and specificity of amino acid encodings after cross-validation.
+1 indicates the encoding providing the best sensitivity (AUC = 0.9683, MCC = 0.8677), whereas 2 means the encoding providing the best specificity (AUC = 0.9338, MCC = 0.6474).}
 \label{fig:cvres}
@@
 \begin{table}[ht]
+\small
 \begin{minipage}{.5\linewidth} 
@@
 \centering
-\caption{The worst sensitivity encoding.} 
+\caption{The best specificity encoding.} 
 \begin{tabular}{l}
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.95\textwidth]{figures/enccomp.png}
-\caption{A comparison of encodings.
+\includegraphics[width=0.95\textwidth]{figures/enccomp.eps}
+\caption{Comparison of encodings.
 A) Properties of amino acids in the best and the worst sensitivity encoding.
@@
 We used four performance measures to evaluate results of cross-validation: specificity, sensitivity, Matthew's Correlation Coefficient ($\phi$ coefficient) and Area Under the Curve (AUC).
-All encodings provide very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
+All encodings provided very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
 The classification of amino acids has the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
-The final signalHsmm algorithm uses the encoding, that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
+The final signalHsmm algorithm uses the encoding that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
 
 \begin{table}[ht]
+\small
 \centering
@@
 
-We examined more deeply encodings with the best and the worst sensitivity.
-In both cases Group 1 tends to contain polar amino acids typical with the average size of residuals.
-In the best encoding, all charged amino acids, both acids and bases (even as weak bases as histidine), also belong to this group.
-This group, nearly absent in h-region, provides very good distinction between parts of signal peptide for best encoding, because is nearly absent in the h-region.
-In the worst encoding, where polar and charged character of the Group 1 is not that explicit, the difference in frequency between regions is also less visible.
+We examined more in details encodings with the best sensitivity and the best specificity (Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}).
+In both cases, the group 1 tends to contain average-sized polar amino acids.
+In the best sensitivity encoding, all charged amino acids, both acidic and basic (also weakly basic histidine)belong to this group.
+These amino acids are nearly absent from h-region and provide very good distinction between regions of signal peptide.
+In the best specificity encoding, where polar and charged character of the group 1 is not such explicit, the difference in its distribution between the regions is also less visible.
+The amino acids belonging to the group 2 have a quite low probability of occurrence in $\alpha$-helix and are very diverse.
+This group includes polar but uncharged serine and threonine as well as hydrophobic tyrosine, aliphatic glycine and proline.
+In the best specificity encoding, the polar character of this group is emphasized by the addition of asparagine and aspartic acid.
+Despite these differences, the distribution of group 2 seems to be comparable in two encodings.
+The both groupings have strongly non-polar and aliphatic group 3 containing isoleucine, leucine, methionine and valine.
+The hydrophobic property of this group in the best sensitivity encoding are even more pronounced by the presence of tryptophan and phenylalanine.
+On the contrary, more polar histidine belongs to the group 3 in the best specificity encoding.
+Because of the hydrophobic character, this group dominates in the h-region in the case of both amino acid classifications.
+The fourth group is the most diverse in both encodings.
+In the case of the best specificity encoding, this group comprises aromatic amino acids: phenylalanine, tryptophan and tyrosine.
+In contrast, the group 4 in the best sensitivity encoding contains only alanine and cysteine, which both are rather small amino acids and tend to appear in $\alpha$-helices.
+This very unique composition seems to be typical of the c-region of signal peptide.
 
-The amino acids belonging to the Group 2 have quite low probability of occurring in an alpha-helix, but are still very diverse, but all .
-Polar but uncharged serine and threonine neighbour with alyphatic glycine and proline.
-In the worst encoding the polar character of the group is emphasized by the addition of asparagine.
-The best grouping contains also highly hydrophobic tyrosine.
-Despite these differences, the frequency of Group 2 seems to be comparable between two encodings.
+The encoding of amino acid plays crucial role in the recognition of signal peptide but does not affect identification of proteins without signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
+The mature protein state in the HSMM model tends to have a more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
 
-The both groupings have strongly non-polar and aliphatic group 3 (see Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}) containing isoleucine, leucine, methionine and valine.
-Hyrophobic properties of this group in the best encoding are even more pronounced by the presence of tryptophan and phenylalanine.
-On the contrary, to the Group 3 in the worst encoding belong also more polar histidine.
-Hence the hydrophobic character of this group, it is dominant over the h-region in case of both amino acid classifications.
-
-The fourth group is the most diverse among both encodings.
-The key is obvious in a case of the worst encoding.
-Phenylalanine, tryptophan and tyrosine are aromatic, highly hydrophobic amino acids.
-In contrast, the Group 4 in the best encoding contains only alanine and cysteine, which both are rather small amino acids and tend to appear in alpha-helices.
-This very unique composition seems to be typical for the c-region of signal peptide.
-
-The encoding of amino acid plays crucial role in the recognition of signal peptide, but does not affect identification of proteins without a signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
-The mature protein state in the HSMM model tends to have more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
-
-\subsection*{Benchmark test}
+\subsection*{Benchmark tests}
 
@@
 \begin{table}[ht]
+\small
 \centering
@@
 
-To check universality of our probabilistic model, we also validated it on signal peptides belonging to specific taxonomic groups.
-As the example, we chose the Plasmodiidae family including malaria parasites, because of her medical significance.
-The testing data set extracted from UniProt data base contained 102 proteins with a signal peptide and 358 without it.
-We repeated the benchmark scheme for different classifiers (Tab.~\ref{tab:bench2010plas}).
-
 \begin{table}[ht]
+\small
 \centering
-\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
+\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to \textsl{Plasmodiidae}.} 
 \begin{tabular}{lrrrr}
@@
 
-signalHsmm has much better performance that other algorithms on proteins belonging to organisms from Plasmodiidae family.
-Since our learns more general decision rules, it is able to grasp the similarities between very variable sequences.
-Without losing the ability to predict the more common cases, signalHsmm is able to also deal well with very atypical signal peptides.
+To check universality of our probabilistic model, we also validated it on signal peptides belonging to specific taxonomic groups.
+As an example, we chose the \textsl{Plasmodiidae} family including malaria parasites because of their medical significance.
+The testing data set extracted from UniProt data base contained 102 sequences with a signal peptide and 358 sequences without it.
+Interestingly, signalHsmm showed much better performance than other algorithms (Tab.~\ref{tab:bench2010plas}).
+Since our algorithm considers more general decision rules, it is able to recognize also atypical signal peptides.
 
@@
 
-The architecture of existing signal peptide predicting software is usually opaque, preventing analysis of decision rules included in the model.
-Without ability to trace the computations, researchers cannot discover which parameters are responsible for predictions.
+The architecture of existing signal peptide predicting software is usually opaque, preventing analysis of decision rules included in the models.
+Researchers cannot also discover  parameters that are responsible for the predictions.
 Considering the biological context of the problem, we see a need for a transparent model of signal peptide.
-signalHsmm is a first step in this direction, providing users with both prediction results and the extandable, readable predictive model.
-
-Aside from a confirmation of the accepted signal peptide model (hydrophibicity of the h-region, polarity of the n-region), we also discovered new relationships.
-We found the alanine is one of the most typical amino acids for the c-region and occurs there far more often than in the other parts of signal peptides and mature proteins.
-The presence of alanine in the signal peptide cleavage site~\cite{1986vonheijnea} cannot explain such over-representation of group containing alanine in the c-region.
-
-The good performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence, but on the physicochemical features of the amino acids.
-Although this finding may seem to offer a minimal improvement when we consider the most popular signal peptides, this generalization pays off when we analyze more specific cases, as signal peptides of \textit{Plasmodiidae}.
-
+SignalHsmm is the first step in this direction providing users with both prediction results and the extendable, readable predictive model.
+Aside from the confirmation of the accepted signal peptide model (hydrophibicity of the h-region and polarity of the n-region), we also found that the alanine is one of the most typical amino acids in the c-region in comparison to others.
+The good performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence but on the physicochemical features of the amino acids. 
 The flexibility and efficient information recovery makes our model unique among similar software.
 The signalHsmm not only shows very good performance but is also adjustable enough to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.
-Morrover, our method can effectively extract information from very small data sets, which in future may lead to new predictors, able to recognize less known signaling sequences.
+Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors able to recognize atypical signaling sequences.
 
 \subsection*{Availability and implementation}
-The signalHsmm prediction web-server is avaible under address: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
-
-signalHsmm is implemented as an R package freely available on CRAN: \url{http://cran.r-project.org/web/packages/signalHsmm}.
-Stand-alone version, in addition to the prediction of signal peptides, offers tools to build, train and test novel signal peptide models.
-
-\section*{Acknowledgments}
-PS and PB devised the HSMM model of signal peptide. MB and PS co-wrote the software. MB gathered data and carried the analysis.
-MB, PS and PM drafted the manuscript. All authors revised the manuscript and approved the final version.
+The signalHsmm prediction web-server is available at: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
+SignalHsmm is implemented as an R package available at: \url{http://cran.r-project.org/web/packages/signalHsmm}.
+Stand-alone version offers prediction and tools to build, train and test novel signal peptide models.
 

### 2015-06-20 paper.tex
small corrections

@@
 Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction.
-Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation.
-The proposed approach is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
+Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability.
+The proposed approach is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides.
 The big advantage of the algorithm is its extensibility.
-Using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
-Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even after training on very small data sets.
-Moreover, it appeared very stable regardless of types of learning data sets.
+By using n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
+Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor despite the fact it was trained on very small data set.
+Moreover, it appears to be very stable regardless of types of learning data sets.
 Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
@@
 Information about the protein destination is included within the very protein in short stretches of amino acid residues called targeting or sorting signals.
-Ones of them are signal peptides, which are located at the N-terminus of proteins.
+One kind of them are signal peptides, which are located at the N-terminus of proteins.
 
 Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum and Golgi apparatus.
-Such proteins can stay inside of these compartments, can be inserted into cellular membranes or exported outside the cell.
-Proteins equipped with signal peptides play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}.
+Such proteins can stay inside these compartments, can be inserted into cellular membranes or exported outside the cell.
+Proteins equipped with signal peptides play crucial role in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}.
 Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.
@@
 It constitutes a core region of signal peptide and usually forms $\alpha$-helix.
-The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}.
+The third part of a signal peptide is a polar, but uncharged c-region.
+It is usually 6 residues long and ends with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}.
 The cleavage site is characterized by a variable amino acid composition.
@@
 Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}.
-Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect regions of signal peptides without realizing limitations of their probabilistic frameworks.
-The used HMMs imply a geometric distribution for duration of regions' length.
-We verified the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was other than geometric (Fig.
+Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect structure of signal peptides regions in their limited probabilistic frameworks.
+The HMMs they use, however, imply a geometric distribution for duration of regions length.
+We studied the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was not geometric (Fig.
 \ref{fig:reglen}).
@@
 \ref{fig:sparch}) does not describe extremely long or short peptides.
-Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures but such probabilistic frameworks have not still been implemented.
+Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures, but such probabilistic frameworks have not yet been implemented.
 
@@
 
-Majority of the signal peptide predicting software use the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
-This method of encoding, however, does not regard relationships between amino acids and differences in their physicochemical properties.
+Majority of the signal peptide predicting software uses the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid.
+This method of encoding, however, does not take into account relationships between amino acids and differences in their physicochemical properties.
 This is disadvantage of such signal peptides' models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
@@
 
-Since the functionality of signal peptides depends the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics.
+Since the functionality of signal peptides depends on the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics.
 The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\cite{1998nielsenprediction}.
@@
 Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06.
-The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, its start and cleavage site.
+The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, containing its start and cleavage site information.
 Sequences with more than one cleavage site were excluded from the final data set.
@@
 We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size.
-The high hydrophobicity is determinants of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region.
+The high hydrophobicity is determinant of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region.
 The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.
@@
 We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide.
-The cross-validation was repeated 60 times.
-At first, we planned 160 repetitions to ensure that every protein without signal peptide was in the learning set with the probability of 0.95, but a very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) for smaller number of repetitions allowed us to relax this condition without decreasing credibility of cross-validation.
+The cross-validation was repeated 60 times to ensure that every protein without signal peptide was included in the learning set with probability higher than 0.5.
+Very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas})  confirms credibility of cross-validation.
 
@@
 Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM). 
-Let us briefly describe the idea of the HMM. 
+Let us first briefly describe the idea of the HMM. 
 Suppose we have a sequence of observations, e.g. amino acids, and we are interested in understanding an underlying cause of their occurrence. 
-HMM aims to answer that question assuming a specific and flexible structure of the problem.
+HMM aims to answer that question assuming a specific and yet flexible structure of the problem.
 HMM consists of two stochastic processes. The first is a discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
@@
 $A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
-Th second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$, which are assumed to occur independently.
-Their distribution depends only on the hidden state which emits it.
-Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
+The second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$,.
+They are assumed to occur independently conditionally on the hidden state which emits it.
+Their distribution probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
 In our case, observations are (encoded) amino acids.
 The main goal in signalHsmm is to find for a given peptide most probable regions boundaries. This is achieved with Viterbi algorithm.
-For a good reference on HMM, see~\citep{1989rabinera}.
+For a good reference on HMM see~\citep{1989rabinera}.
 
@@
 
-The model we used is in fact Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
+The model we used is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
 In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
@@
 where $D$ is the maximum allowed duration.
-As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
+As our datasets are of reasonable size and $D$ is small -- around 30, computational effort is not much 
 higher than in the regular HMM.
 
-Our model has a very specific structure. Ihe hidden states represent signal peptide regions.
+Our model has a very specific structure. The hidden states represent signal peptide regions.
 Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
@@
 
-We examined more in details encodings with the best sensitivity and the best specificity (Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}).
+We examined in more detail encodings with the best sensitivity and the best specificity (Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}).
 In both cases, the group 1 tends to contain average-sized polar amino acids.
@@
 These amino acids are nearly absent from h-region and provide very good distinction between regions of signal peptide.
-In the best specificity encoding, where polar and charged character of the group 1 is not such explicit, the difference in its distribution between the regions is also less visible.
+In the best specificity encoding, where polar and charged character of the group 1 is not so explicit, the difference in its distribution between the regions is also less visible.
 The amino acids belonging to the group 2 have a quite low probability of occurrence in $\alpha$-helix and are very diverse.
@@
 The encoding of amino acid plays crucial role in the recognition of signal peptide but does not affect identification of proteins without signal peptides (compare change in specificity and sensitivity on Fig.~\ref{fig:cvres}).
-The mature protein state in the HSMM model tends to have a more uniform distribution of residues than signal peptide region, which make it more resistant to changes in amino acid groupings.
+The mature protein state in the HSMM model tends to have a more uniform distribution of residues than signal peptide region, which makes it more resistant to changes in amino acid groupings.
 
@@
 The signalHsmm-1987 was able to very accurately predict a signal peptide, even better than predictors trained on richer data sets.
-It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets~\ref{tab:bench2010}.
+It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets (see Tab.~\ref{tab:bench2010}).
 
@@
   \midrule
-signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9416 & 0.9720 & 0.9112 & 0.8848 \\ 
-   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9673 & 0.9579 & 0.9766 & 0.9347 \\ 
+signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.9416 & \textbf{0.9720} & 0.9112 & 0.8848 \\ 
+   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.9673 & 0.9579 & \textbf{0.9766} & \textbf{0.9347} \\ 
   PrediSi \citep{2004hillerpredisi} & 0.8949 & 0.9065 & 0.8832 & 0.7899 \\ 
@@
   signalHsmm-1989 & 0.9562 & 0.9626 & 0.8972 & 0.8617 \\ 
-   \rowcolor[gray]{0.85}signalHsmm-2010 with k-mers & 0.9679 & 0.9673 & 0.9112 & 0.8799 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 with k-mers & \textbf{0.9679} & 0.9673 & 0.9112 & 0.8799 \\ 
    \bottomrule
@@
 signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
-   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & 0.9469 & 0.4028 \\ 
+   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & \textbf{0.9469} & 0.4028 \\ 
   Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
    \rowcolor[gray]{0.85}Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
-  signalHsmm-2010 & 0.9340 & 0.7941 & 0.8939 & 0.6526 \\ 
-   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & 0.7531 \\ 
-  signalHsmm-2010 with k-mers & 0.9334 & 0.9902 & 0.7989 & 0.6767 \\ 
+  signalHsmm-2010 & \textbf{0.9340} & 0.7941 & 0.8939 & 0.6526 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & \textbf{0.7531} \\ 
+  signalHsmm-2010 with k-mers & 0.9334 & \textbf{0.9902} & 0.7989 & 0.6767 \\ 
    \bottomrule
@@
 
-The architecture of existing signal peptide predicting software is usually opaque, preventing analysis of decision rules included in the models.
-Researchers cannot also discover  parameters that are responsible for the predictions.
+The architecture of existing signal peptide predicting software is usually opaque which makes analysis of how its decisions are made impossible.
+Researchers cannot discover which parameters are responsible for the predictions.
 Considering the biological context of the problem, we see a need for a transparent model of signal peptide.
-SignalHsmm is the first step in this direction providing users with both prediction results and the extendable, readable predictive model.
+SignalHsmm is the first step in this direction.
+It provides users good performance as well as interpretability and expendability of predictive model. 
 Aside from the confirmation of the accepted signal peptide model (hydrophibicity of the h-region and polarity of the n-region), we also found that the alanine is one of the most typical amino acids in the c-region in comparison to others.
-The good performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence but on the physicochemical features of the amino acids. 
+The performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence but on the physicochemical features of the amino acids. 
 The flexibility and efficient information recovery makes our model unique among similar software.
-The signalHsmm not only shows very good performance but is also adjustable enough to model properly very specific signal peptides belonging to poorly represented taxonomic groups in databases.
+The signalHsmm is also adjustable enough to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
 Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors able to recognize atypical signaling sequences.

### 2015-06-20 paper.tex
revamped abstract

@@
 \begin{abstract}
-Short stretches of amino acid residues called targeting signals are involved in the proper localization of proteins in a cell, which is essential to maintain their desired function.
-Ones of them are signal peptides, which are responsible for targeting of proteins to endomembrane system and their export outside the cell.
+Signal peptides are responsible for targeting of proteins to endomembrane system and their export outside the cell.
 Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
-Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction.
-Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability.
-The proposed approach is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides.
+They may be recognized with adequate accuracy using learning systems, that are usually 'black-box' models.
+Without an ability to trace decision rules, researchers cannot discover which parameters are responsible for predictions.
+Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation.
+The proposed approach is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides.
 The big advantage of the algorithm is its extensibility.
-By using n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
-Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor despite the fact it was trained on very small data set.
-Moreover, it appears to be very stable regardless of types of learning data sets.
+Using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
+Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even after training on very small data sets.
+Moreover, it appeared very stable regardless of types of learning data sets.
 Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.

### 2015-06-20 paper.tex
making corrections to  abstract

--- 2015-06-20 paper.tex
@@
 \begin{abstract}
-Short stretches of amino acid residues called targeting signals are involved in the proper localization of proteins in a cell, which is essential to maintain their desired function.
-Ones of them are signal peptides, which are responsible for targeting of proteins to endomembrane system and their export outside the cell.
+We provide high-performance tool for recognizing signal peptides, which prediction results are easy to interpret 
+and which structure is readily expandable by the researcher.
+Signal peptides play important role in cell biology as they are responsible for targeting of proteins to endomembrane system and their export outside the cell.
 Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
-Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction.
+They may be recognized with adequate accuracy using learning systems, that are usually 'black-box' models.
+Without an ability to trace decision rules, researchers cannot discover which parameters are responsible for predictions.
 Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability.
@@
 The big advantage of the algorithm is its extensibility.
-By using n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
-Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor despite the fact it was trained on very small data set.
-Moreover, it appears to be very stable regardless of types of learning data sets.
+By using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides.
+Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even though it was trained on very small data sets.
+Moreover, it proved to be very stable regardless of types of learning data sets.
 Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
-It should be emphasized that our model described signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
+It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
 The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.

### 2015-06-20 paper.tex
first abstract sentence again

@@
 \begin{abstract}
-We provide high-performance tool for recognizing signal peptides, which prediction results are easy to interpret 
-and which structure is readily expandable by the researcher.
+signalHsmm using universal, transparent and extendable model of signal peptides, predicts their presence in eukaryotic proteins.
 Signal peptides play important role in cell biology as they are responsible for targeting of proteins to endomembrane system and their export outside the cell.

### 2015-06-20 paper.tex
improved caption

@@
 \includegraphics[width=0.95\textwidth]{figures/enccomp.eps}
-\caption{Comparison of encodings.
-A) Properties of amino acids in the best and the worst sensitivity encoding.
-B) Frequencies of amino acids belonging to the encodings in regions of signal peptide and mature proteins.}
+\caption{Comparison of amino acid encodings with the best sensitivity and the best
+specificity in different regions of signal peptide and mature proteins according to: A) the normalized value of
+property; points represent amino acids; B) frequencies of amino acids in the given region.}
 \label{fig:enccomp}
@@
 The flexibility and efficient information recovery makes our model unique among similar software.
-The signalHsmm is also adjustable enough to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
+signalHsmm is also adjustable enough to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
 Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors able to recognize atypical signaling sequences.

### 2015-06-20 paper.tex
caption improved

@@
 specificity in different regions of signal peptide and mature proteins according to: A) the normalized value of
-property; points represent amino acids; B) frequencies of amino acids in the given region.}
+properties for particular amino acids (points) and B) frequencies of amino acids in the given region.}
 \label{fig:enccomp}

### 2015-06-20 paper.tex
finished!

@@
 \author[1]{Pawe\l{} Mackiewicz}
-\affil[1]{University of Wroc\l{}aw, Department of Genomics, Poland}
-\affil[2]{Wroc\l{}aw University of Technology, Department of Mathematics, Poland}
+\affil[1]{University of Wroc\l{}aw, Department of Genomics, Poland; michalburdukiewicz@gmail.com, blazej.pawel@gmail.com, pamac@smorfland.uni.wroc.pl}
+\affil[2]{Wroc\l{}aw University of Technology, Department of Mathematics, Poland; piotr.sobczyk@pwr.edu.pl}
 
@@
 \begin{abstract}
-signalHsmm using universal, transparent and extendable model of signal peptides, predicts their presence in eukaryotic proteins.
-Signal peptides play important role in cell biology as they are responsible for targeting of proteins to endomembrane system and their export outside the cell.
-Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
-They may be recognized with adequate accuracy using learning systems, that are usually 'black-box' models.
-Without an ability to trace decision rules, researchers cannot discover which parameters are responsible for predictions.
-Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability.
-The proposed approach is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides.
+Signal peptides play an essential role in targeting of proteins to endomembrane system and their export outside the cell.
+The proteins equipped with such peptides are of great importance in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.
+These peptides may be recognized with adequate accuracy using learning systems.
+However, they are usually 'black-box' models therefore unable to trace decision rules and identify parameters responsible for the predictions.
+That is why, we designed a new more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability.
+The proposed approach called signalHsmm is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides.
 The big advantage of the algorithm is its extensibility.
@@
 Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
-It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
+It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textit{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
 The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.
@@
 The third part of a signal peptide is a polar, but uncharged c-region.
-It is usually 6 residues long and ends with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}.
+It is usually 6 residues long and ends with a cleavage site, in which a signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}.
 The cleavage site is characterized by a variable amino acid composition.
@@
 \begin{figure}[ht]\centering
-\includegraphics[width=0.55\textwidth]{figures/SP.png}
-\caption{The organization of typical signal peptide. The figure is not drawn in scale.}
+\includegraphics[width=0.55\textwidth]{figures/SP.eps}
+\caption{The organization of typical signal peptide. The regions are not drawn in scale.}
 \label{fig:sparch}
@@
 Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect structure of signal peptides regions in their limited probabilistic frameworks.
-The HMMs they use, however, imply a geometric distribution for duration of regions length.
-We studied the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was not geometric (Fig.
+The used HMMs, however, imply a geometric distribution for duration of regions length.
+We studied the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the length distribution for every region was not geometric (Fig.
 \ref{fig:reglen}).
@@
 \ref{fig:sparch}) does not describe extremely long or short peptides.
-Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures, but such probabilistic frameworks have not yet been implemented.
+Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures but such probabilistic frameworks have not yet been implemented.
 
@@
 Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure.
-Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfill the designed role.
+Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey \textit{in silico} many artificial peptides to select the ones that would fulfill the designed role.
 
@@
 This method of encoding, however, does not take into account relationships between amino acids and differences in their physicochemical properties.
-This is disadvantage of such signal peptides' models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
+This is disadvantage of such signal peptide models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids.
 In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.
@@
 Since the functionality of signal peptides depends on the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics.
-The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\cite{1998nielsenprediction}.
+The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\citep{1998nielsenprediction}.
 We refined some region recognition criteria to attune the algorithm to less typical signal peptides.
@@
 The negative set comprised 152~272 sequences without any signal peptide annotation.
-Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were removed from the final sets.
+Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were finally removed.
 
@@
 We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size.
-The high hydrophobicity is determinant of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region.
+The high hydrophobicity is a determinant of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region.
 The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.
@@
 In 31\% of cases, the groupings were identical.
-To compare the usefulness of encodings, we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding.
+To compare the usefulness of encodings, we performed a 5-fold cross-validation training the new instance of signalHsmm on every encoding.
 We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide.
-The cross-validation was repeated 60 times to ensure that every protein without signal peptide was included in the learning set with probability higher than 0.5.
-Very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas})  confirms credibility of cross-validation.
+The cross-validation was repeated 60 times to ensure that every protein without signal peptide was included in the learning set with the probability higher than 0.5.
+Very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas})  confirmed the credibility of cross-validation.
 
@@
 $A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
-The second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$,.
-They are assumed to occur independently conditionally on the hidden state which emits it.
-Their distribution probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
-In our case, observations are (encoded) amino acids.
-The main goal in signalHsmm is to find for a given peptide most probable regions boundaries. This is achieved with Viterbi algorithm.
+The second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$.
+They are assumed to occur independently and conditionally on the hidden state that emits it.
+Their distribution are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
+In our case, the observations are (encoded) amino acids.
+The main goal in signalHsmm is to find the most probable regions boundaries for a given peptide. This is achieved with Viterbi algorithm.
 For a good reference on HMM see~\citep{1989rabinera}.
@@
 the number of observations emitted by the hidden state, has a geometric distribution.
-\citep{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational
-complexity. Similar ideas were used for signal peptide recognition, for example by~\citep{2004klla}. 
+\cite{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational
+complexity. Similar ideas were used for signal peptide recognition, for example by~\cite{2004klla}. 
 However, it is still not flexible enough because the empirical regional length distributions (see Fig.~\ref{fig:reglen})
@@
 \centering
-\tikzstyle{block} = [draw,shape=circle, top color=white, bottom color=green!50!red!70 ,minimum size=4em]
-\tikzstyle{blockComp} = [draw,shape=circle, top color=white, bottom color=red!50!black!70 ,minimum size=2.5em]
+\tikzstyle{block} = [draw,shape=circle, top color=green!50!white!70, bottom color=green!50!white!70 ,minimum size=4em]
+\tikzstyle{blockComp} = [draw,shape=circle, top color=red!50!white!70, bottom color=red!50!white!70 ,minimum size=2.5em]
 % diameter of semicircle used to indicate that two lines are not connected
 \def\radius{.7mm} 
-\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=green!50!red!70 ,minimum size=3pt,inner sep=0pt]
-\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=red!50!black!70 ,minimum size=3pt,inner sep=0pt]
+\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=white ,minimum size=3pt,inner sep=0pt]
+\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=white ,minimum size=3pt,inner sep=0pt]
 \def\n{11}
- \tikzstyle{line} = [draw=black, color=black!30!white!50, line width=1.5mm, -latex'] 
- \tikzstyle{line2} = [draw=black, color=red!30!black!50, line width=1.5mm, -latex']
+ \tikzstyle{line} = [draw=black, color=black!70!white!50, line width=1.5mm, -latex'] 
+ \tikzstyle{line2} = [draw=black, color=red!30!white!70, line width=1.5mm, -latex']
 \tikzstyle{frame}=[text=black,above, bottom color=white, top color=blue!50!black!70  ]
-\tikzstyle{frameComp}=[text=black,below, bottom color=white, top color=red!50!black!70  ]
- \tikzstyle{line3} = [color=green!50!red!70, line width=1.5mm]
+\tikzstyle{frameComp}=[text=black,below, bottom color=red!50!white!70, top color=red!50!white!70  ]
+ \tikzstyle{line3} = [color=green!50!black!70, line width=1.5mm]
 \def\names{{"$O_{1,1}$","...","$O_{1,d_1}$", "...","$O_{k, 1}$", "...", "$O_{k, d_k}$"}}%
@@
 \draw[line,->]  (block1) --  (komp3);
-\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
+\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.3cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
 
-\node[block, bottom color=green!50!red!80,] at (4.5+2,0) (block2) {...};
-\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
+\node[block, bottom color=green!50!white!80,] at (4.5+2,0) (block2) {...};
+\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.3cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
 \node[blockComp] at (4.5+2,-2)  (komp4) {\footnotesize \pgfmathparse{\names[3]}\pgfmathresult};
@@
 
-\node[block,text width=3em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
+\node[block,text width=3em, bottom color=green!50!white!70] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
 \node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
@@
 
-\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
+\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
 (1,-1.8) -- (5,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_1$}};
 
-\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
+\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
 (8,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_k$}};
 
-\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
+\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
 (1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
@@
 
-The model we used is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
+The model we used is Hidden semi-Markov Model (HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
 In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
@@
 Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
-Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
+Possible transitions are depicted as arrows in Fig.~\ref{fig:ngramext}.
 Probabilities of observations for the hidden states and hidden states durations were estimated from training data.
 The advantage of HSMM model is not only a better performance but also its straightforwardness.
-Fig.~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.
+Fig.~\ref{fig:ngramext} is easy to interpret for a researcher without any mathematical background.
 
 %model structure
-\begin{figure}
- \begin{minipage}{.5\linewidth} 
-     \resizebox{6.5cm}{!}{%
-    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
-    \label{fig:model_hsmm}
-      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
-      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
-         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
-      %nodes
-      \node[main node]            (start') 	[]						{Start};	     
-      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
-      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
-      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
-      \node[main node, bottom color=green!70!black!70] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};
 
-      %lines
-      \path [line] (start')   edge node [left, color=black] {} (nregion');
-      \path [line] (nregion') edge node [below, color=black] { } (hregion');
-      \path [line] (hregion') edge node [below, color=black] { } (cregion');
-      \path [line] (cregion') edge node [left, color=black] { } (mature');
-      \draw [line] (start') to[out=340,in=200] (mature');
-    \end{tikzpicture} }
-    \caption{The diagram of signalHsmm.}
-    \label{fig:signalHsmm}
-\end{minipage}
- \begin{minipage}{.5\linewidth}
-     \resizebox{6.5cm}{!}{%
-    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
-    \label{fig:model_hsmm_kMer}
-      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
-      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
-         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
-      %nodes
-      \node[main node]            (start')   []  					{Start};	     
-      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-25mm, yshift=35mm] 	{n-region};
-      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-25mm,yshift=35mm] 	{h-region};
-      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-10mm,yshift=-25mm] 	{c-region};
-      \node[main node, bottom color=green!70!black!70] 	(mature') 	[below of=hregion',xshift=15mm, yshift=-90mm] 	{Mature protein};
-
-      \node[main node, bottom color=yellow!70!black!70]   (kmer1) 	[below of=cregion',xshift=35mm,yshift=30mm] 	{Aminoacid 1};
-      \node[main node, bottom color=yellow!70!black!70]   (kmer2) 	[below of=kmer1,xshift=-5mm,yshift=5mm] 	{Aminoacid 2};
-      \node[main node, bottom color=yellow!70!black!70]   (kmer3) 	[below of=kmer2,xshift=-5mm,yshift=5mm] 	{Aminoacid 3};
-      %lines
-      \path [line] (start')   edge node [left, color=black] {} (nregion');
-      \path [line] (nregion') edge node [below, color=black] { } (hregion');
-      \path [line] (hregion') edge node [below, color=black] { } (cregion');
-      \path [line] (cregion') to[out=230,in=90] (mature');
-      \draw [line] (start') to[out=300,in=180] (mature');
-
-      \draw [line] (cregion') edge node [below, color=black] { } (kmer1);
-      \draw [line] (kmer1) edge node [left, color=black] {} (kmer2);
-      \draw [line] (kmer2) edge node [left, color=black] {} (kmer3);
-      \draw [line] (kmer3) edge node [left, color=black] {} (mature');
-
-      %\draw [decorate,line width=3pt,decoration={brace,amplitude=10pt}] ([xshift=5mm, yshift=5mm]kmer1) -- ([yshift=-25mm]kmer3) node [black,midway,xshift=0.6cm] {\footnotesize $P_1$};
-      \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
-      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{black!90} n-gram};
-    \end{tikzpicture} }
-    \caption{The diagram of signalHsmm extended with the n-gram cleavage site model.}
+\begin{figure}[ht]\centering
+\includegraphics[width=0.51\textwidth]{figures/HSMMs.eps}
+\caption{The diagram of simple (A) and extended version of signalHsmm with the n-gram cleavage site model (B).}
 \label{fig:ngramext}
-    \end{minipage}
 \end{figure}
@@
 To improve our model, we added few supplementary states representing specific motives that might occur in the proximity of cleavage site.
-The structure of cleavage sites, more conserved than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.
+The structure of cleavage sites, more conserved than other parts of signal peptide~\citep{2004hillerpredisi}, may be reflected by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.
 Using the biogram software~\citep{biogramPackage}, we extracted n-grams from cleavage sites of signal peptides.
-The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of the general model (see Tab.\ref{tab:best}.
-Selected n-grams representing less common cleavage site motifs were included in the HSMM model as alternative paths at the end of c-region (Fig.~\ref{fig:ngramext}).
+The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of the general model (Tab.
+\ref{tab:best}).
+Selected n-grams representing less common cleavage site motifs were included in the HSMM model as alternative paths at the end of c-region (Fig.~\ref{fig:ngramext}B).
 
@@
 \caption{Comparison of amino acid encodings with the best sensitivity and the best
-specificity in different regions of signal peptide and mature proteins according to: A) the normalized value of
-properties for particular amino acids (points) and B) frequencies of amino acids in the given region.}
+specificity in different regions of signal peptide and mature proteins according to: A) the normalized value of properties for particular amino acids (points) and B) frequencies of amino acids in the given region.}
 \label{fig:enccomp}
@@
 All encodings provided very good AUC (0.93 -- 0.97) and specificity (0.92 -- 0.96).
-The classification of amino acids has the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
-The final signalHsmm algorithm uses the encoding that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (see Fig.~\ref{fig:cvres}).
+The classification of amino acids had the biggest impact on sensitivity, which ranges from 0.66 to 0.94.
+The final signalHsmm algorithm uses the encoding that yields the highest specificity and Matthew's Correlation coefficient as well as the second best AUC (Fig.~\ref{fig:cvres}).
 
@@
 
-We examined in more detail encodings with the best sensitivity and the best specificity (Tab.~\ref{tab:best}, Tab.~\ref{tab:best} and Fig.~\ref{fig:enccomp}).
+We examined in detail encodings with the best sensitivity and the best specificity (Tab.~\ref{tab:best}, Tab.~\ref{tab:worst} and Fig.~\ref{fig:enccomp}).
 In both cases, the group 1 tends to contain average-sized polar amino acids.
-In the best sensitivity encoding, all charged amino acids, both acidic and basic (also weakly basic histidine)belong to this group.
-These amino acids are nearly absent from h-region and provide very good distinction between regions of signal peptide.
+In the best sensitivity encoding, all charged amino acids, both acidic and basic (also weakly basic histidine)   belong to this group.
+These amino acids are nearly absent from h-region and provide very good distinction between regions of signal peptide (Fig.~\ref{fig:enccomp}).
 In the best specificity encoding, where polar and charged character of the group 1 is not so explicit, the difference in its distribution between the regions is also less visible.
@@
 The signalHsmm-1987 was able to very accurately predict a signal peptide, even better than predictors trained on richer data sets.
-It indicates that signalHsmm is very stable and is able to recover the structure of signal peptides from even very small data sets (see Tab.~\ref{tab:bench2010}).
+It indicates that signalHsmm is very stable and able to recover the structure of signal peptides from even very small data sets (Tab.~\ref{tab:bench2010}).
 
@@
 \centering
-\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to \textsl{Plasmodiidae}.} 
+\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to \textit{Plasmodiidae}.} 
 \begin{tabular}{lrrrr}
@@
 To check universality of our probabilistic model, we also validated it on signal peptides belonging to specific taxonomic groups.
-As an example, we chose the \textsl{Plasmodiidae} family including malaria parasites because of their medical significance.
+As an example, we chose the \textit{Plasmodiidae} family including malaria parasites because of their medical significance.
 The testing data set extracted from UniProt data base contained 102 sequences with a signal peptide and 358 sequences without it.
@@
 
-The architecture of existing signal peptide predicting software is usually opaque which makes analysis of how its decisions are made impossible.
-Researchers cannot discover which parameters are responsible for the predictions.
+The architecture of existing signal peptide predicting software is usually opaque, which unables to extract decision rules and determine parameters responsible for the predictions.
 Considering the biological context of the problem, we see a need for a transparent model of signal peptide.
 SignalHsmm is the first step in this direction.
-It provides users good performance as well as interpretability and expendability of predictive model. 
-Aside from the confirmation of the accepted signal peptide model (hydrophibicity of the h-region and polarity of the n-region), we also found that the alanine is one of the most typical amino acids in the c-region in comparison to others.
-The performance of signalHsmm confirms that properties of signal peptides do not depend on their exact sequence but on the physicochemical features of the amino acids. 
+It shows a good performance and provides clear interpretability and expendability of  the predictive algorithm. 
+The model confirmed not only the high hydrophibicity of the h-region and polarity of the n-region but also found that alanine is one of the most typical amino acids in the c-region.
+The performance of signalHsmm indicates that properties of signal peptides do not depend on their exact sequence but on the physicochemical features of the amino acids. 
 The flexibility and efficient information recovery makes our model unique among similar software.
-signalHsmm is also adjustable enough to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
-Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors able to recognize atypical signaling sequences.
+SignalHsmm is also adjustable to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
+Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors specialized in recognition of atypical signaling sequences.
 

### 2015-06-20 encodings_vs_regions.R

@@ -60,6 +60,8 @@ source("plot_tools.R")
 
 levels(deg_region[["region"]]) <- c("n-region", "h-region", "c-region", "Mature\nprotein")
 levels(deg_region[["group"]]) <- paste0("Group ", levels(deg_region[["group"]]))
+levels(deg_region[["enc"]]) <- c("Best\nsensitivity", "Best\nspecificity")
+
 
 p3 <- ggplot(deg_region, aes(x = region, y = freq, fill = enc, colour = enc)) +
   geom_bar(stat = "identity", position = "dodge") + 
@@ -87,6 +89,7 @@ dat_worst <- group_properties(group_worst)
 
 dat_bestworst <- cbind(enc = unlist(lapply(c("best", "worst"), function(i) rep(i, nrow(dat_best)))),
                        rbind(dat_best, dat_worst))
+levels(dat_bestworst[["enc"]]) <- c("Best\nsensitivity", "Best\nspecificity")
 
 # p1 <- ggplot(dat_best, aes(x = critertion, y = value, label = variable)) +
 #   geom_point(size = 5, shape = 21, colour = "blue", fill = adjustcolor("blue", 0.25)) +

### 2015-06-20 paper.tex
unified names

@@
 
-\title{SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides}
+\title{signalHsmm - a novel semi-Markov model of eukaryotic signal peptides}
 
@@
 
-SignalHsmm is not the first software which uses amino acid encoding in signal peptide prediction.
+signalHsmm is not the first software which uses amino acid encoding in signal peptide prediction.
 BLOMAP~\citep{maetschke2005blomap} also employed the similar strategy but considered only substitution matrices.
@@
 Considering the biological context of the problem, we see a need for a transparent model of signal peptide.
-SignalHsmm is the first step in this direction.
+signalHsmm is the first step in this direction.
 It shows a good performance and provides clear interpretability and expendability of  the predictive algorithm. 
@@
 The flexibility and efficient information recovery makes our model unique among similar software.
-SignalHsmm is also adjustable to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
+signalHsmm is also adjustable to model properly very specific signal peptides belonging to taxonomic groups which are poorly represented in databases.
 Moreover, our method can effectively extract information from very small data sets, which in future may lead to new predictors specialized in recognition of atypical signaling sequences.
@@
 The signalHsmm prediction web-server is available at: \url{http://smorfland.uni.wroc.pl/signalhsmm}.
-SignalHsmm is implemented as an R package available at: \url{http://cran.r-project.org/web/packages/signalHsmm}.
+signalHsmm is implemented as an R package available at: \url{http://cran.r-project.org/web/packages/signalHsmm}.
 Stand-alone version offers prediction and tools to build, train and test novel signal peptide models.

### 2015-06-22 signalHSMM_kMer_report.Rmd
adding duration estimated with kernel density estimator

@@
 ```{r, cache=TRUE}
-other_soft <- read.csv2("benchmark_other_new.csv")
+other_soft <- read.csv2("benchmark_other.csv")
 
@@
 benchmark_data <- read.fasta("benchmark_data.fasta", seqtype = "AA")
-benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:218)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
+benchmark_data2 <- lapply(c(benchmark_full,benchmark_data[-c(1:214)]), function(i) i[1L:ifelse(length(i) > 80, 80, length(i))])
 
@@
 ```
+
+Now, what happens when we use kernel density estimation?
+```{r}
+params = structure(c(0.00345173057742305216, 0.16303340280710845733, 0.18328936325298567089, 0.32191072063759601507, 0.12710730373689624328, 0.19479707117476749434, 0.07149501410718961392, 0.05102665325739586671, 0.02061953603483871952, 0.01232173478541280856, 0.00339291179532653139, 0.00314522392528060902, 0.00050543218629833602, 0.00000944936698432368, 0.00000000009143740374, 0.00000000000000000000, 0.00000000000000000004, 0.00000000000000000321, 0.00000000000000000290, 0.00000000000000000501, 0.00000000000000000835, 0.00000000000000000088, 0.00000000000000000146, 0.00000000000000000000, 0.00000000000000000592, 0.00000000000000000000, 0.00000000000000000345, 0.00000000000000000036, 0.00000000000000000132, 0.00000000000000000276, 0.00000000000000000001, 0.00000000000000000462, 0.00119575265523844728, 0.00135312474054827118, 0.00278142092346731536, 0.00184134527771419310, 0.00400555470679759759, 0.01737188174042058122, 0.04070639179659169687, 0.06492241958211328323, 0.10486627831232180064, 0.13185802303839422311, 0.16522286797419358240, 0.14092085504016543807, 0.13473723690401800956, 0.08898241528270439682, 0.05443355717845509839, 0.04403282913502165608, 0.02561338612799781561, 0.01320007003819387639, 0.01069067991482806901, 0.00517725596697340799, 0.00181848832708673882, 0.00093206572991894022, 0.00156817541687766086, 0.00051169053398325657, 0.00077312532731958154, 0.00004559742198773287, 0.00000001037808743568, 0.00000000000000871706, 0.00000000000000000000, 0.00000000000000000000, 0.00000000000000000000, 0.00000000000000000000, 0.00345173057742305216, 0.16303340280710845733, 0.18328936325298567089, 0.32191072063759601507, 0.12710730373689624328, 0.19479707117476749434, 0.07149501410718961392, 0.05102665325739586671, 0.02061953603483871952, 0.01232173478541280856, 0.00339291179532653139, 0.00314522392528060902, 0.00050543218629833602, 0.00000944936698432368, 0.00000000009143740374, 0.00000000000000000000, 0.00000000000000000004, 0.00000000000000000321, 0.00000000000000000290, 0.00000000000000000501, 0.00000000000000000835, 0.00000000000000000088, 0.00000000000000000146, 0.00000000000000000000, 0.00000000000000000592, 0.00000000000000000000, 0.00000000000000000345, 0.00000000000000000036, 0.00000000000000000132, 0.00000000000000000276, 0.00000000000000000001, 0.00000000000000000462, 0.03125, 0.03125, 0.03125,
+0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125,
+0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125,
+0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125,
+0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125, 0.03125,
+0.03125), .Dim = c(32L, 4L),
+.Dimnames = list(NULL, c("n", "h","c", "")))
+
+  parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
+                                     params, pState1, nState1, pTrans1, d=3)
+  parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
+                                     od = parametersSet2$od, params = parametersSet2$params,
+                                     pState2, nState2, pTrans2, d=3)
+  parametersSet4 <- add_k_mer_state2(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
+                                     od = parametersSet3$od, params = parametersSet3$params, 
+                                     pState3, nState3, pTrans3, d=3)
+  parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
+                                     od = parametersSet4$od, params = parametersSet4$params, 
+                                     pState4, nState4, pTrans4, d=3)
+  overall_probs_log <- signalHsmm2010$overall_probs_log
+  maxSignal <- 45
+```
+
+```{r}
+registerDoMC(cores=7)
+n <- length(benchmark_data2)
+results <- foreach(i=1:n) %dopar% {
+    prot <- benchmark_data2[[i]]
+    deg_sample <- na.omit(as.numeric(degenerate(toupper(prot)[1L:min(maxSignal, length(prot))], group_best)))
+
+    viterbi_res <- duration_viterbi2(deg_sample-1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, 
+                                     signalHsmm2010$od, signalHsmm2010$params)
+    viterbi_path <- viterbi_res[["path"]]+1
+    c_site <- ifelse(any(viterbi_path == 4), 
+                     max(which(viterbi_path == 3)), 
+                     length(deg_sample))
+    prob.signal <- viterbi_res[["viterbi"]][c_site, viterbi_path[c_site]]
+    prob.non <- Reduce(function(x, y) x + overall_probs_log[y], deg_sample[1L:c_site], 0)
+    prob.total <- exp(prob.signal - prob.non)
+
+    viterbi_res5 <- duration_viterbi2(deg_sample-1, parametersSet5$pipar, parametersSet5$tpmpar, 
+                                      parametersSet5$od, parametersSet5$params)
+    viterbi_path5 <- viterbi_res5[["path"]]+1
+    c_site5 <- ifelse(any(viterbi_path5 == 4), 
+                      min(which(viterbi_path5 == 4))-1, 
+                      length(deg_sample))
+    c_site5 <- ifelse(any(viterbi_path5 == lastState1), 
+                      min(which(viterbi_path5 == lastState1)), 
+                      c_site5)
+    c_site5 <- ifelse(any(viterbi_path5 == lastState2), 
+                      min(which(viterbi_path5 == lastState2)), 
+                      c_site5)
+    c_site5 <- ifelse(any(viterbi_path5 == lastState3), 
+                      min(which(viterbi_path5 == lastState3)), 
+                      c_site5)
+    c_site5 <- ifelse(any(viterbi_path5 == lastState4), 
+                      min(which(viterbi_path5 == lastState4)), 
+                      c_site5)
+    #get probabilities of signal peptide model
+    prob.signal5 <- viterbi_res5[["viterbi"]][c_site5, viterbi_path5[c_site5]]
+    #get probabilities of no signal peptide model
+    prob.non5 <- Reduce(function(x, y) x + overall_probs_log[y], deg_sample[1L:c_site5], 0)
+    prob.total5 <- exp(prob.signal5 - prob.non5)
+
+    c(1 - 1/(1 + prob.total), 1 - 1/(1 + prob.total5), c_site, c_site5)
+}
+probs <- t(sapply(results, function (x) head(x, 2)))
+cut <- t(sapply(results, function (x) tail(x, 2)))
+```
+
+```{r, echo=FALSE}
+ggplot(data.frame(noMer=probs[,1],kMer=probs[,2], sig = factor(real_labels)), aes(x=noMer,y=kMer, color=sig)) +
+  geom_point() + xlab("No k-mer") + ylab("with k-mer") + 
+  scale_color_discrete("Signal peptide") +geom_abline(slope=1)
+```
+
+```{r, echo=FALSE}
+all_preds2 = all_preds
+all_preds2$signalKmer = probs[,2]
+
+bench_metrics2 <- HMeasure(real_labels, all_preds2,
+                             threshold = c(rep(0.5, 5), 0.1, 0.1, 0.5))[["metrics"]]
+kable(bench_metrics2)
+```

### 2015-07-06 signalHSMM_kMer_report.Rmd
redo report
now report is in concordance with the new signalHsmm

@@
 ```{r}
-  parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
+  parametersSet2 <- add_k_mer_state(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
                                      signalHsmm2010$params, pState1, nState1, pTrans1, d=3)
-  parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
+  parametersSet3 <- add_k_mer_state(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
                                      od = parametersSet2$od, params = parametersSet2$params,
                                      pState2, nState2, pTrans2, d=3)
-  parametersSet4 <- add_k_mer_state2(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
+  parametersSet4 <- add_k_mer_state(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
                                      od = parametersSet3$od, params = parametersSet3$params, 
                                      pState3, nState3, pTrans3, d=3)
-  parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
+  parametersSet5 <- add_k_mer_state(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
                                      od = parametersSet4$od, params = parametersSet4$params, 
@@
 
-    viterbi_res <- duration_viterbi2(deg_sample-1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, 
+    viterbi_res <- duration_viterbi(deg_sample-1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, 
                                      signalHsmm2010$od, signalHsmm2010$params)
@@
 
-    viterbi_res5 <- duration_viterbi2(deg_sample-1, parametersSet5$pipar, parametersSet5$tpmpar, 
+    viterbi_res5 <- duration_viterbi(deg_sample-1, parametersSet5$pipar, parametersSet5$tpmpar, 
                                       parametersSet5$od, parametersSet5$params)
@@
 
-  parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
+  parametersSet2 <- add_k_mer_state(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
                                      params, pState1, nState1, pTrans1, d=3)
-  parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
+  parametersSet3 <- add_k_mer_state(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
                                      od = parametersSet2$od, params = parametersSet2$params,
                                      pState2, nState2, pTrans2, d=3)
-  parametersSet4 <- add_k_mer_state2(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
+  parametersSet4 <- add_k_mer_state(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
                                      od = parametersSet3$od, params = parametersSet3$params, 
                                      pState3, nState3, pTrans3, d=3)
-  parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
+  parametersSet5 <- add_k_mer_state(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
                                      od = parametersSet4$od, params = parametersSet4$params, 
@@
 
+save(benchmark_data2, file = "clean_data.RData")
+
 bench_metrics2 <- HMeasure(real_labels, all_preds2,

### 2015-08-01 paper.tex
updated plasmodium benchmark

@@
 Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases.
-It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textit{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84).
+It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textit{Plasmodium} and their relatives more accurately (with AUC = 0.93) than popular programs (0.79).
 The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.
@@
 On the other hand, signal peptides show great variation and the description presented above (Fig.
-\ref{fig:sparch}) refers to the most ‘typical’ signal peptides.
+\ref{fig:sparch}) refers to the most ???typical’ signal peptides.
 There are  exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}.
@@
 Therefore, signal peptides became the subject of many computational programs to their prediction.
-Many software incorporates ‘black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}.
+Many software incorporates ???black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}.
 However, these models do not provide direct biological information about organization of signal peptides and are not able to predict properly atypical signal peptides.
-Although there are programs that do not share the innate flaws of ‘black-box’ models, they also demand an improvement.
+Although there are programs that do not share the innate flaws of ???black-box’ models, they also demand an improvement.
 Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}.
@@
 \begin{table}[ht]
-\small
 \centering
-\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to \textit{Plasmodiidae}.} 
+\caption{Comparison of Area Under the Curve, H-measure and Matthews Correlation Coefficient for different classifiers considering only proteins belonging to Plasmodiidae.} 
 \begin{tabular}{lrrrr}
@@
   \midrule
-signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
-   \rowcolor[gray]{0.85}PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & \textbf{0.9469} & 0.4028 \\ 
-  Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
-   \rowcolor[gray]{0.85}Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
-  signalHsmm-2010 & \textbf{0.9340} & 0.7941 & 0.8939 & 0.6526 \\ 
-   \rowcolor[gray]{0.85}signalHsmm-1989 & 0.9326 & 0.9216 & 0.8966 & \textbf{0.7531} \\ 
-  signalHsmm-2010 with k-mers & 0.9334 & \textbf{0.9902} & 0.7989 & 0.6767 \\ 
+signalP 4.1 (no tm) \citep{2011petersensignalp} & 0.8356 & 0.7745 & 0.8966 & 0.6420 \\ 
+   \rowcolor[gray]{0.85}signalP 4.1 (tm) \citep{2011petersensignalp} & 0.7928 & 0.6471 & 0.9385 & 0.6185 \\ 
+  PrediSi \citep{2004hillerpredisi} & 0.6597 & 0.3725 & 0.9469 & 0.4028 \\ 
+   \rowcolor[gray]{0.85}Phobius \citep{2004klla} & 0.7963 & 0.6765 & 0.9162 & 0.5991 \\ 
+  Philius \citep{2008reynoldstransmembrane} & 0.7753 & 0.6176 & 0.9330 & 0.5841 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 & \textbf{0.9340} & \textbf{1.0000} & 0.8436 & \textbf{0.7380} \\ 
+  signalHsmm-1989 & 0.9326 & 0.9510 & \textbf{0.8631} & 0.7266 \\ 
+   \rowcolor[gray]{0.85}signalHsmm-2010 with k-mers & 0.9334 & 0.9902 & 0.7989 & 0.6767 \\ 
    \bottomrule

### 2015-08-01 benchmark_signalHsmm.R

@@ -4,7 +4,7 @@ library(seqinr)
 library(hmeasure)
 library(dplyr)
 library(foreach)
-library(doMC)
+#library(doMC)
 other_soft <- read.csv2("benchmark_other.csv")
 
 group_best <- list(`1` = c("r", "n", "d", "q", "e", "h", "k"), 
@@ -66,27 +66,27 @@ lastState2 = lastState1+1+length(kMer2)
 lastState3 = lastState2+1+length(kMer3)
 lastState4 = lastState3+1+length(kMer4)
 
-parametersSet2 <- add_k_mer_state2(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
+parametersSet2 <- add_k_mer_state(kMer1, signalHsmm2010$pipar, signalHsmm2010$tpmpar, signalHsmm2010$od, 
                                    signalHsmm2010$params, pState1, nState1, pTrans1, d=3)
-parametersSet3 <- add_k_mer_state2(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
+parametersSet3 <- add_k_mer_state(kMer2, pipar = parametersSet2$pipar, tpmpar = parametersSet2$tpmpar, 
                                    od = parametersSet2$od, params = parametersSet2$params,
                                    pState2, nState2, pTrans2, d=3)
-parametersSet4 <- add_k_mer_state2(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
+parametersSet4 <- add_k_mer_state(kMer3, pipar = parametersSet3$pipar, tpmpar = parametersSet3$tpmpar, 
                                    od = parametersSet3$od, params = parametersSet3$params, 
                                    pState3, nState3, pTrans3, d=3)
-parametersSet5 <- add_k_mer_state2(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
+parametersSet5 <- add_k_mer_state(kMer4, pipar = parametersSet4$pipar, tpmpar = parametersSet4$tpmpar, 
                                    od = parametersSet4$od, params = parametersSet4$params, 
                                    pState4, nState4, pTrans4, d=3)
 overall_probs_log <- signalHsmm2010$overall_probs_log
 maxSignal <- 45
 
-registerDoMC(cores=4)
[…]

### 2015-08-03 literature/literature.Rmd
started reading literature

---
title: "Literature"
bibliography: "kmer_filo.bib"
output:
  html_document:
    toc: true
    theme: united
---

# One by one

```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
library(knitcitations)
library(RefManageR)
cleanbib()
BibOptions(sorting = "none", check.entries = FALSE)
refs <- read.bibtex("kmer_filo.bib", .Encoding = "UTF-8")
```

```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
cat(paste0("##", citet(refs[["saha_brief_2013"]])))
```

They mention few investigations, most of them uses n-grams. The `r citet(refs[["wang_application_2000"]])` was the most appealing.

```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
cat(paste0("##", citet(refs[["wang_application_2000"]])))
```

Uses aggregating of amino acids (called exchange group). Do not state why they created such clusterings. 'Exchange groups represent conservative re- placements through evolution.' They also sport a 'distance measure', which is more like the IG in the QuiPT.

Cool idea: 'To compensate for the possible loss of information due to
ignoring the other 2-gram patterns, a linear correlation coefficient (LCC) between the values of the 436 2-gram patterns
with respect to the protein sequence S and the mean value
of the 436 2-gram patterns in the positive training dataset
is calculated and used as another input feature value for S.'

They have quite expanded motif usage for the bayesian neural network.

# References

```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
sort(refs, sorting = "ynt")
```

### 2015-08-03 literature/generate_rev.R

@@ -1,0 +1,4 @@
+library(rmarkdown)
+render(input = "./literature/literature.Rmd", output_format = 'html_document')
+
+tmp <- read.bibtex("./literature/kmer_filo.bib")

### 2015-08-04 literature/literature.Rmd
added more literature

@@
 
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
+cat(paste0("##", citet(refs[["xie_discriminative_2015"]])))
+```
+
+Consecutive n-grams + random forests. Good n-gram reference: Manning and Schuetze, 1999. Benchmark with other methods.
+
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
+cat(paste0("##", citet(refs[["behnam_geometric_2013"]])))
+```
+
+HGT as the example of phenomenon missed by alignment-based phylogenetics.
+
+Critique of $D^2$ statistic and introduction of its variant. Introduction of bichromatic closest
+pair. Article focused on the speed.
+
 # References

### 2015-08-04 literature/literature.Rmd
started with a review reading

@@
 
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
+cat(paste0("##", citet(refs[["behnam_geometric_2013"]])))
+```
+
+HGT as the example of phenomenon missed by alignment-based phylogenetics.
+
+Critique of $D^2$ statistic and introduction of its variant. Introduction of bichromatic closest
+pair. Article focused on the speed.
+
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
+cat(paste0("##", citet(refs[["yang_alignment_2013"]])))
+```
+
+Don't understand. K0mer based, may be used for phylogenetics. Does not require specyfying $k$.
+
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
+cat(paste0("##", citet(refs[["bonham-carter_alignment-free_2014"]])))
+```
+
+Good review.
+Description of: factor frequencies, BBC (mutual information of DNA 2-grams), FFP (number of words commong among two sequences), Jensen-Shannon Divergence test (a test using Kullback-Leibler Divergence), suffix trees by k-mer frequencies (Euclidean distance between frequency counts, get optimum $k$ called $l$), improved complete composition vector (distance between normalized frequency of positioned k-mers with different $k$),
+
 # References

### 2015-08-12 literature/literature.Rmd
fiddling with the literature

@@
 
-They mention few investigations, most of them uses n-grams. The `r citet(refs[["wang_application_2000"]])` was the most appealing.
+They mention few investigations, most of them uses n-grams. The `r tmp <- citet(refs[["wang_application_2000"]])` was the most appealing.
 
@@
 ```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
-cat(paste0("##", citet(refs[["yang_alignment_2013"]])))
+cat(paste0("##", citet(refs["yang_alignment_2013"])))
 ```
@@
 
+```{r, echo=FALSE,message=FALSE,warning=FALSE,results='markup'}
+search()
+
+works.cited
+```
+
 # References
-
-```{r, echo=FALSE,message=FALSE,warning=FALSE,results='asis'}
-sort(refs, sorting = "ynt")
-```
